% !TEX TS-program = lualatex
\documentclass[11pt, a4paper]{article}

% LuaLaTeX handles UTF-8 natively; fontspec lets the document pick up any
% Unicode glyphs present in the default font, including mathematical script
% characters used in listings (ℕ, ℤ, ℚ, ℝ, ℂ, Ø, and similar). We keep the
% `newunicodechar` fallbacks for body text so amsmath glyphs (\mathbb) are
% used in paragraphs.  Code listings use the `minted` package (Pygments-
% backed) with native UTF-8 — no per-character literate table needed.
\usepackage{fontspec}
\usepackage{newunicodechar}
\usepackage[english]{babel}    % Language support
\usepackage{geometry}          % Adjust page margins
\geometry{margin=1in}

% --- Recommended Math & Science Packages ---
\usepackage{amsmath, amssymb, amsthm} % Essential math tools
\usepackage{mathtools}
\usepackage{mathrsfs}

% Theorem environments (amsthm is loaded above).
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\newtheorem{conjecture}{Conjecture}
\usepackage{graphicx}                  % Support for images
\usepackage{hyperref}                  % Clickable links & cross-refs
\usepackage{booktabs}                  % High-quality tables
% csquotes is loaded later — after `minted` (which pulls in `fvextra`)
% to silence the fvextra/csquotes load-order warning.

\newunicodechar{∩}{\ensuremath{\cap}}
\newunicodechar{ℚ}{\ensuremath{\mathbb{Q}}}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{𝔽}{\ensuremath{\mathbb{F}}}
\newunicodechar{∃}{\ensuremath{\exists}}
\newunicodechar{π}{\ensuremath{\pi}}
\newunicodechar{⊂}{\ensuremath{\subset}}
\newunicodechar{⊆}{\ensuremath{\subseteq}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{∉}{\ensuremath{\notin}}
\newunicodechar{∧}{\ensuremath{\land}}
\newunicodechar{∨}{\ensuremath{\lor}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≥}{\ensuremath{\geq}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{β}{\ensuremath{\beta}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{Ω}{\ensuremath{\Omega}}
\newunicodechar{↔}{\ensuremath{\leftrightarrow}}
\newunicodechar{χ}{\ensuremath{\chi}}
\newunicodechar{Σ}{\ensuremath{\Sigma}}
\newunicodechar{Τ}{\ensuremath{\mathrm{T}}}

% For code listings
\usepackage{xcolor}
% Landscape tables:
\usepackage[figuresright]{rotating}
\usepackage[table]{xcolor}
\usepackage{pdflscape}
\usepackage{tabularx}
\usepackage{booktabs}


% Graphs
\usepackage{graphviz}

\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, calc, backgrounds}
\usepackage{tikz-cd}

\usepackage{draftwatermark}
\SetWatermarkScale{5}

\usepackage[style=numeric, sorting=none]{biblatex}
\addbibresource{references.bib}     % Points to your .bib file (include the extension)

% `minted` (Pygments-backed) handles UTF-8 natively in code listings —
% no per-character literate table needed.  Requires `lualatex -shell-escape`
% (the build pipeline in docs/paper/Makefile and the CI install step in
% the project Makefile are updated accordingly).
\usepackage{minted}
\usepackage{csquotes}  % Load after minted/fvextra to silence load-order warning.

% Use JuliaMono for code listings — designed specifically for Unicode-rich
% source (Julia uses the full math repertoire) with native coverage of
% Greek (φ, Σ, Τ, χ, ...), math double-struck (ℕ, ℤ, ℚ, ℝ, ℂ, 𝔻, 𝔽, ...),
% math-fraktur, math-script, sub/superscripts, and arrows.  Replaces the
% earlier `lmmono10` (no Greek) and `Fira Mono` (no math-double-struck)
% choices made in the lstlisting era.  Installed locally via
% `brew install --cask font-juliamono`; CI install adds `fonts-juliamono`
% (Ubuntu 22.04+) under ci-install-paper-deps in the project Makefile.
% Scale=0.90 keeps glyph metrics close to the document's serif body.
\setmonofont{JuliaMono}[Scale=0.90]

\setminted{
  fontsize=\small,
  breaklines=true,
  breakanywhere=true,
  frame=single,
  framesep=2mm,
  autogobble,
}
\usemintedstyle{friendly}

% C++ is the default code-listing language in this paper; declare short
% block + inline aliases so source markup reads `\begin{cpp}...\end{cpp}`
% and `\cppinline{...}` rather than the verbose `\begin{minted}{cpp}` /
% `\mintinline{cpp}{...}` forms.  The optional `[name]` first argument
% names the environment directly; without it minted would produce
% `\begin{cppcode}` and `\cpp{...}`, which would collide with our \cpp
% short alias usage.
\newminted[cpp]{cpp}{}
\newmintinline[cppinline]{cpp}{}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}

\newcommand{\bind}{\mathbin{>\!\!>\mkern-6.7mu=}}

% --- Document Metadata ---
\title{Sets Are Rules, Not Buckets:\\
  Intensional Programming in C++23 via \texttt{dedekind}\thanks{%
    Draft intended for submission to \emph{Programming}~\cite{programming_journal},
    the journal for programming-related research.
    The paper's scope is deliberately narrower than the companion project
    report: this document focuses on the sets-as-predicates thesis and
    compile-time structural pruning; the broader library (categorical
    foundations, algebra tower, numeric tower, geometry, linear algebra,
    morphologies, Python facade) is documented in the companion Report
    distributed alongside the source tree.}}
\author{The \texttt{dedekind} Authors}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent\emph{Named algebraic structure at the type level, discharged
by the compiler, narrows the gap between abstract mathematics and
machine code without a separate proof assistant.} The underlying
theory is not new---ETCS~\cite{lawvere1964etcs}, Wadler's
parametricity and propositions-as-types
reading~\cite{wadler1989theorems,wadler2015propositions}, and
proof-assistant-grounded formalisation in Coq, Agda, and Lean have
been studied for decades. What is new is the \emph{cost-benefit
calculus}: modern transformer-based coding assistants substantially
compress the time and expertise needed to author opt-in trait
specialisations, axiom-shape concepts, and the surrounding library
scaffolding, turning what was an academic specialism into an
actionable workflow for working software engineers.
The meta-claim is not ours to originate; we make it here so the body's executable exhibit can be read in that light, and the §\ref{sec:acknowledgements} disclosure documents how the calculus actually shifted on this paper specifically (which tools, which tasks, what oversight pattern).

We present \texttt{dedekind}~\cite{dedekind2026}, an
open-source C++23 library (Apache-2.0), as that exhibit. It embeds
formal set-theoretic and categorical structures directly into the
type system and targets \emph{mainstream C++23}---concepts,
non-type template parameters, named modules---rather than a research
compiler. The design rule is \emph{intensional first, realize when
you mean it}: a set is represented as a predicate $P : A \to \Omega$
over an ambient carrier type $A$, not as a heap-allocated container, with
two practical consequences:
\begin{itemize}
  \item Set operations (intersection, complement, union) compose as
        predicate combinators at compile time, enabling the
        \textsc{llvm} backend to perform \emph{structural pruning}:
        logically empty intersections collapse into zero-instruction
        machine code.
  \item Numeric realisation is deferred to explicit boundaries,
        cleanly separating abstract algebraic reasoning (e.g.\ over
        the naturals $\mathbb{N}$) from hardware-constrained IEEE~754
        floating-point~\cite{ieee754_2019} semantics.
\end{itemize}
The paper identifies \emph{Axiomatic Systems Programming} as a
nascent paradigm in which the compiler is a structural gate, not a
theorem prover, and demonstrates the discipline through a textbook
linear-programming instance that reduces at compile time to its
optimal vertex as a typed constant; the same reduction over a
dual-number ring additionally recovers forward-mode automatic
differentiation~\cite{elliott2018simple} as a sibling compile-time
collapse.

\noindent A subsidiary observation we offer alongside: working in the disciplined subset of C++23 the paper develops carries a striking resemblance, by structural analogy (§\ref{sec:jlt}), to using \texttt{clang++} as a System~F type-checker, with optimised machine code emerging as a side effect rather than from a separate compilation step.  We name the resemblance because the body's exhibits invite it; a formal correspondence is not in scope.
\end{abstract}

\tableofcontents

\vspace{0.75em}
\noindent\textbf{Keywords:} C++23; concepts; non-type template
parameters; compile-time verification; Axiomatic Systems Programming;
type-directed collapse; linear programming; forward-mode automatic
differentiation.

% ============================================================
\section{Introduction}
\label{sec:introduction}
% ============================================================

A software developer who wants to expose mathematical abstractions in a
programming language faces a recurring tension.
Algebraic structures---groups,
rings, fields, vector spaces---are defined in the abstract by laws which are hard to enforce: associativity,
commutativity, identities, inverses, closure, totality and many more.
The concrete carriers\footnote{We use ``carrier'' for the underlying set $A$ of a $\Sigma$-algebra in the strict universal-algebra sense (Burris--Sankappanavar~\cite{burris1981universalalgebra}), distinct from the algebra $(A, \Sigma_A)$ itself.  Where the prose says ``the carrier $\mathrm{T}$'' it is shorthand for the pair inhabited under the relevant concept.  Section~\ref{sec:alg-sigma} makes the layering precise; the HAS-A posture of named carriers (algebra HAS-A carrier $\Leftrightarrow$ carrier IS-PART-OF algebra in the mereological reading) is encoded mechanically by the \cppinline{HasCarrier<Algebra, Carrier, Ops...>} concept in \texttt{algebra:universal}, which composes \cppinline{IsAlgebra<Carrier, Ops...>} (algebra-side closure under the operations) with the parthood primitives \cppinline{IsPartOfRelation} and \cppinline{arrow_drill_down} from \texttt{category:mereology}.  The sibling \cppinline{IsTotalAlgebra<T, Ops...>} concept names the totality tier between closure (\cppinline{IsAlgebra}) and the per-axiom laws hierarchy in \texttt{category:total} (\cppinline{IsRing}, \cppinline{IsField}, \dots): closure plus \cppinline{IsTotal<T, Op>} on each operation, the gate the laws-tier concepts are built on.  Pinned end-to-end on $\mathbb{F}_2 = (\mathbb{B}, \oplus, \wedge)$ as the concrete-algebra witness for both tiers.} a programming language offers violate those laws conditionally, sometimes silently:
sets must fit into memory or cache, functions may throw runtime exceptions or return \cppinline{nullptr};
a 32-bit signed \cppinline{int} is not an additive group (overflow, \cppinline{INT_MIN} has no
negation), an IEEE~754 \cppinline{double}~\cite{ieee754_2019} may overflow, underflow, be \cppinline{NaN},
and is not associative under addition in the bit-perfect sense.
The options are then to 
i) ignore the divergence and try to mitigate or silence bugs on specific inputs,
ii) police it at runtime, or 
iii) encode it in the software build process so the build system can police and maybe exploit it.
In real life, the pragmatic choice is often the first one: the programmer defines a \cppinline{std::vector<std::vector<double>>}, calls it a tensor and gets on with life.

Here, we explore and report what might happen if a library implemented \emph{in a mainstream systems programming language} takes the third
option seriously and consistently defaults to mathematical rigor and build-time constraints as opposed to risking runtime errors.
The hope being to find better answers to questions such as the following:
Will we be able to explain to the type system that an inner product of two vectors is a scalar while the outer product is a matrix of rank 1?
Will it recognize that the intersection of a set of integers with a set of real numbers must return a subset of the integers?
What might the type of a function space even look like?
Will it be possible to design a system that can switch between symbolic operations on intensional concepts such as transfinite sets and extensional properties or is symbolic algebra a capability that will remain locked away in languages which are somehow more expressive?
What happens if, either by accident or because of inherent complexity, neither the compiler nor the programmer can come up with the answer in a sufficiently short period of time? Is it possible to design escape hatches when the limits of computability are fast approaching?

Turing showed us a long time ago that determining the limits of computability up-front is not achievable \emph{in the general case} and so we can anticipate the outcome: a hopefully growing collection of \emph{existential proofs and fragments} that facilitate computation where we know it to be feasible.
A plausible and occasionally tractable process to support this might look about as sketched in Figure~\ref{fig:dedekind-bridge}:
1. find pre-validated \emph{existing proofs} and turn them into specifications,
2. translate the specifications into source code and tests,
3. type-check the code and run the tests, reviewing the output,
4. compare and contrast with the original textbook specification (1.).

Math goes one way; C++ comes back.
The claim is bounded --- only the part of mathematics that fits a disciplined fragment of C++23 is realised, and even then we make no global claim of having proved the realisation faithful.
Instead, a growing body of \emph{empirical} evidence is aggregated and compiled.
The discharge is mechanical, the architecture is set up so the evidence accumulates in a single direction.
If the compiler accepts the axiom being imported, then the compiler should also be able to validate or sometimes even derive the corresponding theorems~\cite{wadler2015propositions}.
The remainder of this paper, starting with §\ref{sec:axiomatic}, is intended to illustrate how this was done in some detail and what was found in the process.


\begin{figure}[htbp]
\centering
\begin{tikzcd}[column sep=huge, row sep=large, ampersand replacement=\&]
\mathbf{Cat} \arrow[r, shift left=2, "\text{1. cite and specify}"] \&
\texttt{dedekind} \arrow[l, shift left=2, "\text{4. compare and contrast}"] \arrow[r, shift left=2, "\text{2. encode and review}"] \&
\mathbf{Cpp} \arrow[l, shift left=2, "\text{3. type-check and test}"]
\end{tikzcd}
\caption{The target operational model.  \texttt{dedekind} aims at being a faithful intermediary between $\mathbf{Cat}$ (mathematical structure in general, axiomatically defined) and a fragment of $\mathbf{Cpp}$ (the category of C++ types, values and arrows between them).
Math $\to$ library: an interpretation maps a species (a ring, a vector space, a topos primitive) to a corresponding \texttt{dedekind} concept / trait / module.
Library $\to$ math: the library's declarations cite the underlying axiomatic source so the resulting concept has a textbook anchor.
Library $\to$ C++: the library instantiates concrete carriers (\cppinline{Rational<long>}, \cppinline{Vec2V<T>}, \dots) under the named structure.
C++ $\to$ library: the compiler checks each carrier via \cppinline{static_assert} and concept-constraint discharge --- the propositional-fragment instance of the Curry--Howard--Lambek correspondence~\cite{lambek1988introduction,wadler2015propositions}.
}
\label{fig:dedekind-bridge}
\end{figure}


The choice of anchoring the library's domain in category theory on the mathematical side follows a well-trodden path~\cite{pierce2002tapl}.
Here, a minimalist type checker does not need to do much more than confirm whether the output of one function matches the input of another (morphism composition).
Rewrite rules (e.g.\ natural transformations) and optimization enter the picture at a later stage.
Once this basic rapport is established, the path follows an established progression:
1. cartesian-closed categories~\cite{lambek1988introduction} giving rise to a simply-typed $\lambda$-calculus~\cite{lambek1988introduction} on sets;
2. the elementary theory of the category of sets (ETCS)~\cite{lawvere1964etcs}; and
3. universal algebra and abstract algebra~\cite{burris1981universalalgebra,lang2002algebra}.

On the encoding side, the choice is more open --- any sufficiently expressive type system would do --- but we landed on modern C++ for the following reasons.
The goal is an embedding in an existing system (an eDSL in a library) as opposed to the development of a new language or compiler (a language or framework), and so the host language should meet the following requirements:
a) it should be mainstream so that tooling is provided for;
b) its runtime requirements should be modest enough to embed in the Python and JVM ecosystems and to target resource-constrained environments in the high-performance-computing or embedded-systems sense;
c) its type system should be rich enough to perform many of the desired type checks.
Notably, while the resulting DSL should resemble mathematical notation so that Figure~\ref{fig:dedekind-bridge} can be operationalized, developer ergonomics in the broader sense were deliberately not given a high priority at this point in time.

Our resulting posture is explicit: \texttt{dedekind} is a systems-programming
library that treats mathematical concepts as first-class compile-time
objects.
The library is implemented in C++23, and the compile-time
concept machinery is its primary interface; the consumer we design
for is another C++ author who would otherwise bolt ad-hoc structure
onto hard to verify numeric types.
Choosing to land the semantics at the
compile-time layer---rather than as runtime abstractions, or as a DSL
lowered to a runtime solver---is a design commitment that shapes what
the compiler and the library can do for the user and what they cannot.

\paragraph{The thesis.} In one sentence: \emph{named algebraic structure
at the type level, discharged by the compiler, narrows the gap
between abstract mathematics and machine code without a separate
proof assistant.}
Figure~\ref{fig:dedekind-bridge} illustrates this overarching goal.
Every section of the paper serves that sentence.

\paragraph{How is this new?}
Compile-time verification of algebraic structure is not new in the
abstract. Agda and Coq host it natively; Liquid Haskell attaches
refinement predicates to types; C++ template-metaprogramming traditions
(most visibly Boost.Hana~\cite{chagnon2016hana} and
CTRE~\cite{dusikova2018ctre}) have demonstrated type-level computation
at scale for over a decade; autodiff libraries such as
Ceres-Solver~\cite{ceres-solver} and ENOKI~\cite{jakob2019enoki} have
carried fragments of the dual-number story into production C++. What this
paper adds is an end-to-end exhibition of the technique---named
algebraic structure, discharged by the compiler, with the LLVM
backend carrying the discharge through to optimised machine
code---in standard C++23, without a proof assistant, a custom
type checker, or a bespoke language extension. \texttt{dedekind}
is the existence proof. The viewpoint on which it rests we name
\emph{Axiomatic Systems Programming}: an engineering discipline
of practising, not merely describing, type-level structural verification
in mainstream compiled code. The discipline transfers to any language
hosting System-F-style polymorphism with structural concept-style
constraints; the active research line on extending System F with
modern type-system disciplines is published at length (see, e.g.,
the DOT calculus underlying
Scala~3~\cite{amin2016essence} and the capability-tracking line
of work~\cite{boruchgruszecki2023capturing}). The contribution of
this paper is an articulation of that discipline restricted to
mainstream C++23, together with a worked library that discharges
its claims mechanically.

\paragraph{What shifted the cost--benefit calculus.}
Two recent developments are the \emph{key enablers}, not tricks or shortcuts.
First, the C++23 standard~\cite{cpp23standard} delivers the three features the technique needs --- \cppinline{concept}s, NTTPs, and named modules --- in a mainstream toolchain with an installed industrial base; previous template-metaprogramming traditions had to encode these implicitly at steep readability and review cost.
Second, modern coding-assistant LLMs~\cite{chen2021evaluating} compress the time and expertise needed to author opt-in trait specialisations, axiom-shape concepts, and the surrounding library scaffolding, turning what was once a deep-template-metaprogramming specialism into an actionable workflow for working software engineers.
The correctness story is unchanged --- the type system and \cppinline{static_assert} obligations are what the compiler discharges, both checked mechanically --- but the cost story is, and that is what makes the discipline viable at industrial scale rather than as a paper-only artefact.

While this realization targets C++ explicitly for the reasons stated earlier, much of what follows is however \emph{not} specific to C++.
Concepts have close analogues in Java (sealed types plus generic constraints),
C\# (constrained generics, augmented by Roslyn source generators and
analysers), and Python (PEP~484 type hints under a strict
\texttt{mypy}/\texttt{pyright} pass). Any language whose type system
can name a predicate over compile-time data and check it mechanically
is a candidate host for the same technique. What C++23 contributes on
top is the LLVM backend sitting downstream of the type checker:
discharged proofs do not only \emph{verify}, they also \emph{optimise},
and the connection between the two is what lets the LP vertex collapse
to a literal return rather than to a skipped runtime check.

\paragraph{Intended target.}
The discipline lands well in CPU-first high-performance computing and embedded
machine-learning deployments---small fixed-topology problems with
hard runtime-footprint constraints---where the technique's
compile-time-substrate posture sits one layer below existing
embedded ML runtimes such as TFLite Micro~\cite{david2021tflm}.

\paragraph{The compiler as a poor man's theorem prover} is
a pedagogical framing that runs through this paper.
The motivation is primarily structural, not economic:
\texttt{clang++} \emph{is} a formal system in the technical sense --- a typing
derivation it accepts is a proof in the propositional fragment
against a fixed inference rule-set.
The resulting architecture imposes this division of labour:
i) the engineer carries an honesty obligation up front~\cite{nspe_ethics} (NSPE, canons III + IV; truthfulness in public statements; faithful agency) and
ii) the compiler discharges every \cppinline{static_assert}, \cppinline{requires} or \cppinline{constexpr} evaluation that follows. 
The Curry--Howard reading~\cite{wadler2015propositions} grounds this technically; codified engineering ethics ground it
professionally: the library treats a trait specialisation in a public source code repository as a public statement subject to the canons. 
The argument is bounded: software claims in concurrent systems (TCP retransmit-timer correctness, scheduler fairness, end-to-end latency) often cannot be reduced to type-level structural assertions.
The physics of latency will usually not allow it.
The paper covers the fragment where the runtime model is \emph{physically localized} on trusted hardware and the reduction is available in principle.
Further, unlike a theorem prover the compiler is not actively searching for proofs above and beyond the built-in look-up rules.
It has no tactic language, no resolution procedure, no proof assistant in the loop.
What it \emph{does} have is a type system strong enough to name a predicate,
a \cppinline{static_assert} strong enough to discharge it, and an LLVM
backend that will happily remove what has already been decided.

\paragraph{What this buys, concretely.} From the design rule (``name
the structure in the type, let the compiler discharge the laws'') and
a small handful of primitives (concepts, NTTP-carried predicates,
halfspaces, cuts), the library builds, at compile time, a numeric
tower ($\mathbb{N}$, $\mathbb{Z}$, $\mathbb{Q}$, $\mathbb{R}$ via
Dedekind cuts, $\mathbb{C}$, dual numbers), a small linear-algebra
layer, a calculus layer (numerical / polynomial / dual-number AD),
and an optimisation layer (the LP vertex enumerator of
Section~\ref{sec:lp-centrepiece}). No runtime solver ships with any
of it. The same machinery lifts arrows over functorial families
($T \mapsto T^2 \mapsto M_2(T)$ etc.), pins ring monomorphisms
($T \hookrightarrow M_2(T)$ as a monadic unit), and extracts optima
from polytopes (the LP reduction of
Section~\ref{sec:lp-centrepiece} is a co-Kleisli
arrow);\footnote{%
Concretely: each of the scalar / column / row / square shape
constructors $(T, T^2, {}^t T^2, M_2(T))$ is functorial in $T$ and
ships an $\textsc{IsFunctor}$ witness, so an arrow $f: T \to T$
lifts elementwise without per-carrier boilerplate.  The scalar
embedding $T \hookrightarrow M_2(T)$ is the unit $\eta_{M_2}(s) =
s \cdot I_2$ of the shape monad.  The LP vertex extraction is a
co-Kleisli arrow $\varepsilon_{\mathrm{LP}}: \mathrm{Polytope}(T)
\to T^2$.  The library carries these as typed constants in the
emitted object file via the same primitives that discharge the
algebraic laws.  The discipline extends to polynomial calculus: the
formal derivative of a polynomial is a \texttt{static\_assert}-checked
identity, and the Fundamental Theorem of Calculus (Part~II) is
discharged at translation time as numerical agreement between a
trapezoidal quadrature and a closed-form antiderivative
(\texttt{dedekind.algebra:polynomial} and
\texttt{dedekind.analysis:ftc} partitions in the companion
repository).} the categorical vocabulary that names this machinery
(Table~\ref{tab:cpp-stlc-mapping}) is therefore not decorative ---
intensional sets in the sense of the
\emph{nLab}~\cite{nlab:structural_set_theory} (predicates into a
subobject classifier, situated in a Cartesian-closed
ambient~\cite{lambek1988introduction}) are the same machinery
discharging set membership and lifting structure-preserving arrows.

\paragraph{The three claims this paper substantiates.}
\begin{enumerate}
  \item Sets are \emph{rules} (predicates), not \emph{buckets} (containers).
  \item Computation is \emph{intensional until explicitly realized}.
  \item The compiler is used as a \emph{verification and optimization engine}
        for algebraic invariants.
\end{enumerate}
The remainder of the paper develops these three claims and demonstrates
each by a reduction that compiles to a constant. We make no pretence that a complete
and faithful taxonomy of mathematics is achievable; quite the contrary, we assert that this process is at best open-ended.
What we \emph{do} report is a small set of findings that are both interesting
and true: each claim has a worked exhibit in the library and a
mechanical witness the compiler discharges.


% ============================================================
\section{An Intersection of C++23 and Category Theory}
\label{sec:axiomatic}
% ============================================================

The strategy of this section is a three-step intersection.
Both category theory and modern C++23 are large and occasionally intimidating beasts; each, however, admits a tractable subset with somewhat simpler atomic building blocks and clear composition rules.
The \texttt{dedekind} library lives at their intersection.

\subsection{A Goldilocks Zone in Category Theory: $\mathbf{Alg}(-)_{\beth_1}$}
\label{sec:alg-sigma}

The choice of anchoring the library's domain in category theory on the mathematical side is neither new nor arbitrary and follows a well-trodden path.
The authoritative textbook in the context of programming might be \emph{Types and Programming Languages}~\cite{pierce2002tapl}.
For the more casual reader of this text, it should be sufficient to remember that category theory is a book-keeping device.
It tells us what clicks together (the arrows compose), what doesn't, and how things might be rearranged or simplified.

In the context of \texttt{dedekind}, we focus on a fragment of $\mathbf{Cat}$ that most readers will be familiar with: $\mathbf{Alg}(-)$, i.e.\ algebras in the universal-algebra sense.\footnote{The unrelated algebraic-geometry usage of ``algebraic set'' for the zero locus of a system of polynomials is \emph{not} what we mean.}\footnote{The dash is a typographic placeholder meaning ``fill in any signature here'' (Mac Lane's standard argument-blank convention~\cite{maclane1971categories}, same family as $\mathrm{Hom}(-, -)$ for the hom-bifunctor and $(-)^{\mathrm{op}}$ for the opposite-category functor); it is \emph{not} a unary minus.  The notation reappears in §\ref{sec:cpp-stlc-mapping}.}
Here, we adopt algebra textbook terminology where the universal-algebra category is $\mathbf{Alg}(\Sigma)$~\cite{burris1981universalalgebra} and $\Sigma$ denotes a fixed algebraic signature, i.e.\ a list of operation symbols with their arities.
However, \emph{we deliberately leave $\Sigma$ unbound for now}; the intent is to defer the formalisation of concrete operations.
The net result is a focus on sets equipped with operations realising as-yet-unspecified signatures $\Sigma$.
Concretely, $\mathbf{Alg}(\Sigma) \subset \mathbf{Alg}(-)$ for any fixed $\Sigma$; varieties such as $\mathbf{Mon}$ arise as $\mathbf{Alg}(\Sigma_{\mathrm{Mon}})$ for the monoid signature, and $\mathbf{Alg}(-)$ itself is a small fragment of $\mathbf{Cat}$ (with $\mathbf{Set}$ entering not as a containing supercategory but via the standard forgetful functor $U: \mathbf{Alg}(-) \to \mathbf{Set}$ that drops operations and keeps the carrier).
While the binding of concrete operator-shape signatures such as $\neg, \vee, \wedge, \cup, \cap, +, \times, \ldots$ is discussed in later sections~\ref{sec:cpp-stlc-mapping}, \ref{sec:sets-rules}, we can already name some familiar examples: the Booleans $\mathbb{B}$ with logical operations, the integers $\mathbb{Z}$ with addition and subtraction, and so on.

In functional programming terminology, these algebras correspond to $F_\Sigma$-algebras: for each signature $\Sigma$ there is a polynomial endofunctor $F_\Sigma$ on $\mathbf{Set}$, and Pierce~\cite{pierce1991basic} (§2.2) gives the isomorphism $\mathbf{Alg}(\Sigma) \cong F_\Sigma\text{-}\mathbf{Alg}$ that lets the same fragment read coherently to both audiences.
We note though that while $\mathbf{Alg}(\Sigma)$ is only a small fragment of $\mathbf{Cat}$ we expect that it carries enough algebraic vocabulary to host carriers, operators, and structure-preserving arrows in the shape an engineering library actually needs and so we use it as a starting point for a structuralist mathematical style~\cite{marquis2022structuralist}.

We further restrict our attention to the \emph{small} slice $\mathbf{Alg}(\Sigma)_{\beth_1}$ --- the $\beth_1$-small algebras, i.e.\ algebras whose underlying carrier has cardinality at most that of the continuum.\footnote{The standard mathematical workaround for ``a category whose objects are all sets / all categories'' is a Grothendieck universe hierarchy~\cite{maclane1971categories} (Mac Lane, §I.6): pick a universe $U$, define $\mathbf{Set}_U$ as the category of $U$-small sets, then let $\mathbf{Set}_U$ live as an object of a larger-universe $\mathbf{Cat}_{U'}$.  The hierarchy stratifies; no level contains itself.  The $\beth_1$ subscript here pins the smallness ceiling at the cardinality of the continuum --- enough to admit $\mathbb{R}$-sized carriers, small enough to forbid proper-class self-reference.}
Pinning the object-size at $\beth_1$ admits the transfinite carriers required for Dedekind-cut arithmetic ($\mathbb{R}$, §\ref{sec:numeric-tension}) and the continuum-sized $\mathbb{C}$, $\mathbb{D}$ used in §\ref{sec:linalg}, while keeping a foundational ceiling that forbids the self-referential paradoxes of the proper-class meta-categories $\mathbf{Set}$ and $\mathbf{Cat}$.


In the language of software architecture, the picture might be described as a hub-and-spoke layout.
Each hub is a carrier type $\mathrm{T}$, and the spokes are the maps that compose with it.
Operator overloads ($\mathrm{T} \times \mathrm{T} \to \mathrm{T}$), free-function actions on the carrier, trait specialisations registering structural facts about $\mathrm{T}$, and named embedding arrows $\mathrm{T} \to \mathrm{S}$ crossing to a sibling hub.
The library lifts this pattern directly into the build graph.
Each named module declares one or more hubs and owns its spokes.
The module and partition DAG is the diagram of inter-hub arrows.
Figure~\ref{fig:hub-spoke-bnz} sketches three concrete algebraic hubs ($\mathbb{B}$, $\mathbb{N}$, $\mathbb{Z}$) with one arrow of each kind.

\begin{figure}[htbp]
\centering
\begin{tikzcd}[column sep=5em, row sep=4em, ampersand replacement=\&]
\mathbb{B} \arrow[loop above, "\neg"]
   \arrow[r, bend left=20, "\mathfrak{i}"]
\& \mathbb{N} \arrow[loop above, "x^2"]
   \arrow[l, bend left=20, "x \,\%\, 2 = 0"]
   \arrow[r, bend left=20, "-x"]
\& \mathbb{Z} \arrow[loop above, "-x"]
   \arrow[l, bend left=20, "x^2"]
\end{tikzcd}
\caption{
An example subset of $\mathbf{Alg}(\Sigma)$: the carriers $\mathbb{B}$, $\mathbb{N}$, $\mathbb{Z}$.
Each carrier is depicted with its underlying set as the single object ($\{\mathrm{false}, \mathrm{true}\}$, $\{0, 1, 2, \ldots\}$, $\{\ldots, -1, 0, 1, \ldots\}$) and selected endo-operations as loops: the affine translation $x \mapsto x \oplus 1$ on $\mathbb{B}$ (i.e.\ logical negation, which on $\mathbb{F}_2$ is the additive translation by 1), squaring $x \mapsto x^2$ on $\mathbb{N}$, negation $x \mapsto -x$ on $\mathbb{Z}$.
This is a cartoon of the carrier-as-a-category picture, in the spirit of Mac Lane~\cite{maclane1971categories} §I.1's monoid $\leftrightarrow$ one-object-category bijection.
Equivalently (Pierce~\cite{pierce1991basic} §2.2): each loop is a component of the bundled structure arrow $\alpha : F_\Sigma(A) \to A$, where $A$ is the carrier, for the carrier's signature.
A strict realisation would draw one morphism per carrier element, impractical here, so only a few representative loops are shown.
The set's elements are not drawn separately; in this layer they live \emph{inside} the single object, exactly as in the category $\mathbf{Mon}$ of monoids.
Inter-hub arrows are typed maps $\mathrm{T} \to \mathrm{S}$.
$\mathbb{B} \rightleftarrows \mathbb{N}$: forward $\mathfrak{i}$ the canonical embedding, reverse the parity classifier.
$\mathbb{N} \rightleftarrows \mathbb{Z}$: forward $x \mapsto -x$ (negation, well-typed as $\mathbb{N} \to \mathbb{Z}$ because $\mathbb{N}$ is not closed under additive inverse; the canonical inclusion from the Grothendieck group completion is just $n \mapsto n$ and is not drawn), reverse $x \mapsto x^2$ (squaring closes into $\mathbb{N}$ from any $\mathbb{Z}$ input).
Most arrows on each hub and most inter-hub arrows are suppressed for clarity. The diagram shows one of each kind.
}
\label{fig:hub-spoke-bnz}
\end{figure}

The general thrust of the architecture is then to view these small, algebraic categories as universal underlying carrier sets, with specific subsets characterised in the ETCS sense by a subobject classifier $\Omega$.
A predicate $P : A \to \Omega$ maps ambient elements to \emph{truth values}, and $\Omega$ is the type those truth values live in.
Change $\Omega$, and you change what counts as membership.
The default is the one every programmer expects: $\Omega = \mathbb{B} = \{\top, \bot\}$, classical two-valued logic.
Plug it in and you get ordinary set theory.
But set structures with undecidable membership (e.g.\ the Mandelbrot set) can be supported this way also.
Swap it for Kleene $\mathbb{K}_3 = \{-1, 0, +1\}$ and something more interesting happens.
In the case of \cppinline{double}, the extra value \emph{Unknown} is where \texttt{NaN} lives.
In classical logic, $P \lor \neg P = \top$ is a theorem.
With \texttt{NaN} in play, that theorem quietly breaks, because neither $x < \texttt{NaN}$ nor
$x \geq \texttt{NaN}$ is true.
In $\mathbb{K}_3$, $P \lor \neg P = \mathit{Unknown}$ is a calm, unremarkable identity.
The paradox was never in the floats; it was in our choice of $\Omega$.
$\Omega$ parameterises membership truth values while the $\beth_1$ ceiling parameterises carrier size; the two are orthogonal.

\subsection{A Goldilocks Zone in C++: the Juliet Posture $\mathbf{Jlt} \subset (\mathbf{CCC} \cap \mathbf{Cpp})$}
\label{sec:jlt}

In Section \ref{sec:alg-sigma}, a mathematical Goldilocks zone was identified as $\mathbf{Alg}(-)_{\beth_1}$.
Here, we identify its engineering counterpart: the Juliet Posture $\mathbf{Jlt}$, a structural subcategory of the intersection $\mathbf{CCC} \cap \mathbf{Cpp}$.
This is the zone where C++ types exhibit the Cartesian Closed symmetry required to inhabit our algebraic hubs.

\begin{proposition}
The intersection $\mathbf{CCC} \cap \mathbf{Cpp}$ is non-empty and reifiable within the C++23 translation unit.
\end{proposition}

\begin{proof}
To prove the existence of this intersection, we exhibit a mapping from the physical substrate of $\mathbf{Cpp}$ to the axiomatic requirements of a Cartesian Closed Category (the internal language of $\mathbf{STLC}$).

\begin{enumerate}
    \item \textbf{Selection of Objects (The Carrier):}
    We select the sub-universe of types satisfying \cppinline{std::regular}, which provides default-construction, copy-semantics, and equality (\cppinline{operator==}).  Equational extensionality on morphisms ($a = b \implies f(a) = f(b)$) is not given by \cppinline{std::regular} alone but follows on the disciplined fragment of Appendix~\ref{app:jlt-systemf} (in particular: no public-API specialisation, no behaviour-branching SFINAE), under which functions cannot inspect type internals to break the read.  In that fragment, \textbf{Objects} behave as sets of values with stable identity.
    \item \textbf{Selection of Morphisms (The Arrows):}
    We select \cppinline{constexpr} lambda expressions on the disciplined fragment.  \cppinline{constexpr} alone does not guarantee purity (captures may be mutable) nor termination (template instantiation is Turing-complete in general); the pure, terminating, effect-free reading is recovered specifically by Appendix~\ref{app:jlt-systemf}'s admissibility rules (no \cppinline{reinterpret_cast}, no SFINAE branching, bounded instantiation depth).  Under this discipline, the selected functions behave as \textbf{Morphisms} satisfying identity and associativity of composition.
    \item \textbf{Construction of Products (Cartesian):}
    The existence of \cppinline{std::pair} and structural binding provides the \textbf{Universal Property of Products}.  For any two objects $A, B$, there exists a product $A \times B$ with projection morphisms $\pi_1, \pi_2$, satisfying the Cartesian requirement.
    \item \textbf{Construction of Exponentials (Closed):}
    First-class lambdas provide the \textbf{Internal Hom-object} ($B^A$): a lambda value \cppinline{[](A) -> B { ... }} reifies an arrow $A \to B$ as a C++ value.  This admits the \textbf{Currying} transformation, mapping $f: A \times B \to C$ to $\hat{f}: A \to C^B$, which on the disciplined fragment realises the bijection $\mathrm{Hom}(A \times B, C) \cong \mathrm{Hom}(A, C^B)$ that closure requires.
    \item \textbf{The Admissibility Gate:}
    The C++23 \cppinline{concept} mechanism acts as the \textbf{type-level admissibility gate}: it mechanically verifies that a given type $T \in \mathbf{Cpp}$ exposes the operator-shape and member-API surface required to inhabit $\mathbf{CCC}$.  This is the C++-side \emph{gate}, not the topos-level subobject classifier $\Omega$ of §\ref{sec:alg-sigma}; the bona fide $\Omega$ machinery lives one layer up in the codebase's \texttt{category:topoi} partition, and \cppinline{concept} is the type-system's complement to it.
\end{enumerate}

We have therefore exhibited a terminal object (\cppinline{void} / \cppinline{std::monostate}), finite products, and exponentials within the disciplined \cppinline{constexpr} subset of C++ defined by Appendix~\ref{app:jlt-systemf}'s admissibility rules; the intersection $\mathbf{CCC} \cap \mathbf{Cpp}$ is inhabited, and the \textbf{Juliet Posture} is the intentional selection of this inhabitant.  The requirements are recapitulated in Table \ref{tab:cpp23-ccc-mapping}.
\end{proof}

\begin{table}[h]
\centering
\caption{The C++23 Realization of Cartesian Closed Category (CCC) Requirements}
\label{tab:cpp23-ccc-mapping}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{C++23 Pillar} & \textbf{Categorical Requirement} & \textbf{Role in the Juliet Posture (\(\mathbf{Jlt}\))} \\ \midrule
\cppinline{std::regular} & \textbf{Objects as Sets} & Default-construction, copy-semantics, equality.  Equational extensionality on the disciplined fragment. \\
\cppinline{constexpr} + admissibility rules & \textbf{Pure Morphisms} & Side-effect-free, terminating composition (purity / termination follow from Appendix~\ref{app:jlt-systemf}, not from \cppinline{constexpr} alone). \\
\cppinline{concept} & \textbf{Type-level admissibility gate} & Filters \(\mathbf{Cpp}\) noise; complement of the topos-level \(\Omega\) (§\ref{sec:alg-sigma}). \\
Lambdas / \cppinline{std::pair} & \textbf{Exponentials / Products} & Reifies the closed and Cartesian properties. \\ \bottomrule
\end{tabular}
\end{table}


We therefore carefully chose elements from C++ guided by two criteria.
First, alignment with the universal-algebra fragment $\mathbf{Alg}(\Sigma)$ identified in §\ref{sec:alg-sigma}.
Second, the desire to integrate functionality the standard library already ships, in particular \cppinline{std::regular} value-types and the primitive carriers (\cppinline{int}, \cppinline{double}, \cppinline{bool}, \dots), without wrapping them in \cppinline{struct} boilerplate or paying a runtime cost.
We can compare and contrast this choice problem with three textbook schools~\cite{pierce2002tapl, cardelli1996typesystems} of type identification:
\begin{itemize}
\item nominal (declared lineage (often inheritance); checked at compile time but blind to behaviour),
\item duck (structural by call-site method probe; checked at runtime, too late for the optimiser),
\item static structural typing (concept-based; behaviour checked at compile time without a declared lineage).
\end{itemize}

C++23 \cppinline{concept}, \cppinline{constexpr}, and \cppinline{std::regular} realise the third school.
We call the resulting discipline the \emph{Juliet Posture}, after the observation that a rose by any other name would smell as sweet~\cite{shakespeare1597romeo}.
The library cares about the \emph{scent} (structural invariants), not the \emph{lineage} (declared name).

% Source: src/main/modules/dedekind/algebra/ring.cppm (concept definition, ll. 110-116)
\begin{listing}[ht]
\caption{The Juliet Posture: structural compile-time matching regardless of nominal provenance.}
\begin{cpp}
// Concept identifying the syntactic ring operators
template <typename T>
concept HasRingOperators = requires(T a, T b) {
  { a + b  } -> std::same_as<T>;
  { a - b  } -> std::same_as<T>;
  { -a     } -> std::same_as<T>;
  { a * b  } -> std::same_as<T>;
};

 // Operators exist on int, but (int, +), not a monoid, because associativity certification has been withheld:
static_assert(HasRingOperators<int>); 
static_assert(!IsAssociative<int, std::plus<int>>); 

// Arithmetic operators are absent on bool but associativity is certified for logical operators:
static_assert(!HasRingOperators<bool>);
static_assert(!IsAssociative<bool, std::logical_and<bool>>);
\end{cpp}
\end{listing}

In practice this boils down to compile-time structural typing across three feature tiers, each strictly more expressive than the last:
(i)~\cppinline{constexpr} branching via template specialisation (the C++<20 baseline);
(ii)~\cppinline{concept}-constrained templates, which lift specialisation from a \emph{compiles-or-doesn't} probe to a named structural predicate (the C++20/23 step the Juliet Posture rests on);
(iii)~developer-supplied claims and axioms via \emph{trait specialisations} (\cppinline{is_associative_v<T, Op>}, \cppinline{is_commutative_v<T, Op>}, \dots) that downstream concepts (\cppinline{IsRing}, \cppinline{IsField}) consult to discharge a single \cppinline{static_assert} obligation.
The first two tiers are mechanically decidable on the disciplined fragment the library polices.
Template instantiation is Turing-complete in general~\cite{veldhuizen2003templates}.
Decidability is recovered by the bounded-instantiation-depth and no-template-template-recursion rules collected in Appendix~\ref{app:jlt-systemf}.
The third tier is the manual-certification path.
The C++ standard distinguishes in [concepts.lib] between \emph{satisfying} a concept (syntactic) and \emph{modelling} it (semantic)~\cite{cpp23standard}.
C++ Core Guideline T.20~\cite{cppcoreguidelines} prescribes manual certification for the semantic-requirements clause the spec deliberately refuses to enforce mechanically.
The standard's \emph{ill-formed, no diagnostic required} clause is its bite, the C++ analogue of \emph{ex falso quodlibet}.
An \texttt{axiom} keyword was considered in early C++0x drafts for in-language semantic certification and ultimately withdrawn~\cite{isocpp_concepts_history}.
The manual-certification pattern is the path the committee left open.\footnote{The C++ standard's \emph{syntactic} / \emph{semantic} split is fundamentally \textbf{decidable vs.\ undecidable} --- what the spec commits to enforcing mechanically vs.\ what it deliberately refuses to enforce --- and does not coincide with the static-vs-dynamic semantics split familiar from PL theory~\cite{pierce2002tapl}.  C++'s \emph{syntactic} bucket subsumes that tradition's \emph{static semantics} (type checking, concept resolution, auto-derived theorems via \cppinline{static_assert}) because those are all things the compiler verifies mechanically; the C++ \emph{semantic} bucket is roughly what resists mechanical decision in the spec's chosen target structure.  The committee's vocabulary optimises for responsibility allocation (compiler vs.\ user) rather than for type-theoretic precision; the mismatch is real but learnable.}

The trait-registry mechanics --- which axioms propagate under which constructors, how per-carrier specialisations are organised, how the \emph{Honest Rejection} policy maps to varieties, and how the algebra-side machinery composes with the C++-side concept layer --- are deferred to §\ref{sec:pruning} (\emph{Algebraic Lattice}) and §\ref{sec:numeric-tension} (\emph{The Traits Registry: Abstract Naturals vs.\ IEEE Numerics}).

Call this disciplined subset $\mathbf{Jlt} \subset \mathbf{Cpp}$ (mnemonically: \emph{Juliet}, after the \emph{rose by any other name} reading of structural-over-nominal typing).
In $\mathbf{Jlt}$ the objects are \cppinline{std::regular} types and the values they classify, and the morphisms are the arrows the language admits between them.
The Juliet Posture deliberately rejects type erasure on the categorical surface: erased witnesses are opaque to both the concept checker and the LLVM optimizer, so erased layers degrade the gate's sharpness and the structural-pruning power that §\ref{sec:compiler-wall} and §\ref{sec:pruning} rest on.

\subsection{A Common Ground: $\mathbf{Ddk} = \mathbf{Jlt} \cap \mathbf{Alg}(-)_{\beth_1}$}
\label{sec:cpp-stlc-mapping}

The \texttt{dedekind} types and combinators are the inhabitants of $\mathbf{Jlt} \cap \mathbf{Alg}(-)_{\beth_1}$.
These are C++23 types in the disciplined fragment $\mathbf{Jlt}$ (§\ref{sec:jlt}) that simultaneously carry universal-algebra structure $\mathbf{Alg}(\Sigma)$ (§\ref{sec:alg-sigma}) for some signature $\Sigma$ the carrier supplies operators for, with the size restriction from §\ref{sec:alg-sigma} carried through.
The dash notation $\mathbf{Alg}(-)_{\beth_1}$ is the family $\bigcup_\Sigma \mathbf{Alg}(\Sigma)_{\beth_1}$ over the signatures the library names.
Per §\ref{sec:alg-sigma}, $\Sigma$ stays unbound here, and the per-shape \cppinline{Has*Operators} concepts pin it later.
The intersection is small and that is the point.

\paragraph{Honest Rejection: the headline policy.}
$\mathbf{Ddk}$ ships with a few carriers certified safe by default but refuses many other things at the gate.
One safe carrier is $\mathbb{B}$ as the Galois field $\mathbb{F}_2$ (exclusive-or for addition and conjunction for multiplication). It has finite cardinality, no rounding, no overflow.
The laws hold trivially.
Others are $\mathbb{N}$, $\mathbb{Z}$, $\mathbb{Q}$.
The latter as \cppinline{Rational<I>} for any wide-enough integer carrier $I$.
Arithmetic on $\mathbb{Q}$ is exact: associativity and distributivity are pinned through the trait registry, and the operator-surface witness \cppinline{HasFieldOperators<Rational<I>>} matches today.
The strict axiomatic \cppinline{IsField} witness on $\mathbb{Q}$ is parked behind an architectural gate (the \cppinline{IsTotal} ladder admits periodic, idempotent, and saturating paths but not yet an exact path); the operator surface is what we ship today, the strict axiomatic certificate is queued.
\cppinline{int} satisfies \cppinline{HasRingOperators} (the four arithmetic operators close on \cppinline{int}), but signed-overflow undefined behaviour defeats associativity.
The strict \cppinline{IsRing} concept therefore refuses it.
IEEE \cppinline{double} satisfies \cppinline{HasFieldOperators}, but rounding non-associativity defeats the field axioms.
\cppinline{IsField} therefore refuses it.
Structural typing (§\ref{sec:jlt}) sees the operator surface mechanically.
The laws are a separate axis the C++ standard library, principled in its silence~\cite{cpp23standard}, refuses to commit to.

% Sources: src/main/modules/dedekind/algebra/galois.cppm (line 335);
%          src/main/modules/dedekind/numbers/integer.cppm (lines ~340--350);
%          src/main/modules/dedekind/category/species.cppm (line 1068).
\begin{listing}[ht]
\caption{Honest Rejection in code: $\mathbf{Ddk}$ ships the Galois field $\mathbb{F}_2$ and the integers $\mathbb{Z}$ as a strict commutative ring safe by default; \cppinline{int} is refused at the gate.  All four static\_asserts compile in $\mathbf{Ddk}$ today.}
\label{lst:honest-rejection}
\begin{cpp}
// 𝔹 = 𝔽₂: bool under (XOR, AND) is the Galois field on two elements —
// exact, physical, finite, no rounding, no overflow.  The strict
// axiomatic IsGaloisField concept binds mechanically (composing the
// axiomatic category::IsField with a finite-cardinality opt-in), with
// no user-side trait specialisation required.
static_assert(IsGaloisField<bool, std::bit_xor<bool>, std::bit_and<bool>>);

// ℤ as default_integer (= SignedExtensionalCardinal<>): the variant
// signed-integer-by-fiat carrier with sign-magnitude periodic wrap —
// no UB, totality via the periodic path through IsTotal.  The strict
// axiomatic IsCommutativeRing chain binds: associativity, commutativity,
// distributivity, additive identity / inverse, and multiplicative
// identity all certified through the species-trait registry.
static_assert(IsCommutativeRing<default_integer,
                                std::plus<default_integer>,
                                std::multiplies<default_integer>>);

// int satisfies the ring operator surface (the four arithmetic
// operators close on int) but fails associativity under signed-overflow
// UB.  The strict axiomatic concepts refuse int at the gate.
static_assert(HasRingOperators<int>);
static_assert(!is_associative_v<int, std::plus<int>>);
\end{cpp}
\end{listing}

\paragraph{The fast lane: opt in on a per-carrier basis.}
A library user pushes the button by writing per-carrier trait specialisations that register \cppinline{is_associative_v<T, Op> = true} (or \cppinline{is_commutative_v}, \cppinline{is_invertible_v}, and so on).
For carriers that already clear the architectural \cppinline{IsTotal} gate (periodic, idempotent, or saturating arithmetic), the trait specialisations are sufficient and the downstream concepts (\cppinline{IsRing}, \cppinline{IsField}) discharge.
Carriers whose totality lives outside that ladder (e.g.\ exact-arithmetic types like \cppinline{Rational<I>}, or signed integer types whose overflow is UB) need an additional architectural lift on \cppinline{IsTotal} before the strict concepts will match.
The user has audited their inputs to a regime where the law actually holds.
Examples are integer arithmetic on a domain where overflow cannot occur, IEEE \cppinline{double} accepted with associativity-by-fiat for a specific application, or a custom carrier whose modular wrap-around the user proves consistent with the surrounding logic.
The opt-in is the user's pinky-swear in their part of the multiverse.
They own the proof that the swear is honest.
IFNDR is the standard's bite if the swear was a lie.
This is the operational discharge of the satisfies/models distinction §\ref{sec:jlt} introduces.
Closure on the operator surface is what \cppinline{Has*Operators} records mechanically.
Lawfulness is what the trait registry records intentionally.
The full mechanics live in §\ref{sec:pruning} (\emph{Algebraic Lattice}).
That section covers which axioms propagate under which constructors, how per-carrier specialisations are organised, and how the \emph{Honest Rejection} default composes with the universal-algebra reading $\mathbf{Mon} \subset \mathbf{Var}(\Sigma) \subset \mathbf{Alg}(\Sigma) \subset \mathbf{Alg}(-) \subset \mathbf{Cat}$~\cite{burris1981universalalgebra} (§II.10--II.11).
The paper's $\mathbf{Mon}$-side framing is in §\ref{sec:alg-sigma}.

\paragraph{Two consequences worth naming.}
First, structural typing makes membership definable \emph{after the fact}.
A pre-existing standard-library type such as a \cppinline{std::pair} or a \cppinline{std::optional} becomes an inhabitant of $\mathbf{Ddk}$ the moment a \cppinline{concept} accepting it is written, with no modification to the carrier; likewise, a standard-library range type satisfying \cppinline{std::ranges::view} can be admitted the same way.
Nominal-typing alternatives (declared lineage, base classes, virtual interfaces) would require carriers shipped in the C++ standard a decade ago to opt in upstream, which is impossible.
Second, the disciplined fragment in which $\mathbf{Ddk}$ inhabits also reads as a faithful image of the simply-typed $\lambda$-calculus and System~F under Lambek--Scott~\cite{lambek1988introduction} and Pierce~\cite{pierce2002tapl}.
This is a verification artefact rather than a flow-relevant claim, but worth naming for the reader who wants to chase the bridge.
Details of this observation are discussed in Appendix~\ref{app:jlt-systemf}.
The row-by-row C++23 $\to$ STLC mapping is in Appendix~\ref{app:ddk-rosetta}, alongside the operator-shape rosetta and the textbook commutative-square diagram for Listing~\ref{lst:isfunctor} below.

\paragraph{Higher-kinded inhabitants.}
$\mathbf{Ddk}$'s residents are not only scalar carriers like the finite fields and exact-arithmetic types of Listing~\ref{lst:honest-rejection}.
The intersection extends to higher-kinded structure --- functors, monads, and other arrow-shaped constructions --- when the same Juliet-Posture concept-gate matches on a hub-and-spoke type that names the structure (§\ref{sec:alg-sigma}'s hub-and-spoke architectural reading at the next level up).
Listing~\ref{lst:isfunctor} exhibits the canonical example: an \cppinline{IsFunctor} concept encoding the textbook functor laws, paired with a worked \texttt{maybe\_functor} hub (Listing~\ref{lst:maybe_functor}) that promotes \cppinline{std::optional} into a functor without modifying the carrier.

% Source: src/main/modules/dedekind/category/functor.cppm (IsFunctor concept + maybe_functor instance, abridged)
\begin{listing}[ht]
\caption{The Functor concept, abridged from \texttt{dedekind.category:functor}. A worked example is listed in~\ref{lst:maybe_functor}. The textbook commutative-square that the concept body encodes is in Appendix~\ref{app:ddk-rosetta}.}
\label{lst:isfunctor}
\begin{cpp}
template <typename F>
concept IsFunctor = IsArrow<F> && requires {
  typename F::Σ_cat;                         // source category
  typename F::Τ_cat;                         // target category
  typename F::template Shape<int>;           // object map  A -> F(A)
} && requires(F f, typename F::Σ_cat::Arrow f_c) {
  { f.φ(f_c) } -> IsArrow;                   // morphism lift  f -> F(f)
  // Domain / Codomain of the lifted arrow track the object map:
  // Dom(F(f)) = F(Dom(f)),  Cod(F(f)) = F(Cod(f))
  // (full requires-clause elided; see source).
};
}
\end{cpp}
\end{listing}

\begin{listing}[ht]
\caption{A worked functor instance (Maybe), abridged from \texttt{dedekind.category:functor}.  The functorial obligation lives in the \emph{hub} type \cppinline{maybe_functor<T>} (§\ref{sec:alg-sigma}'s hub-and-spoke convention), not in the carrier \cppinline{std::optional<T>}.  The static\_assert below is the Curry--Howard discharge that \texttt{maybe\_functor} inhabits the concept.}
\label{lst:maybe_functor}
\begin{cpp}
// The optional-value Maybe functor: Set<T> -> Set<std::optional<T>>.
template <typename T>
struct maybe_functor {
  // IsArrow surface (the hub's outer typing).
  using ArrowKind = hub_arrow_tag;
  using Σ_cat = Set<T>;                      // source category
  using Τ_cat = Set<std::optional<T>>;       // target category
  using Domain   = Σ_cat;                    // hub Domain  = source category
  using Codomain = Τ_cat;                    // hub Codomain = target category
  constexpr Τ_cat operator()(const Σ_cat&) const noexcept { return {}; }

  // Functorial action.
  template <typename U> using Shape = std::optional<U>;
  template <typename Fn>           // Fn = the functorial action arrow
    requires IsArrow<std::remove_cvref_t<Fn>>
  constexpr auto φ(Fn&& f) const {
    return arrow([f = std::forward<Fn>(f)](std::optional<T> const& m) {
      return m.has_value() ? std::optional{std::invoke(f, *m)} : std::nullopt;
    });
  }
};

static_assert(IsFunctor<maybe_functor<int>>);  // mechanical verification
\end{cpp}
\end{listing}

The hub--carrier split (\cppinline{maybe_functor<T>} carries the functorial proof obligation; \cppinline{std::optional<T>} stays a Plain Old C++ Object) is the §\ref{sec:alg-sigma} architectural pattern played at the next kind level.
This is the same after-the-fact membership above: \cppinline{std::optional} ships in the C++ standard and cannot opt into anything upstream, but the moment a hub naming the functorial structure passes the concept gate, the carrier inhabits $\mathbf{Ddk}$ as a functor.

The carrier-lattice's higher-kinded constructions are populated by several functorial families, each a textbook universal-property construction; representative anchors include:
\emph{Frac} (field of fractions: a left adjoint to the inclusion $\mathbf{Field} \hookrightarrow \mathbf{IntegralDomain}$, giving $\mathbb{Q} = \operatorname{Frac}(\mathbb{Z})$, with $\operatorname{Cuts}$ as a sister carrier-strength completion giving $\mathbb{R} = \operatorname{Cuts}(\mathbb{Q})$);
\emph{Free} (free module: a left adjoint to the forgetful $U: \mathrm{Mod}_R \to \mathbf{Set}$, giving $V_n(R) = \operatorname{Free}_n(R) = R^n$ and dually the internal hom $\operatorname{End}_R(V_n) = M_n(R)$);
and \emph{Quotient} (quotient by a congruence: an H operation in the HSP sense, giving $\mathbb{C} = \mathbb{R}[i]/(i^2+1)$ and $\mathbb{D} = \mathbb{R}[\varepsilon]/(\varepsilon^2)$).
Each construction is named in the carrier-lattice diagram (Figure~\ref{fig:carrier-lattice}) and exhibited concretely in §\ref{sec:linalg}; the discipline of this section is that each functor lifts at the type level the moment a hub naming it passes the concept gate, and the corresponding carrier (\cppinline{Rational<I>}, \cppinline{Vec2V<T>} / \cppinline{Matrix2x2V<T>}, \cppinline{Complex<R>} / \cppinline{Dual<F>}) inhabits $\mathbf{Ddk}$ at that kind level without modifying the underlying scalar storage.

\paragraph{The intersection's deliberate scope.}
Two distinct mechanisms compose into the intersection $\mathbf{Ddk} = \mathbf{Jlt} \cap \mathbf{Alg}(-)_{\beth_1}$, and naming them separately matters because they pay for different things.
The $\beth_1$ ceiling from §\ref{sec:alg-sigma} is \emph{foundational}: it stops the encoding from courting the Russell-style self-reference of a meta-category-of-everything, the same way ZFC's missing set-of-all-sets is shaped to forbid such an object.
The no-type-erasure stance from §\ref{sec:jlt} is \emph{mechanical}: it preserves the concrete type-level information the concept checker and the LLVM backend rely on, and is what buys the compile-time decidability and pruning power the rest of the paper trades on.
The visible cost is what neither mechanism gives away for free: meta-categorical statements (the textbook $\mathrm{Disc} \dashv U$, the size-hierarchy adjunctions, free / forgetful pairs across $\mathbf{Cat}$) cannot land at the type level, since they range over the proper-class meta-categories $\mathbf{Set}$ and $\mathbf{Cat}$.
They survive in $\mathbf{Ddk}$ as documented prose-level CT vocabulary, anchored where possible by mechanically-witnessed \emph{restrictions} to a single small slice.
A representative anchor: the $\mathrm{Disc} \dashv U$ adjunction between $\mathbf{Set}$ and $\mathbf{Cat}$ is referenced in prose in \texttt{category:etcs}, but the bona fide \cppinline{HasAdjunctionShape} / \cppinline{IsAdjunction} witnesses match on the discrete-restriction representatives \cppinline{disc_self_endofunctor_t<C>} and \cppinline{disc_self_unit_t<C>} reified in \texttt{category:natural}, where $F = U = \mathrm{Id}$ and $\eta = \epsilon$ degenerate to the identity natural transformation on a single $C$.
The 2-categorical layer above thus becomes the engineer's honesty obligation, named by reference rather than by type-level proof --- a price the encoding pays in exchange for the foundational soundness of $\beth_1$ and the gate-sharpness of no-erasure, and one §\ref{sec:compiler-wall} revisits at the wall where the type checker also stops.

%\clearpage  % Flush §2.3's deferred floats (Listing 3 in particular) before §3.

% ============================================================
\section{Sets and Sequences}
\label{sec:sets-rules}
% ============================================================

% ============================================================
\subsection{Intensional First, Realize When You Mean It}
\label{sec:intensional}
% ============================================================

\subsubsection{The Design Rule}

The central methodology of \texttt{dedekind} is captured by a single rule:
\begin{quote}
  \emph{Intensional first, realize it when you mean it.}
\end{quote}
The thrust of this rule is to leverage \emph{lazy evaluation}---both at the type level
(compile-time symbolic substitution) and at the value level (deferred numeric approximation)---as
a principled defence against \emph{Premature Materialization}: the loss of mathematical
structure that occurs when a computation is forced to a concrete representation earlier than
necessary.

An \emph{intensional} description specifies a mathematical object by its defining properties.
An \emph{extensional} representation enumerates its members or computes its value.
A halfspace \(\{x \in \mathbb{Q} \mid x \leq 4\}\) is intensional; the list of its
rational points between $0$ and $4$ is extensional; and a function that materialises
the former into the latter is the auditable crossing of that boundary. The Dedekind cuts
of Section~\ref{sec:dedekind-cuts}---where a real number is named by the pair of rational
sets below and above it---push the same distinction to its limit: the real is
intensional forever, and only bounded-precision queries cross the boundary.
This framing is close in spirit to intensional database semantics (for example,
Majki\'c's intensional FOL treatment for peer-to-peer data integration
\cite{majkic2009intensional}), but our engineering contrast with mainstream SQL
is explicit: SQL evaluation is typically defined over already-realized relations,
while \texttt{dedekind} preserves predicate-level structure in the type system and
defers materialization to an explicit realization boundary.

The methodological lineage also runs through Abelson and
Sussman's~\cite{abelson1996sicp} \emph{programming by wishful thinking}: name
the procedure you wish you had, use it, and supply its definition only when the
surrounding code forces a concrete answer.  SICP's medium for this discipline
is Scheme---an untyped lambda calculus where the wished-for procedure is
\emph{notational} until evaluation reaches it.  Our medium is the C++23 type
checker, which performs a typed lambda calculus on our behalf at compile time:
the wished-for object inhabits a predicate-set type, the wishes are tracked
structurally as type-level constraints, and the realization boundary is the
single point at which the type system permits an extensional value to be
produced.  Where SICP defers \emph{evaluation} (a runtime move), we defer
\emph{materialization} (a compile-time-to-runtime move), and the type checker
is the auditor that ensures the wishes were coherent before the realization
happens.

\paragraph{Forms before carriers.}
The library reifies the discrete-foundations triple structurally
rather than starting from a machine-integer carrier:
$\mathbb{B}$ as the Subobject Classifier $\Omega$ (ETCS
Axiom~7), $\mathbb{N}$ as Lawvere's Natural Numbers Object (NNO;
ETCS Axiom~9~\cite{lawvere1964elementary}), and $\mathbb{Z}$ via
two universal properties on a single carrier --- initial object in
$\mathbf{CRing}$ and free abelian group on the commutative
monoid $(\mathbb{N}, +, 0)$ (the Grothendieck construction; see
Mac Lane~\cite{maclane1971categories} for the standard
treatment).\footnote{Each Form sits behind a three-layer chain
\emph{Form $\to$ Carrier $\to$ Symbol}: a universal-property
concept (\cppinline{IsNNO}, \cppinline{IsInitialRing},
\cppinline{IsGrothendieckGroup}), a canonical inhabitant
(\cppinline{Cardinality} for $\mathbb{N}$ with $\aleph_0$
saturation; \cppinline{SignedCardinality} for $\mathbb{Z}$ with
$\pm\aleph_0$ and \cppinline{NaZ}), and the practitioner-facing
alias.
The negafibonacci recurrence $F_{n-1} = F_{n+1} - F_n$ is the
canonical exhibit: the backward step requires exactly the
operator (subtraction) that distinguishes $\mathbb{Z}$ from
$\mathbb{N}$, and the variant carriers escalate to
$\pm\aleph_0$ rather than overflowing as
\cppinline{int}/\cppinline{unsigned int} would.}
The same Form-then-Carrier discipline continues up the
tower.\footnote{$\mathbb{Q} = \operatorname{Frac}(\mathbb{Z})$,
$\mathbb{R} = \operatorname{Cuts}(\mathbb{Q})$,
$\mathbb{C} = \operatorname{Cplx}(\mathbb{R})$,
$\mathbb{D} = \operatorname{Dual}(\mathbb{R})$, and the rank lift
$\operatorname{Free}_n(R) = R^n$ /
$\operatorname{End}_R(R^n) = M_n(R)$ are named as functors and
itemised in the carrier-rosetta tables in
§\ref{sec:carrier-rosetta} of the appendix; carriers are named by
the universal property that defines them, not by the storage
scheme they happen to use.}

\subsection{The Comprehension View: Sets Are Rules, Not Buckets}

Here is the move. A set is not a bucket. A set is a rule:
\[
  S = \{x \in A \mid P(x)\}.
\]
The right-hand side is the definition most mathematicians write down first and then
silently forget, reaching instead for a list or a \texttt{std::vector}. In
\texttt{dedekind} we take that right-hand side literally. The set $S$ is the
\emph{type} whose inhabitants are elements of the ambient species $A$ satisfying
$P$. Nothing is enumerated, nothing is allocated, nothing is stored. The
predicate \emph{is} the set.

That shift---what Lawvere called the comprehension view~\cite{lawvere1964etcs,
lawvere2003sets}---is the one idea this section needs. Once you accept it,
everything downstream (intersection, complement, emptiness, pruning) falls out
as predicate algebra. The subobject classifier $\Omega$ becomes a knob you can
turn: classical \texttt{bool} for exact domains, Kleene K3 for floats, anything
else you can register.

\subsection{Set Operations as Predicate Composition}

If a set is a predicate, a set operation is whatever operation you would apply to
the predicates. The dictionary writes itself:

\begin{align*}
  S_1 \cap S_2 &\triangleq \{x \in A \mid P_1(x) \land P_2(x)\} \\
  S_1 \cup S_2 &\triangleq \{x \in A \mid P_1(x) \lor P_2(x)\} \\
  \overline{S_1}  &\triangleq \{x \in A \mid \neg P_1(x)\}
\end{align*}

Every operation on the left is an operation on $\Omega$ on the right. Intersection
is conjunction; union is disjunction; complement is negation. No values are touched,
because there are no values yet. The combinators compose at compile time and produce
a new type---a new rule---which the compiler is then free to inspect.

\subsection{Structural Pruning in Practice}

The payoff of the comprehension view is that the LLVM backend can \emph{prune} a
set expression to a no-op when its predicate is statically unsatisfiable.

% Source: src/test/cpp/modules/dedekind/python/showcase_03_halfspace_contradiction.cpp
\begin{listing}[htbp]
\caption{Contradictory predicates collapse to the empty-set type $\varnothing$. Uses the \cppinline{Halfspace}-NTTP DSL to pin both bounds in the type signature, so the contradiction is resolved at translation time.}
\label{lst:pruning}
\begin{cpp}
constexpr auto n = element<ℕ>;

// Two halfspace-structured sets, with pivots (5 and 3) carried as NTTPs.
constexpr auto gt_5 = Set{n | (n > bound<5>)};
constexpr auto lt_3 = Set{n | (n < bound<3>)};

// Compile-time theorem: the meet IS the empty set on ℕ.
constexpr Ø<Cardinality> empty_meet = gt_5 & lt_3;
static_assert(empty_meet == Ø{});
\end{cpp}
\end{listing}

The generated binary contains no membership-test code for
\cppinline{empty_meet}: the type-level contradiction rule resolves the
result to $\varnothing$ \emph{before} code generation, and the
computability tier of $\varnothing$ is fully top
(extensional/classical/element-in-type).
Listing~\ref{lst:pruning_ir} makes this mechanically observable: the
witness function for set membership constant-folds to \texttt{ret i1
false} in the emitted LLVM IR.

% Source: src/test/cpp/modules/dedekind/python/showcase_03_halfspace_contradiction.ll
\begin{listing}[htbp]
\caption{LLVM IR for the empty-meet witness in Listing~\ref{lst:pruning} (\texttt{clang++-22 -O2}): the membership query constant-folds to \texttt{ret i1 false}.  The complementary cardinality-1 case (singleton collapse) and other halfspace reductions are exercised in \texttt{pruning\_showcases\_test.cpp} but not repeated here --- one source / IR pair carries the structural-pruning claim.}
\label{lst:pruning_ir}
\begin{cpp}
define noundef zeroext i1 @witness_empty_halfspace_meet() local_unnamed_addr #0 {
  ret i1 false
}
\end{cpp}
\end{listing}

\subsection{Algebraic Properties Enable Stronger Pruning}

Beyond interval arithmetic, \emph{algebraic} properties of the predicates provide
additional pruning opportunities.
The \texttt{:species} registry records which operations are associative, commutative,
idempotent, and so forth.
These traits inform the compiler about structural relationships between sets:

\begin{itemize}
  \item $S \cap \varnothing = \varnothing$ (intersection with the empty set is empty).
  \item $S \cap \overline{S} = \varnothing$ (intersection with the complement is empty).
  \item If $S_1 \subseteq S_2$ then $S_1 \cap S_2 = S_1$ (subset absorption).
\end{itemize}

Each of these equalities can be discharged as a \cppinline{static_assert} once the
relevant algebraic witnesses are in the trait registry, reducing the intersection to
the simpler set at compile time.


% ============================================================
\section{Algebraic Lattice}
\label{sec:pruning}
% ============================================================

% TODO(future-edit): this section is the natural home for the deeper
% trait-registry discussion that §\ref{sec:jlt} (Juliet Posture) only
% gestures at.  The framing established there is the two-axis split:
%   (i)  closure on the operator surface --- validated by inspecting
%        the standard library (overload resolution, concept binding);
%   (ii) algebraic laws (associativity, commutativity, distributivity,
%        identities, inverses, totality) --- validated and certified
%        by hand, per carrier, in the trait registry.
% This section should expand axis (ii) with: how the registry is
% organised (per-axiom variable templates like is_associative_v),
% how propagation works (HSP under quotients/products; partial
% propagation under End_R), and how the audit rosettas in
% §\ref{sec:carrier-rosetta} measure the standard library's silence.
% The §\ref{sec:species} listing already exhibits the registry style;
% the work is integrating it as a coherent narrative thread rather
% than a standalone listing.  Keep §3 lean --- if registry detail
% accretes there, push it down here.

% TODO(future-edit, draft figure): Figure~\ref{fig:two-axes} below is a
% first-pass sketch of the two-axis split.  Disposition for the editing
% pass: this figure is primarily an EDITING-TIME MNEMONIC --- a thinking
% aid that helps keep the §\ref{sec:jlt}/§\ref{sec:pruning} narrative
% honest about which axis any given paragraph is talking about.  It may
% or may not earn its place at submission time; the decision is downstream
% of how the surrounding registry prose lands.  Three ways it could go:
%   (i)  load-bearing --- promoted to a numbered figure that the body
%        prose explicitly cites in both §\ref{sec:jlt} and §\ref{sec:pruning};
%   (ii) folded --- absorbed into the rosetta tables (§\ref{sec:carrier-rosetta},
%        Tables~\ref{tab:rosetta-quotient}, \ref{tab:rosetta-rank}) as a
%        compact column/row legend instead of a standalone figure;
%   (iii) dropped --- the rosettas + the §\ref{sec:jlt} prose paragraph
%        carry the framing without the figure.
% The carrier placements are indicative, not audited --- the authoritative
% data lives in the carrier rosettas.  Other open questions for the polish pass:
%   (a) y-axis as a discrete law-stack (closure / +assoc / +comm / +id /
%       +inv) as drawn, or a 2D law matrix (one axis per law) with carriers
%       as cells?  The stack reads cleaner; a matrix is more honest about
%       non-monotone cases (matrix mult: associative AND non-commutative).
%   (b) cross-link to the §\ref{sec:species} registry listing so the
%       reader sees the figure's "above this line" zone realised in code.
%   (c) third axis (universe vs. classifier).  §\ref{sec:jlt}'s closing
%       footnote names this axis: Ω<T> is the carrier-typed universe (the
%       constant-true predicate "T as its own set"); <Tower>Of<> is the
%       ETCS characteristic morphism χ_T : tower-ambient → Ω of T as a
%       subobject of its tower (e.g.\ NaturalNumbersOf rejects negatives
%       on int via the canonical ℕ ↪ ℤ inclusion).  For the bottom of the
%       tower (bool) the two collapse; for ℕ/ℤ/ℚ/ℝ/ℂ/𝔻 they are distinct
%       exports.  Two readings of how to add this to the figure:
%         - depth-axis: extrude the 2D plot into 3D (universe / classifier
%           plane).  Probably visually overloaded; not recommended.
%         - secondary diagram in §\ref{sec:pruning}: a small commutative
%           square showing universe Ω<T> vs.\ classifier χ_T side by side
%           with the tower embedding ℕ ↪ ℤ as the connecting arrow.
%           Lighter touch and reads as a textbook ETCS exhibit.
%       Decision deferred until rosetta polish — if §\ref{sec:pruning}
%       prose ends up needing visual support for the universe/classifier
%       distinction, the secondary-diagram option is the lower-cost
%       addition.
\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  >={Stealth[length=2.5mm]},
  carrier/.style={circle, fill=black, inner sep=1.5pt},
  every node/.style={font=\small},
]
  % background bands
  \fill[gray!8] (-0.05,-0.05) rectangle (10.4,1.05);
  \fill[blue!4] (-0.05,1.05) rectangle (10.4,5.6);
  \node[anchor=west, font=\footnotesize\itshape, gray] at (5.2,0.5) {std::library territory (overload resolution)};
  \node[anchor=west, font=\footnotesize\itshape, blue!50!black] at (5.2,5.3) {trait registry territory (certified by hand)};

  % axes
  \draw[->, thick] (0,0) -- (10.2,0) node[anchor=west] {axis 1: operator surface};
  \draw[->, thick] (0,0) -- (0,5.5) node[anchor=south] {axis 2: algebraic laws};

  % horizontal ticks
  \foreach \x/\lbl in {1.5/{none}, 4.5/{\texttt{HasRingOperators}}, 8/{\texttt{HasFieldOperators}}} {
    \draw[thick] (\x,-0.1) -- (\x,0.1);
    \node[anchor=north] at (\x,-0.15) {\lbl};
  }

  % vertical ticks
  \foreach \y/\lbl in {1/{closure only}, 2/{+ associativity}, 3/{+ commutativity}, 4/{+ identities}, 5/{+ inverses}} {
    \draw[thick] (-0.1,\y) -- (0.1,\y);
    \node[anchor=east] at (-0.15,\y) {\lbl};
  }

  % carriers
  \node[carrier, label={[label distance=0pt]above:{\texttt{int}}}] at (4.5,1) {};
  \node[carrier, label={[label distance=0pt]above:{\texttt{unsigned}}}] at (4.5,3) {};
  \node[carrier, label={[label distance=0pt]above:{\texttt{double}}}] at (8,1) {};
  \node[carrier, label={[label distance=0pt]above:{\texttt{Rational<int>}}}] at (8,5) {};
  \node[carrier, label={[label distance=0pt]above:{\texttt{Matrix<R>}}}] at (4.5,2) {};

  % silence demarcation
  \draw[densely dashed, gray!70!black, thick] (-0.05,1.05) -- (10.4,1.05);
\end{tikzpicture}
\caption{\textbf{The two axes of the trait registry (draft).}  Axis 1 (horizontal) is the operator surface: which types compose under which operators.  This axis is validated by \emph{inspection} of the standard library --- overload resolution and concept binding answer it without further human input.  Axis 2 (vertical) is the algebraic-law stack: closure-only, then $+$ associativity, $+$ commutativity, $+$ identities, $+$ inverses.  Above the dashed line the standard library is silent: each law must be \emph{certified by hand}, per carrier, in the trait registry of §\ref{sec:species}.  Carrier placements illustrate the typical shape of the silence: \texttt{int} closes the ring surface but signed-overflow UB defeats associativity (\textsc{ub}); \texttt{unsigned} reaches commutativity via modular arithmetic; \texttt{double} closes the field surface but IEEE-754 rounding defeats associativity; \texttt{Rational<int>} reaches the full field stack; matrix multiplication is associative without being commutative (a non-monotone point this stacked axis under-resolves --- see TODO note in source).  The placements are indicative; the audited data lives in the carrier rosettas (Tables~\ref{tab:rosetta-quotient}, \ref{tab:rosetta-rank}).}
\label{fig:two-axes}
\end{figure}

The third pillar.  The carrier lattice of Figure~\ref{fig:carrier-lattice}
in load-bearing form: HSP propagation lifts axioms componentwise
along quotients (\textbf{H}, e.g. \texttt{Complex},
\texttt{Dual}) and direct products (\textbf{P}, e.g. $V_n(R)$,
\texttt{Diagonal}); localizations such as \texttt{Frac} sit alongside
HSP rather than inside it, and $\operatorname{End}_R$ is the
partial-axiom rank-$n$ extension.  The algebraic lattice's compile-time-reduction
showcases live in the sister section §\ref{sec:linalg}
\emph{Linear Algebra}: linear programming
(§\ref{sec:lp-centrepiece}), forward-mode automatic differentiation
(§\ref{sec:ad-elliott}), and the
$\mathbb{C} / \mathbb{D} \hookrightarrow M_2(\mathbb{Q})$
embedding (§\ref{sec:lin-alg-embedding}).  Each of these is the
algebraic lattice acting through the disciplined
fragment of §\ref{sec:axiomatic}.


\paragraph{Reading Figure~\ref{fig:carrier-lattice}.}
The Algebraic Lattice is a 2D oblique projection of a 3D cube ---
eight corners, twelve edges, three Fraktur-letter axes
($\mathfrak{A}$ carrier-strength, $\mathfrak{R}$ rank,
$\mathfrak{Q}$ quotient).  $\mathbb{B}$ sits at the bottom-front-left;
$M_2(X(\mathbb{C}))$ at the top-back-right.

Where the carrier axis stops is partly conventional.  We place
$\mathbb{C}$ at the front-right corner because the textbook chain
ends there, but the same diagram with $\mathbb{R}$ at the corner would
push $\mathbb{C}$ to a back-face node $C(\mathbb{R}) =
\mathbb{R}[i]/(i^2 + 1)$ via the quotient axis; with $\mathbb{Q}$ at
the corner, $\mathbb{R}$ joins it via $\mathrm{Cuts}$.  All three
readings agree on the math; the carrier-vs-quotient line is partly
interpretive
(§\ref{sec:carrier-rosetta}).\footnote{A parallel \emph{computability}
lattice (representation Int / Ext; logic K3 / Cl; knowledge Opq / Elt)
sits orthogonal to the three axes drawn --- the algebraic tower says
\emph{which algebra}; the computability tier says \emph{how much the
compiler knows about a set built on it}
(§\ref{sec:gap}, Definition~\ref{def:computability-tier}).  We don't
draw the second cube; one 3D figure carries the weight.}

The lattice is unbounded above.  Iterated quotients, tensor products,
$p$-adic completions and similar each yield a strictly larger object
without a canonical maximum, so the visual top-right-back corner is
an illustrative stopping point and the rendering is necessarily a
finite slice.\footnote{Binary meets (largest shared substructure) are
canonical; binary joins (smallest common extension) exist as pushouts
but require ambient choice, so the structure is better thought of as
a bounded-below lattice (informally, a meet-semilattice with
constructive joins).  Other dimensions the figure deliberately omits:
higher rank ($n > 2$; \#368), predicate-set / order-theoretic
constructions (candidate sibling \emph{Geometric Tower}; future
Figure~1b), iterated $\mathfrak{Q}$ (jet bundles; \#504), and the
Honest-Rejection machine-substrate boundary (\texttt{bool},
\texttt{uint}, \texttt{int}, \texttt{double} embed into
$\mathbb{B} / \mathbb{N} / \mathbb{Z} / \mathbb{Q}$ via named
\texttt{embed\_*} arrows in \texttt{dedekind.numbers}).  Per-cell
compile-status of any specific instance is in the trait registry
(\#499); the figure carries the structural shape.}

The three axes don't all commute: matrix and quotient functors
commute when both sides are defined; different quotient functors
commute when their generating variables are independent (so
$D \circ C \cong C \circ D$); but $\mathfrak{A}$ steps can fail to
type-check after $\mathfrak{R}$ (producing modules, not rings) or
after a quotient that destroys an integral domain (so
$\mathrm{Frac}(D(R))$ is undefined).  Worth knowing on first read; the
specific cells fall out of the type system at the call site.

\begin{figure}[p]
\centering
\begin{tikzpicture}[
  scale=1.0,
  every node/.style={font=\small},
  baseline=(current bounding box.center)
]
  % --- Cube-corners-only schematic of the Algebraic Lattice ---
  %   X axis (right):  carrier strength 𝔄  (Frac, Cuts, Cplx — see Table A)
  %   Y axis (up):     rank             𝔯  (Free_n, End_R    — see Table B)
  %   Z axis (back):   quotient         𝔔  (D, C, D∘C, ...   — see Table C)
  \def\dx{4.0}
  \def\dy{3.0}
  \def\zxoff{2.5}
  \def\zyoff{2.0}

  % Front face (z = 0) corners.
  \node (B)    at (0,    0)    {$\mathbb{B}$};
  \node (C)    at (\dx,  0)    {$\mathbb{C}$};
  \node (M2B)  at (0,    \dy)  {$M_2(\mathbb{B})$};
  \node (M2C)  at (\dx,  \dy)  {$M_2(\mathbb{C})$};

  % Back face (z = 1) corners.
  \node[opacity=0.7] (XB)    at (\zxoff,         \zyoff)         {$X(\mathbb{B})$};
  \node[opacity=0.7] (XC)    at (\dx + \zxoff,   \zyoff)         {$X(\mathbb{C})$};
  \node[opacity=0.7] (M2XB)  at (\zxoff,         \dy + \zyoff)   {$M_2(X(\mathbb{B}))$};
  \node[opacity=0.7] (M2XC)  at (\dx + \zxoff,   \dy + \zyoff)   {$M_2(X(\mathbb{C}))$};

  \tikzset{
    edge/.style={->, >=stealth, semithick},
    edgeback/.style={->, >=stealth, semithick, opacity=0.55},
    diag/.style={->, >=stealth, dashed, thick}
  }

  % X-axis edges (4): "carrier strength" 𝔄 — Frac / Cuts / Cplx; see Table A.
  \draw[edge]     (B)    -- node[midway, fill=white, inner sep=2pt]
                              {$\mathfrak{A}$} (C);
  \draw[edge]     (M2B)  -- node[midway, fill=white, inner sep=2pt]
                              {$\mathfrak{A}$} (M2C);
  \draw[edgeback] (XB)   -- node[midway, fill=white, inner sep=2pt, opacity=0.7]
                              {$\mathfrak{A}$} (XC);
  \draw[edgeback] (M2XB) -- node[midway, fill=white, inner sep=2pt, opacity=0.7]
                              {$\mathfrak{A}$} (M2XC);

  % Y-axis edges (4): "rank" 𝔯 — Free_n / End_R; see Table B.
  \draw[edge]     (B)    -- node[midway, fill=white, inner sep=2pt]
                              {$\mathfrak{R}$} (M2B);
  \draw[edge]     (C)    -- node[midway, fill=white, inner sep=2pt]
                              {$\mathfrak{R}$} (M2C);
  \draw[edgeback] (XB)   -- node[midway, fill=white, inner sep=2pt, opacity=0.7]
                              {$\mathfrak{R}$} (M2XB);
  \draw[edgeback] (XC)   -- node[midway, fill=white, inner sep=2pt, opacity=0.7]
                              {$\mathfrak{R}$} (M2XC);

  % Z-axis edges (4): "quotient" 𝔔 — D / C / D∘C / ...; see Table C.
  \draw[diag] (B)   -- node[midway, sloped, above, font=\scriptsize]
                          {$\mathfrak{Q}$} (XB);
  \draw[diag] (C)   -- node[midway, sloped, above, font=\scriptsize]
                          {$\mathfrak{Q}$} (XC);
  \draw[diag] (M2B) -- node[midway, sloped, above, font=\scriptsize]
                          {$\mathfrak{Q}$} (M2XB);
  \draw[diag] (M2C) -- node[midway, sloped, above, font=\scriptsize]
                          {$\mathfrak{Q}$} (M2XC);

  % --- Axis-direction labels ---
  \node[below=12pt, font=\itshape\small] at (\dx/2, 0)
    {carrier-strength $\mathfrak{A}$};
  \node[left=12pt, font=\itshape\small, rotate=90] at (0, \dy/2)
    {rank $\mathfrak{R}$};
  \node[font=\itshape\small, rotate=30, opacity=0.85]
    at ($(B)!0.5!(XB) + (-0.5, 0.5)$)
    {quotient $\mathfrak{Q}$};
\end{tikzpicture}
\caption{\emph{The Algebraic Lattice.}
3D oblique-projection lattice; eight cube corners, three Fraktur-axis
parameters.  $\mathfrak{A}$ (carrier-strength) follows the chain
$\mathbb{B} \to \mathbb{N} \to \mathbb{Z} \to \mathbb{Q} \to \mathbb{R}
\to \mathbb{C}$; the early steps use the natural-numbers-object
($\mathbb{B} \to \mathbb{N}$) and the Grothendieck ring construction
($\mathbb{N} \to \mathbb{Z}$), the later steps the
\emph{Frac} / \emph{Cuts} / \emph{Cplx} functors
($\mathbb{Z} \to \mathbb{Q} \to \mathbb{R} \to \mathbb{C}$;
Tables~\ref{tab:carriers-exact}--\ref{tab:rosetta-textbook-vs-cpp}).
$\mathfrak{R}$ (rank) lifts Scalar $\to V_n(R) \to M_n(R)$ via
$\mathrm{Free}_n$ and $\mathrm{End}_R$ (Table~\ref{tab:rosetta-rank}).
$\mathfrak{Q}$ (quotient) realises an extension $R \to X(R)$ at a
polynomial ideal --- the degree-2 instances dominate the worked
showcases: $X = D$ gives $R[\varepsilon]/(\varepsilon^2)$ (dual
numbers), $X = C$ gives $R[i]/(i^2 + 1)$ (complex extension); higher
degrees enter via $X = G$, $R[x]/(p(x))$ for $p$ irreducible (Galois
extension; e.g.\ $\mathbb{F}_{64} = \mathbb{F}_2[x]/(x^6 + x + 1)$
at degree~6).  See Table~\ref{tab:rosetta-quotient} for the full
enumeration.
The bottom-front-left corner $\mathbb{B}$ is the lattice bottom; the
top-right-back corner $M_2(X(\mathbb{C}))$ is an illustrative
stopping point on an unbounded-above lattice.  The reading aids,
commutation rules, off-figure dimensions, and the bounded-below /
no-unique-top caveat are discussed in the body prose immediately
preceding this figure.}
\label{fig:carrier-lattice}
\end{figure}

% --- Rosetta tables for the rank ($\mathfrak{R}$) and quotient ($\mathfrak{Q}$)
%     axes of the Algebraic Lattice (Figure~\ref{fig:carrier-lattice}). ---

\begin{table}[htbp]
\centering
\caption{\textbf{Rank-axis rosetta ($\mathfrak{R}$).}  Per-rank
object, its structure-preserving construction, the algebraic concept
witnessing it, and the C++23 carrier.  The chain $T \to V_n(T) \to
M_n(T)$ exhibits the rank-axis steps of
Figure~\ref{fig:carrier-lattice}; $n = 2$ is the worked example,
higher ranks via combinators (\texttt{DirectSum},
\texttt{BlockUpperTriangular}, \texttt{Kronecker}; issue~\#368).}
\label{tab:rosetta-rank}
\small
\begin{tabular}{@{}l l l l@{}}
\toprule
\textbf{Object} & \textbf{Construction} & \textbf{Concept} & \textbf{C++23 carrier} \\
\midrule
$T$ (scalar) & identity & \cppinline{IsRing} / \cppinline{IsField} & \cppinline{T} \\
$V_n(T) = T^n$ & $\mathrm{Free}_n$ (\textbf{P} in HSP; Burris--Sankappanavar~\cite{burris1981universalalgebra} §II.10) & \cppinline{IsProductAlgebra<V>} & \cppinline{Vec2V<T>} \\
$M_n(T) = \mathrm{End}_T(V_n)$ & internal hom in $\mathbf{Mod}_R$ (\emph{not} HSP; Mac Lane~\cite{maclane1971categories} §VII) & \cppinline{is_endomorphism_ring_v<M, V>} & \cppinline{Matrix2x2V<T>} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\caption{\textbf{Quotient-axis rosetta ($\mathfrak{Q}$) — all rows
are \textbf{H} instances of HSP
(Burris--Sankappanavar~\cite{burris1981universalalgebra} §II.10).}
Per-quotient realisation: the polynomial ideal, the resulting
algebra, the structural property, and the C++23 carrier.  Each row
is a value of the parameter $X$ in Figure~\ref{fig:carrier-lattice};
compositions ($X = D \circ C$ etc.) commute when their generators
are independent.  Carrier-side concept across the table:
\cppinline{IsQuotientAlgebra<Q>}.}
\label{tab:rosetta-quotient}
\small
\begin{tabular}{@{}l l l l l@{}}
\toprule
\textbf{$X$} & \textbf{Polynomial ideal} & \textbf{Algebra} & \textbf{Structure} & \textbf{C++23 carrier} \\
\midrule
$D$ (Dual) & $(\varepsilon^2)$ & $R[\varepsilon]/(\varepsilon^2)$ & nilpotent (never integral) & \cppinline{Dual<F>} (\#504 generalises) \\
$C$ (Cplx) & $(i^2 + 1)$ & $R[i]/(i^2 + 1)$ & integral-domain extension when $i^2{+}1$ irreducible; field extension when $R$ is a field$^a$ & \cppinline{Complex<R>} (\#505 generalises) \\
$D \circ C \cong C \circ D$ & $(i^2 + 1, \varepsilon^2)$ & $R[i, \varepsilon]/(i^2{+}1, \varepsilon^2)$ & nilpotent dual of the complex extension & nested \cppinline{Dual<Complex<R>>} \\
$G$ (Galois) & $(q(x))$ for $q$ irreducible & $\mathbb{F}_p[x]/(q(x))$ & finite field $\mathbb{F}_{p^n}$ & \cppinline{𝔽64} ($\mathbb{F}_2[x]/(x^6{+}x{+}1)$, PR \#444) \\
\bottomrule
\end{tabular}

\vspace{0.5em}
{\footnotesize
\textsuperscript{a} The structural type of $C(R) = R[i]/(i^2 + 1)$
depends on both the irreducibility of $i^2 + 1$ over $R$ \emph{and}
on whether $R$ is itself a field.  When $i^2 + 1$ is irreducible over
$R$, the result is an integral-domain extension --- a field extension
when $R$ is already a field (e.g.\ $C(\mathbb{Q}) = \mathbb{Q}[i]$ is
a field; $C(\mathbb{R}) = \mathbb{C}$ is a field), an integral-domain
extension when $R$ is an integral domain but not a field (e.g.\
$C(\mathbb{Z}) = \mathbb{Z}[i]$, the Gaussian integers, is a Euclidean
domain but not a field).  When $i^2 + 1$ is reducible (e.g.\ over
$\mathbb{F}_2$ where it factors as $(x + 1)^2$), the quotient has
nilpotents and is isomorphic to $D(\mathbb{F}_2)$ (cf.\ \#505).
This is a concrete example of the "same construction, different
result depending on base" pattern the quotient axis encodes.  Per-cell compile-status (which $R \times X$
combinations the experimental library admits today vs which are pending
implementation) is tracked in the trait registry of \#499 as positive /
negative \texttt{static\_assert} witnesses.}
\end{table}


\subsubsection{Dedekind Cuts as the Canonical Example}
\label{sec:dedekind-cuts}

Dedekind's construction --- $r \in \mathbb{R}$ as the partition
$(L, U)$ of $\mathbb{Q}$ into rationals strictly below $r$ and those
at or above --- translates the intensional-first rule directly: the
two sets are type-level predicates, and \emph{the predicate is the
number}. The library's canonical worked example is Dedekind's own
$\sqrt{2}$: a rational $q$ is in the lower cut iff $q^{2} < 2$, a
decidable predicate over $\mathbb{Q}$ that compiles into a finite
type-level object --- no approximation, no series summation:

% Source: src/main/modules/dedekind/numbers/symbolic.cppm (Sqrt2_Symbolic, ll. 24-38)
\begin{listing}[htbp]
\caption{$\sqrt{2}$ as an intensional lower cut over $\mathbb{Q}$, from \texttt{dedekind.numbers:symbolic}. The predicate is the number.}
\label{lst:euler}
\begin{cpp}

template <typename Q>
constexpr auto Sqrt2_Symbolic() {
  const dedekind::sets::Ω<Q, TernaryLogic> universe{};
  // The lower cut L_sqrt2 = { q in Q | q^2 < 2 }.
  return ambient_set<Q>([universe](const Q& q) {
    if constexpr (std::floating_point<Q>) {
      if (std::isnan(q)) { return Ternary::Unknown; }
    }
    const auto in_cut =
        (q * q < static_cast<Q>(2)) ? Ternary::True : Ternary::False;
    return TernaryLogic::AND(universe(q), in_cut);
  });
}
\end{cpp}
\end{listing}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.3]
  % Number line
  \draw[->, thick] (-0.4, 0) -- (5.5, 0) node[right] {$\mathbb{Q}$};
  \foreach \x/\l in {0/0, 1/1, 2/2, 3/3, 4/4, 5/5} {
    \draw (\x, -0.08) -- (\x, 0.08);
    \node[below] at (\x, -0.12) {\small $\l$};
  }
  % The cut at sqrt(2) (approximately 1.414)
  \draw[thick, red, dashed] (1.414, -0.75) -- (1.414, 1.0);
  \node[red, above] at (1.414, 1.0) {$\sqrt{2}$};
  % Lower set L_sqrt2
  \draw[blue, line width=2.5pt, opacity=0.45] (-0.25, 0.4) -- (1.414, 0.4);
  \node[blue, above] at (0.6, 0.45) {\small $L_{\sqrt{2}} = \{\, q \in \mathbb{Q} \mid q^{2} < 2 \,\}$};
  % Upper set U_sqrt2
  \draw[orange, line width=2.5pt, opacity=0.55] (1.414, -0.4) -- (5.3, -0.4);
  \node[orange, below] at (3.5, -0.45) {\small $U_{\sqrt{2}} = \{\, q \in \mathbb{Q} \mid q^{2} \geq 2 \,\}$};
  % Sample rationals
  \foreach \x in {0, 0.5, 1, 1.25, 1.4} { \fill[blue] (\x, 0.4) circle (1.4pt); }
  \foreach \x in {1.5, 2, 3, 4, 5} { \fill[orange] (\x, -0.4) circle (1.4pt); }
\end{tikzpicture}
\caption{A Dedekind cut names $\sqrt{2}$ by partitioning $\mathbb{Q}$.
The lower half $L_{\sqrt{2}}$ (blue) and upper half $U_{\sqrt{2}}$
(orange) are both compile-time predicates. In \texttt{dedekind}, the
real number \emph{is} the partition: no approximation is stored, no
square root is computed, and the compiler still knows everything
about $\sqrt{2}$ that the predicate $q^{2} < 2$ expresses.}
\label{fig:dedekind-cut}
\end{figure}

No floating-point approximation is computed; no Newton iteration runs.
Asking ``is $\tfrac{7}{5}$ less than $\sqrt{2}$?'' reduces to
$\tfrac{49}{25} < 2$, decidable at translation time. Approximation to
a concrete float is available later, at a specific precision, when
the programmer asks for it. The intensional object survives
indefinitely without losing reasoning power.



\textbf{Showcase 2 — Lattice/square singleton in $\mathbb{C}$.}
The natural-number lattice (Gaussian integers, $0 \leq \mathrm{Re}, \mathrm{Im} \leq 3$)
intersected with the square $[0.5,1.5]^2$ contains exactly $c_3 = 1+i$.
The compiler proves all four membership assertions at translation time and folds the
exported function to a single \cppinline{ret i1 true}.

% Source: src/test/cpp/modules/dedekind/python/showcase_02_lattice_singleton.cpp
% (Abridged: the source ships four static_asserts on representative test
% points plus an extern "C" witness function; the listing below shows two
% representative static_asserts.  See the source file for the full set.)
\begin{listing}[htbp]
\caption{Showcase 2 source: lattice/square singleton in $\mathbb{C}$.}
\label{lst:showcase02_src}
\begin{cpp}

constexpr auto c = element<ℂ>;

constexpr auto natural_lattice_in_c =
    Set{c |[](const Complex<double>& z) {
      return is_integral_coordinate(z.real()) &&
             is_integral_coordinate(z.imag()) &&
             (z.real() >= 0.0) && (z.real() <= 3.0) &&
             (z.imag() >= 0.0) && (z.imag() <= 3.0);
    }};
constexpr auto square_c1_c2 = Set{c |[](const Complex<double>& z) {
  return (z.real() >= 0.5) && (z.real() <= 1.5) &&
         (z.imag() >= 0.5) && (z.imag() <= 1.5);
}};
constexpr auto lattice_square_intersection = natural_lattice_in_c & square_c1_c2;
using CLogic = typename decltype(lattice_square_intersection)::logic_species;

static_assert(lattice_square_intersection(Complex<double>{1.0, 1.0}) == CLogic::True);
static_assert(lattice_square_intersection(Complex<double>{0.0, 1.0}) == CLogic::False);
\end{cpp}
\end{listing}

% Source: src/test/cpp/modules/dedekind/python/showcase_02_lattice_singleton.ll
\begin{listing}[htbp]
\caption{Showcase 2 LLVM IR (\texttt{clang++-22 -O2}): constant true return.}
\label{lst:showcase02_ir}
\begin{cpp}

define noundef zeroext i1 @witness_lattice_square_singleton() local_unnamed_addr #0 {
  ret i1 true
}
\end{cpp}
\end{listing}


For the previously shown cardinality-1 reduction on $\mathbb{N}$, the same
meet operator, applied to two halfspaces whose bounds pin exactly one
integer, yields a result whose type \emph{is} \cppinline{Singleton<4>}, with
the unique inhabitant in the template parameter.


\paragraph{FIXME (\#362) --- showcase: carrier strength-reduction.}
The showcases above collapse the \emph{cardinality} of an intensional set
(transfinite $\mapsto$ finite $\mapsto$ singleton); a complementary showcase,
not yet listed in the showcase grid, would collapse the \emph{ambient set}
itself.  Mathematically: $A \subset \mathbb{N}$, $B \subset \mathbb{Z}$,
$\mathbb{N} \hookrightarrow \mathbb{Z}$ implies $A \cap B \subset \mathbb{N}$
(not $\subset \mathbb{Z}$), and dually $A \cup B \subset \mathbb{Z}$ widens
along the embedding.  The library now ships an existential proof of the
rule on the variant pair $(\texttt{Cardinality}, \texttt{SignedCardinality})$
--- $\texttt{Set}\langle\mathbb{N}\rangle \cap \texttt{Set}\langle\mathbb{Z}\rangle$
type-collapses to $\texttt{Set}\langle\mathbb{N}\rangle$, locked in by
unit tests --- but the mechanism does not yet appear as an IR-fixture
showcase with a \cppinline{ret i1 true / false} witness on a heterogeneous
intersection.  Adding such a showcase makes the carrier-axis collapse
empirically observable on the same footing as the cardinality-axis ones
above; it sits naturally next to Showcases~3 and~4 in the grid.  The full
carrier-promotion framework (any pair in the lattice $\mathbb{N} \subset
\mathbb{Z} \subset \mathbb{Q} \subset \mathbb{R} \subset \mathbb{C}$, not
just the variant pair) is tracked under issue \#362 and discussed in
\texttt{docs/design/carrier-lattice.md}.

Further showcases in the source tree exercise the same machinery on
$\mathbb{R}$ ambient carriers (showcase~5), non-point intervals with observable
cardinality (showcase~6: $-21 < n \leq 21$ on $\mathbb{Z}$, $|\cdot|=42$),
integer lattice intersected with real-valued bounds (showcase~7), and
2D rectangles via structural cartesian product (showcase~8: $42 \times 11 =
462$). All are IR-verified.

\paragraph{The machine layer below $\mathbb{N}$: \texttt{std::unsigned\_integral} is $\mathbb{Z}/2^w\mathbb{Z}$.}
The lattice $\mathbb{N} \subset \mathbb{Z} \subset \mathbb{Q} \subset \mathbb{R}
\subset \mathbb{C}$ rests on a machine layer occupied by the standard library's
primitive integer families. The honest textbook reading of
\cppinline{std::unsigned_integral} is that each width-$w$ instance is the
finite cyclic ring $\mathbb{Z}/2^w\mathbb{Z}$ under modular arithmetic --- a
\emph{commutative ring}, \emph{not} the textbook
$\mathbb{N}$.\footnote{The mod-wrap arithmetic gives every element an additive
inverse: $u + (2^w - u) \equiv 0 \pmod{2^w}$. The textbook $\mathbb{N}$ is a
commutative semiring \emph{without} additive inverse; calling \texttt{unsigned
int} ``natural numbers'' is a category-theoretic error in the direction of
claiming \emph{more} structure than $\mathbb{N}$ actually has.} The library's
$\mathbb{N}$ lives on the variant carrier \cppinline{Cardinality} (the
saturating commutative monoid, no $-$); the \cppinline{std::unsigned_integral}
layer is the canonical quotient of monoids
$\mathbb{N} \twoheadrightarrow \mathbb{Z}/2^w\mathbb{Z}$ that sits
\emph{above} $\mathbb{N}$ rather than coinciding with
it.\footnote{This framing is the correct reading of the project's
\emph{Honest Rejection} policy: a hardware species is rejected as a witness
for an idealised mathematical structure when its laws differ from those of the
idealised structure --- and the rejection is symmetric in direction, so that
``too much structure'' (\texttt{unsigned int}'s modular inverse) is rejected
on the same footing as ``too little'' (\texttt{int}'s undefined-behaviour
overflow that defeats closure).} The truthy-positivity injection
$\mathbb{B} \hookrightarrow \texttt{std::unsigned\_integral}$ and the
structure-forgetting lift
$\texttt{std::unsigned\_integral} \to \texttt{Cardinality}$ make the embedding
chain $\mathbb{B} \hookrightarrow \texttt{std::unsigned\_integral} \to
\mathbb{N}$ explicit at the type level.  The full discussion lives
in the report chapter on the carrier lattice.

\paragraph{The signed sibling: \texttt{std::signed\_integral} carries the operator surface but not the axiomatic ring.}
The signed-integer family (\texttt{int}, \texttt{long}, \texttt{long long})
presents the \emph{dual} category-theoretic error to its unsigned sibling.
Where \texttt{unsigned int} claims \emph{more} structure than $\mathbb{N}$
has, \texttt{int} claims \emph{less} structure than $\mathbb{Z}$
has.\footnote{The literal C++ ring operators ($+$, binary $-$, unary $-$,
$*$) all close on \texttt{int} and the order layer is fine; the operator
surface really is there.  But \emph{signed-overflow is undefined behaviour},
which means closure-under-arithmetic fails as a stable axiom.  The library's
posture is therefore honest in both directions: \cppinline{HasRingOperators<int>}
fires (the surface is real); \cppinline{IsRing<int, std::plus, std::multiplies>}
does not (the axiomatic-ring witness would be a false claim under UB).  The
exact-$\mathbb{Z}$ reading lives one carrier step deeper, on
\cppinline{SignedCardinality} (with $\pm\aleph_0$ saturation and
\cppinline{NaZ}).}  The Kleene-truth-value version of the embedding chain is
$\mathbb{K3} \hookrightarrow \texttt{std::signed\_integral} \to \mathbb{Z}$,
factored exactly as the Boolean / unsigned chain is: \texttt{embed\_}$\mathbb{K}3$\texttt{\_}$\mathbb{Z}$\texttt{\_}
encodes the three Kleene values into the signed-int machine layer (\texttt{False}
$\mapsto -1$, \texttt{Unknown} $\mapsto 0$, \texttt{True} $\mapsto 1$;
post-PR~\#439), and the lift $\texttt{std::signed\_integral} \to \mathbb{Z}$
(called \texttt{embed\_sint\_}$\mathbb{Z}$\texttt{\_}, post-issue~\#418 / renamed under~\#457) lands
in the variant $\mathbb{Z}$-proxy.\footnote{Both rejections are symmetric in
form: each carrier's claim is mismatched with the textbook structure it would
purport to inhabit, and the type system refuses the mismatch in both
directions.  An engineer who claims that the carrier their code uses is a ring
``in spirit'' is making the kind of public statement the truthfulness canons
of \cite{nspe_ethics} are about; the type system's refusal to accept the
\cppinline{IsRing<int>} witness is the mechanical analogue of an external
review catching the claim.}

% ============================================================
\section{Linear Algebra}
\label{sec:linalg}
% ============================================================

\subsection{Linear Programming: The Optimum as a Typed Constant}
\label{sec:lp-centrepiece}

Stare at a linear program for a moment. The polytope is an intersection
of halfspaces---a set, in exactly the comprehension sense of
Section~\ref{sec:sets-rules}. The objective is a linear functional. The
optimum is the one vertex of the polytope at which the objective is
largest.

The previous section's reductions collapsed a set to a simpler set. An
optimum is also a collapse: it reduces a set to a single point. The move
of this section is to lift that collapse into the type system. If every
halfspace and every objective coefficient is named at the type level,
the compiler has enough information to enumerate the vertices, run the
feasibility checks, compare the objective values, and emit the winner
as a literal. The LP solver is the C++ compiler.

\paragraph{Showcase~9 (centrepiece).} Consider a textbook LP:
\begin{align*}
  \text{maximise} \quad & 3x + 2y \\
  \text{subject to} \quad & x + y \leq 4 \\
                          & 2x + y \leq 6 \\
                          & x, y \geq 0
\end{align*}
The feasible region is a polygon in $\mathbb{Q}^2$
(Figure~\ref{fig:lp-polytope}); the optimum is the vertex
$(2, 2)$ where the non-axis-aligned constraints $x + y = 4$ and
$2x + y = 6$ are simultaneously binding (objective value $10$). In
\texttt{dedekind.optimization:lp}, constraints are NTTPs and the
reduction is a variadic template whose output carries the vertex at the
type level:

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.85]
  % Axes
  \draw[->, thick] (-0.3, 0) -- (5.3, 0) node[right] {$x$};
  \draw[->, thick] (0, -0.3) -- (0, 5.5) node[above] {$y$};
  % Gridlines
  \draw[gray, very thin, opacity=0.5] (0, 0) grid (5, 5);
  \foreach \i in {1,2,3,4,5} {
    \node[below left, font=\scriptsize] at (\i, 0) {$\i$};
    \node[below left, font=\scriptsize] at (0, \i) {$\i$};
  }
  % Feasible region: polygon (0,0)-(3,0)-(2,2)-(0,4)
  \fill[blue!18] (0,0) -- (3,0) -- (2,2) -- (0,4) -- cycle;
  \draw[blue!40!black, thick] (0,0) -- (3,0) -- (2,2) -- (0,4) -- cycle;
  % The two non-trivial constraint boundaries (extended past feasible region)
  \draw[green!50!black, dashed, thick] (-0.1, 4.1) -- (4.5,-0.4)
    node[pos=0.92, below right, font=\small] {$x+y=4$};
  \draw[red!60!black, dashed, thick] (-0.1, 6.2) -- (3.4, -0.8)
    node[pos=0.92, below right, font=\small] {$2x+y=6$};
  % Objective level line through the optimum: 3x+2y=10 passes through (2,2) and (0,5)
  \draw[violet!70!black, dotted, thick] (-0.1, 5.15) -- (3.5, -0.25)
    node[pos=0.08, above right, font=\small] {$3x+2y=10$};
  % Feasible vertices
  \fill[black] (0,0) circle (2.5pt);
  \fill[black] (3,0) circle (2.5pt);
  \fill[black] (0,4) circle (2.5pt);
  % Highlight the optimum
  \fill[red] (2,2) circle (3.5pt);
  \draw[red, thick] (2,2) circle (6pt);
  % Vertex labels with objective values
  \node[below left, font=\small] at (0,0) {$(0,0)\!:\!0$};
  \node[below right, font=\small] at (3,0) {$(3,0)\!:\!9$};
  \node[above left, font=\small] at (0,4) {$(0,4)\!:\!8$};
  \node[above right, font=\small, red] at (2.1,2.05) {$\mathbf{(2,2)\!:\!10}$};
  % Objective direction arrow
  \draw[->, thick, violet!80!black] (4.0, 3.2) -- (4.6, 3.6);
  \node[violet!80!black, font=\small] at (4.3, 2.95) {$c = (3,2)$};
\end{tikzpicture}
\caption{The feasible polygon of Showcase~9. Vertices are labelled
with the objective value $3x + 2y$ at each. The active-set
enumeration considers all $\binom{4}{2} = 6$ pairs of constraints;
four pairs produce feasible intersections, and the optimum is the one
vertex (red) where both non-axis-aligned constraints
$H_1\colon x+y\leq 4$ and $H_2\colon 2x+y\leq 6$ bind simultaneously.
At translation time the compiler performs this enumeration; the
emitted binary contains only the literal coordinates of $(2,2)$.}
\label{fig:lp-polytope}
\end{figure}

% Source: src/test/cpp/modules/dedekind/python/showcase_09_lp_vertex_typed_constant.cpp
\begin{listing}[htbp]
\caption{Compile-time LP: the optimum IS a type. From \texttt{showcase\_09\_lp\_vertex\_typed\_constant.cpp}.}
\label{lst:lp-centrepiece}
\begin{cpp}

using Rat = Rational<SignedExtensionalCardinal<>>;  // canonical exact ℚ
using H1 = Halfspace2D<Rat, Rat{1}, Rat{1}, Rat{4}>;     //  x +  y <= 4
using H2 = Halfspace2D<Rat, Rat{2}, Rat{1}, Rat{6}>;     // 2x +  y <= 6
using H3 = Halfspace2D<Rat, Rat{-1}, Rat{0}, Rat{0}>;    //  x      >= 0
using H4 = Halfspace2D<Rat, Rat{0}, Rat{-1}, Rat{0}>;    //       y >= 0

using Optimum = decltype(maximize<Rat, Rat{3}, Rat{2}, H1, H2, H3, H4>());
static_assert(std::same_as<Optimum, Vec2<Rat, Rat{2}, Rat{2}>>);
\end{cpp}
\end{listing}

The reduction enumerates vertex candidates at compile time (all
$\binom{4}{2} = 6$ pairs of binding constraints), solves each $2 \times 2$
active-set system via \texttt{Invertible2x2}'s closed-form Cramer
inverse (exact over $\mathbb{Q}$), filters by feasibility against the
non-active constraints, and picks the objective-maximising feasible
vertex. Every step is a template specialisation; no runtime iteration
survives.

\paragraph{IR witness.} The paper's claim is not prose; it is
translation-time mechanics. Two \texttt{extern "C"} functions in the
fixture return the numerator of the optimum's $x$- and $y$-coordinates.
At \texttt{-O2}, \texttt{clang}'s optimiser folds the entire LP
reduction to literal integers:

% Source: src/test/cpp/modules/dedekind/python/showcase_09_lp_vertex_typed_constant.ll
\begin{listing}[htbp]
\caption{Emitted LLVM IR for Showcase~9. The six candidate solves, six feasibility checks, and six objective comparisons collapse to constant returns.}
\label{lst:lp-ir}
\begin{cpp}

define noundef i64 @witness_lp_optimum_x() ... {
  ret i64 2
}

define noundef i64 @witness_lp_optimum_y() ... {
  ret i64 2
}
\end{cpp}
\end{listing}

The IR is checked into the repository as a fixture
(\texttt{showcase\_09\_lp\_vertex\_typed\_constant.ll}); a build-time
gate (\texttt{make ir-fixture-check}) fails CI if optimiser drift ever
loses the collapse. The binary contains no LP solver; no simplex
tableau; no active-set machinery. More precisely: the vertex's
coordinates are non-type template parameters of the result type
(\cppinline{Vec2<Rat, Rat{2}, Rat{2}>}), so the C++23 NTTP machinery
is what carries the answer through to type-level. The input is a
constexpr value passed as NTTP; the output is a type whose NTTPs
encode the coordinates; the runtime call returning a coordinate
collapses to a literal load in the emitted IR.  Because the reduction
lands in \texttt{Rational<...>} rather than IEEE floating point, the
\cppinline{ret i64 2} output is bit-identical across ARM and x86
back-ends and host floating-point ABIs --- the typed constant has
no rounding mode to disagree about.  This is the embedded /
cross-platform-determinism guarantee the paper has been claiming
throughout, made concrete by the absence of any \texttt{double}-typed
reduction in the IR.

\subsection{Forward-Mode AD as a Sibling Realisation of Compile-Time Derivatives}
\label{sec:ad-elliott}

Swap the ring: re-run the same LP reduction over
$\mathbb{Q}[\varepsilon]/(\varepsilon^2)$ instead of $\mathbb{Q}$.
The optimum and its derivative with respect to a chosen input come
out as one typed constant; no separate derivative pass. We follow
Elliott's framing~\cite{elliott2018simple} that forward-mode AD is
the chain-rule-carrying product $(f, f')$, with the dual-number ring
making the chain rule fall out of the ring laws.\footnote{The same
abstract derivative is computed numerically by
\texttt{dedekind.analysis:ftc::derivative\_at} (central difference)
and symbolically by
\texttt{dedekind.algebra:polynomial::RigPolynomial::derive}. The
\texttt{Dual<F>} carrier slots into any NTTP-parameterised carrier
---including the LP's \texttt{Halfspace2D<T, a, b, c>}---without
special-casing, because its primal and tangent are structural
fields.}

\paragraph{Showcase: parametric LP sensitivity.} Perturb $H_1$'s
bound by $\varepsilon$ (replace $x + y \leq 4$ with
$x + y \leq 4 + \varepsilon$). The active set is unchanged for
infinitesimal perturbation; the Cramer solve now runs over duals:

% Source: src/test/cpp/modules/dedekind/optimization/lp_test.cpp
\begin{listing}[htbp]
\caption{Parametric LP over \texttt{Dual<Rat>}. From \texttt{lp\_test.cpp}.}
\label{lst:lp-dual}
\begin{cpp}

using D = Dual<Rat>;
using H1P = Halfspace2D<D, D{Rat{1L}}, D{Rat{1L}},
                        D{Rat{4L}, Rat{1L}}>;  // bound = 4 + epsilon
using H2D = Halfspace2D<D, D{Rat{2L}}, D{Rat{1L}}, D{Rat{6L}}>;
using H3D = Halfspace2D<D, D{Rat{-1L}}, D{Rat{0L}}, D{Rat{0L}}>;
using H4D = Halfspace2D<D, D{Rat{0L}}, D{Rat{-1L}}, D{Rat{0L}}>;

constexpr auto v = maximize_value<D, D{Rat{3L}}, D{Rat{2L}},
                                  H1P, H2D, H3D, H4D>();
static_assert(v.x.val == Rat{2L}  && v.x.der == Rat{-1L});
static_assert(v.y.val == Rat{2L}  && v.y.der == Rat{2L});
\end{cpp}
\end{listing}

\noindent
The primal $(2, 2)$ is the original optimum; the tangents $(-1, +2)$
match the hand derivation $x^*(t) = 2 - t$, $y^*(t) = 2 + 2t$ on the
active set $\{x + y = 4 + t,\; 2x + y = 6\}$. The LP reduction was
written once over an unconstrained ring-like carrier; instantiating
it over the dual-number ring discharges sensitivity analysis as a
side effect of the ring laws. The same
$\mathtt{Dual}\circ\mathtt{Complex}\circ\mathtt{Rational}$ layering
extends to holomorphic automatic differentiation at compile time;
we leave its application to a follow-up.

\subsection{Linear Algebra: \texorpdfstring{$\mathbb{C}$ and $\mathbb{D}$}{ℂ and 𝔻} as Subrings of \texorpdfstring{$M_2(\mathbb{Q})$}{M₂(ℚ)}}
\label{sec:lin-alg-embedding}

The classical regular representations
$a + bi \mapsto
\bigl(\begin{smallmatrix} a & -b \\ b & a \end{smallmatrix}\bigr)$
and
$a + b\varepsilon \mapsto
\bigl(\begin{smallmatrix} a & b \\ 0 & a \end{smallmatrix}\bigr)$
are ring monomorphisms $\mathbb{C}, \mathbb{D} \hookrightarrow
M_2(\mathbb{Q})$ (see, e.g., Lang~\cite{lang2002algebra}). The
library claims both identities at the type level over
\cppinline{Rational<SignedExtensionalCardinal<>>} and the compiler
discharges them as bit-perfect \texttt{static\_assert}s, with the
defining relation $\varepsilon^2 = 0$ lifting to
$\bigl(\begin{smallmatrix} 0 & 1 \\ 0 & 0 \end{smallmatrix}\bigr)^2
= 0$ on the matrix side.

In the HSP vocabulary of §\ref{sec:axiomatic}, both embeddings are
components of a single natural-transformation schema
$\operatorname{Cplx}, \operatorname{Dual} \Rightarrow
\operatorname{End}_R \circ \operatorname{Free}_2$ — composing an H
operation (quotient by $(i^2+1)$ or $(\varepsilon^2)$) with a P
operation ($\operatorname{Free}_2$) and the internal-hom
$\operatorname{End}_R$ — with target $M_2(R) =
\operatorname{End}_R(R^2)$.  The worked instance ships at $R =
\mathbb{Q}$; the listing below discharges
$\operatorname{Cplx}(\mathbb{Q}), \operatorname{Dual}(\mathbb{Q})
\hookrightarrow M_2(\mathbb{Q})$ mechanically.

% Source: src/test/cpp/modules/dedekind/linear_algebra/embeddings_test.cpp
\begin{listing}[htbp]
\caption{Matrix embeddings of complex and dual numbers over $\mathbb{Q}$.}
\label{lst:lin-alg-embeddings}
\begin{cpp}

using Rat = Rational<SignedExtensionalCardinal<>>;
constexpr Complex<Rat> z{Rat{1}, Rat{2}}, w{Rat{3}, Rat{-5}};
constexpr Dual<Rat>    d{Rat{2}, Rat{3}}, e{Rat{-1}, Rat{5}};
static_assert(as_matrix2x2(z + w) == as_matrix2x2(z) + as_matrix2x2(w));
static_assert(as_matrix2x2(z * w) == as_matrix2x2(z) * as_matrix2x2(w));
static_assert(as_matrix2x2(d + e) == as_matrix2x2(d) + as_matrix2x2(e));
static_assert(as_matrix2x2(d * e) == as_matrix2x2(d) * as_matrix2x2(e));
constexpr Dual<Rat> eps{Rat{0}, Rat{1}};
static_assert(as_matrix2x2(eps) * as_matrix2x2(eps) == zero_matrix2x2_v<Rat>);
\end{cpp}
\end{listing}

\subsection{From Cardinality Pruning to Type-Directed Collapse}
\label{sec:gap}

Step back from the showcases for a moment. What is the pattern? In every
one of them we started with an intensional set---a predicate---and we
ended with a named extensional object whose \emph{type} carries the
answer: $\varnothing$, a singleton, an order interval of known
cardinality, an LP vertex. The collapse is not incidental; it is a
rewrite rule, running inside the type system, at translation time. To
make the claim precise, we first formalise ``how much the compiler
knows about a set'' as a lattice, then state the theorem as monotone
motion up that lattice.

\begin{definition}[Computability tier]
\label{def:computability-tier}
For a compile-time set $S$ over an ambient carrier $T$, the
\emph{computability tier} is the triple
\[
  \mathcal{T}(S) \;=\; \bigl(\,\mathsf{R}(S),\; \mathsf{L}(S),\; \mathsf{K}(S)\,\bigr)
  \;\in\; \{\mathrm{Int},\mathrm{Ext}\} \times \{\mathrm{K3},\mathrm{Cl}\} \times \{\mathrm{Opq},\mathrm{Elt}\}
\]
whose components are the \emph{representation} tier ($\mathsf{R}$:
intensional vs.\ extensional, i.e.\ predicate vs.\ elements-in-type),
the \emph{logic} tier ($\mathsf{L}$: Kleene three-valued vs.\
classical two-valued membership), and the \emph{knowledge} tier
($\mathsf{K}$: elements opaque vs.\ carried in the type). Each axis
carries the order $\mathrm{Int} \preceq \mathrm{Ext}$,
$\mathrm{K3} \preceq \mathrm{Cl}$,
$\mathrm{Opq} \preceq \mathrm{Elt}$; the product order is
componentwise. The resulting Boolean lattice $(\mathcal{T}, \preceq)$
has bottom $\bot = (\mathrm{Int},\mathrm{K3},\mathrm{Opq})$ (``a raw
predicate over a floating-point carrier with no inhabitants known'')
and top $\top = (\mathrm{Ext},\mathrm{Cl},\mathrm{Elt})$ (``a named
finite set whose members are template parameters and whose membership
is decidable'').
\end{definition}

Every showcase of Section~\ref{sec:pruning} is a motion inside this
lattice. The empty-halfspace reductions of Showcase~3 go $(\mathrm{Int},\mathrm{K3},\mathrm{Opq}) \mapsto \top$
in one step (the result is $\varnothing$); the cardinality-1 reduction
of Showcase~4 and the LP reduction of Showcase~9 do the same. The
reductions never slide downward. With that in hand:

\begin{theorem}[Type-Directed Collapse, totally-ordered 1D case]
\label{thm:type-directed-collapse-1d}
Let $T$ be a totally ordered carrier, and let
\[
  P_i(x) \;=\; (x \ge a_i), \quad i = 1, \dots, m,
  \qquad
  Q_j(x) \;=\; (x \le b_j), \quad j = 1, \dots, n,
\]
be lower-bound and upper-bound halfspace predicates over $T$ whose
pivots $a_i, b_j \in T$ are carried at the type level as non-type
template parameters. Define the effective bounds
$a^\star \coloneqq \max_i a_i$ if $m \ge 1$ and $a^\star \coloneqq
\bot$ if $m = 0$, and $b^\star \coloneqq \min_j b_j$ if $n \ge 1$
and $b^\star \coloneqq \top$ if $n = 0$ (where $\bot, \top$ denote
the carrier's bottom and top elements, taken as
$\pm\infty$-style sentinels when $T$ is unbounded). Then the meet
\[
  S \;=\; \bigcap_{i=1}^{m} \{x \in T \mid P_i(x)\}
        \;\cap\;
        \bigcap_{j=1}^{n} \{x \in T \mid Q_j(x)\}
\]
is resolved at compile time to exactly one of:
\begin{enumerate}
  \item $\varnothing$, when $a^\star > b^\star$;
  \item a \cppinline{Singleton<v>}, when $T$ is integral and
        $a^\star = b^\star = v$;
  \item an \cppinline{OrderInterval}~$[a^\star, b^\star]$
        otherwise.
\end{enumerate}
\end{theorem}

\begin{proof}
By induction on $m + n$.

\smallskip
\noindent\emph{Base case} ($m + n = 0$): no constraints; $S = T$,
which is the order interval $[\bot, \top]$ (case 3) by the boundary
convention.

\smallskip
\noindent\emph{Inductive step.} Assume the claim holds for $k$ total
constraints. Add one more; without loss of generality (by the
symmetry between lower and upper bounds) take it to be a fresh
lower-bound predicate $P_*(x) = (x \ge a_*)$. The inductive
hypothesis gives a reduced form $S_k$ of one of the three shapes.
Form $S = S_k \cap \{x \mid x \ge a_*\}$. Two compile-time
comparisons of $a_*$ against the bounds of $S_k$ (where present)
suffice:

\begin{itemize}
  \item \textbf{Redundant.} If $S_k$ has lower bound $a$ and
        $a_* \le a$, then $\{x \mid x \ge a_*\} \supseteq S_k$ and
        $S = S_k$; the form is preserved.
  \item \textbf{Tightening.} If $S_k$ is the interval $[a, b]$ (or
        unbounded above) and $a < a_* \le b$ (or no upper bound),
        then $S = [a_*, b]$ in case 3; if $T$ is integral and
        $a_* = b$, then $S = \{a_*\}$ (case 2).
  \item \textbf{Contradicting.} If $S_k$ has upper bound $b$ and
        $a_* > b$, or if $S_k$ is already $\varnothing$, then
        $S = \varnothing$ (case 1).
\end{itemize}

\noindent
The same case-split, with rôles of $a_i$ and $b_j$ exchanged,
discharges the symmetric upper-bound addition. Each case lands in
one of the three forms, completing the induction.
\end{proof}

\paragraph{Theorems for free.}
The structural shape of Theorem~\ref{thm:type-directed-collapse-1d}
is Wadler's \emph{Theorems for Free!}~\cite{wadler1989theorems}
move at the algebraic register. Wadler showed that polymorphic
types in the typed lambda calculus discharge a class of theorems by
parametricity alone, with no further proof obligation on the
caller. Here, the analogous discharge runs from \emph{carrier-axiom
declaration} (\cppinline{HasTotalOrderOperators<T>} +
\cppinline{IsTotallyOrdered<T>}, plus the integral-carrier
witness when case 2 is to fire) to a carrier-level reduction of any
finite intersection of half-line constraints. The engineer's
honesty obligation is the axiom set; the collapse is
type-level free. The proof above is one demonstration that the
discharge is mechanical---the inductive argument does not depend on
which carrier or which pivot type, only on the axiom slots being
filled.

\paragraph{Existential anchor.}
The theorem is realised in the library by witnesses that
materialise at the static-assertion layer; the canonical exhibit on
$\mathbb{N}$ is the typed-constant collapse
\[
  \{x \in \mathbb{N} \mid x \ge 2\} \,\cap\, \{x \in \mathbb{N} \mid x \le 5\}
  \;\longrightarrow\;
  \texttt{OrderInterval}\langle 2, 5\rangle,
\]
checked at compile time. Section~\ref{sec:pruning} catalogues these
witnesses; the LLVM~IR fixtures pin the resulting machine-code
collapse downstream of the type-level reduction.

\begin{conjecture}[Type-Directed Collapse, n-dimensional generalisation]
\label{conj:type-directed-collapse-nd}
Let $T_1, \dots, T_d$ be totally ordered carriers and let
$\{P_k\}_{k=1}^{N}$ be halfspace predicates over the product
$T_1 \times \dots \times T_d$ whose pivots are carried at the type
level. The meet
\[
  S \;=\; \bigcap_{k=1}^{N} \{(x_1, \dots, x_d) \mid P_k(x_1, \dots, x_d)\}
\]
admits a compile-time reduction to a typed normal form
generalising the three cases of
Theorem~\ref{thm:type-directed-collapse-1d} (empty, vertex-singleton,
or a polytope record carrying its active constraint set), under a
suitable axiom set on the product carrier (total order per axis,
closure of the meet under the product, and a vertex-enumeration
discipline for the polytope case). Writing $S_k$ for the
single-constraint set, the reduction is monotone up the
computability-tier lattice:
\[
  \mathcal{T}(S) \;\succeq\; \bigvee_{k} \mathcal{T}(S_k),
\]
and strictly so on at least one axis in every case the library
exhibits.

\smallskip
\noindent\emph{Status of claim.} This conjecture is \emph{not}
proved in the paper. It is supported by the same exhibition pattern
as the 1D theorem: each case is checked into the library as a
\cppinline{static_assert}/\cppinline{STATIC_CHECK} witness and,
where relevant, as an LLVM~IR fixture that fails CI on optimiser
drift (see Section~\ref{sec:pruning}). The 2D linear-programming
showcase of Section~\ref{sec:lp-centrepiece} discharges one
non-trivial instance via active-set vertex enumeration; a general
inductive proof over pivot configurations and dimensions $d$ is left
open.
\end{conjecture}

\begin{corollary}[Iterated collapse]
\label{cor:iterated-collapse}
Any finite composition of type-directed-collapse reductions is itself
monotone up $\mathcal{T}$. In particular, no sequence of library
operations can ever return a set that is \emph{less} decidable than
its inputs.
\end{corollary}

So: not ``a faster way to do set algebra'', but a rewrite rule that
happens to live in the C++ type system, and whose action is
monotone-ascending in a lattice of decidability. The halfspace case is
the smallest non-trivial instance we know how to prove mechanically.
Four further axes appear to obey the same pattern; the project tracker
carries one existential proof per axis---one worked instance, one IR
fixture, one compile-time witness---rather than a general framework.
We would rather ship one reduction that fully collapses than four
sketches that do not.

% Tracker cross-reference for maintainers (invisible in PDF):
%   Carrier tightening      -> issue #362
%   Smooth optimization     -> issue #363
%   Linear programming      -> issue #364  (demonstrated here)
%   Lattice-law rewriting   -> issue #365
\begin{itemize}
  \item \textbf{Carrier tightening.} Intersecting a subset of
    $\mathbb{N}$ with a subset of $\mathbb{R}$ should land in $\mathbb{N}$:
    the carrier lattice contributes its own structural reduction via the
    existing canonical embeddings (\cppinline{embed_uint_sint_},
    \cppinline{embed_ℚ_ℝ}, \ldots).
  \item \textbf{Smooth optimization.} The stationary set
    $\{x \mid \nabla f(x) = 0\}$ for a quadratic $f$ collapses to a
    \cppinline{Singleton<x*>} with the minimizer in the template parameter.
    Same reduction pattern, applied at the extremum operator.
    \emph{Scope note:} the polytope / linear-programming domain is not
    incidental.  Active-set enumeration over halfspaces is
    constructively bounded at compile time precisely because halfspaces
    have finite combinatorial structure; non-polyhedral domains
    (KKT-stationary-point search over arbitrary smooth $f$) require
    symbolic root-finding which does not terminate at template-
    instantiation time.  Quadratic $f$ is the worked-instance
    boundary; non-linear extensions are post-paper work tracked
    under \#363.
  \item \textbf{Linear programming} (demonstrated in
    Section~\ref{sec:lp-centrepiece} and Section~\ref{sec:ad-elliott}).
    \cppinline{maximize(c, polytope)} reduces, for compile-time-sized LPs, to
    a typed vertex via type-level active-set enumeration, and the same
    reduction over a dual-number carrier yields sensitivity analysis
    with no extra code path: the solver is the compiler.
  \item \textbf{Lattice-law rewriting.} Distributivity and
    absorption applied at the type level rewrite
    $(A \cup B) \cap \lnot A$ to $B \cap \lnot A$ \emph{before} local
    reduction rules fire, exposing collapses that a purely local analysis
    would miss.
\end{itemize}

Each axis sits on the foundation shared with Section~\ref{sec:pruning}: NTTP
predicates, structured \cppinline{operator&} dispatch, and the computability
tiers visible in the showcases. Taken together they sketch a program of
which the present halfspace reductions are one installment.

The polynomial-calculus exhibit (companion-repository
\texttt{:polynomial} / \texttt{:ftc} partitions) belongs to the same
family but at the \emph{function} layer: a polynomial reduced under
\cppinline{derive} (a ring homomorphism) or a numerical integrator
yields another intensional object whose identity with a target is
discharged by the compiler.

\paragraph{Scaling beyond atomic carriers via structure-preserving combinators.}
The same partition---atoms below, runtime above---governs the
linear-algebra surface the paper exhibits at the $2 \times 2$ scale.
The library ships \texttt{Invertible2x2}, \texttt{Matrix2x2V},
\texttt{Vec2V} as statically-validated atoms; larger systems are
\emph{not} reached by inflating the atom (we do not aim for
$n \times n$ inversion in templates), but by composing atoms through
\emph{structure-preserving combinators} that propagate the
invariants up.  \texttt{DirectSum<A, B>} carries invertibility from
its factors to the block-diagonal whole and is shipped today;
$4 \times 4$ invertibility is therefore already a one-combinator
step from the $2 \times 2$ atom, with $(A \oplus B)^{-1} =
A^{-1} \oplus B^{-1}$ exact over $\mathbb{Q}$ and verifiable by
\texttt{static\_assert}:

% Source:
% src/test/cpp/modules/dedekind/linear_algebra/matrix_test.cpp
\begin{cpp}
TEST_CASE("linear_algebra:matrix --- Tier 1: "
          "DirectSum preserves invertibility") {
  constexpr Invertible2x2<Rat, Rat{1L}, Rat{2L},
                          Rat{3L}, Rat{4L}> block_M{};
  // 90° rotation over ℚ.
  constexpr Invertible2x2<Rat, Rat{0L}, Rat{-1L},
                          Rat{1L}, Rat{0L}> block_rot{};
  constexpr DirectSum<decltype(block_M),
                      decltype(block_rot)> ds{};
  constexpr auto ds_inv = ds.inverse();

  using IdentityDirectSum =
      DirectSum<Identity2x2<Rat>, Identity2x2<Rat>>;
  STATIC_CHECK(ds * ds_inv == IdentityDirectSum{});
  STATIC_CHECK(ds_inv * ds == IdentityDirectSum{});
}
\end{cpp}

\noindent \texttt{BlockUpperTriangular<A, B, D>} extends the same
propagation to triangular blocks under a degenerate-Schur condition
(also shipped, also testable as a single \texttt{static\_assert}); a
future Kronecker combinator preserves orthogonality and lattice
structure compositionally.  Each combinator is itself a small
structurally-checked unit; their composition gives $2^k \times 2^k$
matrices whose invariants follow by induction without exponential
blowup in the templates.  The remaining roadmap items
% Backlog: issues #366 (block-Schur invertibility recursion, full
% non-degenerate case), #368 (catalog of well-behaved matrix
% carriers --- orthogonal, rotation, diagonal), #369 (structure-
% preserving combinators including Kronecker).
fall out of the same Axiomatic-Systems-Programming discipline rather
than requiring a different paradigm, and are tracked in the project
backlog rather than discharged here.  The $2 \times 2 \to 4 \times 4$
step is a witness, not a promise; everything beyond is the natural
recursion of the same combinator pattern.

\paragraph{Carrier choices, error messages, ergonomics.}
The carrier-vs-overflow trade-offs are catalogued in
Section~\ref{sec:carrier-rosetta}'s Rosetta tables: showcases default
to \cppinline{Rational<SignedExtensionalCardinal<>>} (exact, no
silent overflow); programmers who trade exactness for native
performance pick a primitive row and accept the named failure mode
(signed-overflow UB on \cppinline{int}; non-associative rounding on
\cppinline{float}/\cppinline{double}).  The combinators of
Section~\ref{sec:pruning} bind to \cppinline{HasRingOperators}
rather than any particular carrier, so the swap is a one-line change
at the \cppinline{using} declaration.  Error messages, by contrast,
are a more universal C++ pain: a failed \texttt{static\_assert} in a
deep reduction can produce a multi-page expansion trail that no
library can fully repair.  The realistic ambition is to \emph{fail
fast at the named axiom} --- the first message the reader sees should
be the predicate that genuinely broke, not the cascade of downstream
failures it triggered.  The library structures concept ladders so
that each rung adds a single named requirement, and uses targeted
\cppinline{static_assert} messages inside internal reductions, so a
violation surfaces at the smallest witness rather than deep inside
overload resolution.\footnote{Worked diagnostic: a rank-deficient
\cppinline{Invertible2x2<Rat, 1, 2, 2, 4>} input produces a single
named message (\texttt{Invertible2x2 requires full rank: a*d - b*c
!= 0}) at the \texttt{static\_assert} site with one \texttt{note}
pointing at the instantiation, no substitution-failure cascade.  The
predicate sits inside the class body as a literal
\texttt{static\_assert} rather than as a \cppinline{requires}-clause
to be probed, so the diagnostic fires at the named axiom directly.}

In short: the compiler wall has a predictable shape---depth, steps,
combinatorics, numeric range, diagnostics---and each is either
pushable by a flag, avoidable by staying within the instance regime
for which this technique is intended, or a known limitation we do not
claim to have solved. The paper's contribution lies entirely to the
left of that wall.

\paragraph{Frictions: where C++23 and IEEE numerics break the chain.}
The axioms cannot always be discharged. The friction points encountered while
building the library are observational, not advocacy, and each is
pinned at a concrete file in source (with the relevant lines flagged
in the corresponding partition's witness blocks) so a reader can
verify the mechanism rather than take the claim on faith.
\emph{Integer-promotion shape mismatch}: \cppinline{bool + bool} and
\cppinline{bool * bool} return \cppinline{int} rather than
\cppinline{bool}; \cppinline{unsigned short + unsigned short} /
\cppinline{unsigned short * unsigned short} and
\cppinline{unsigned char + unsigned char} /
\cppinline{unsigned char * unsigned char} likewise promote to
\cppinline{int}. The functor surface
(\cppinline{std::plus<bool>}, \cppinline{std::multiplies<unsigned short>})
returns the carrier by signature, but the literal \cppinline{+} /
\cppinline{*} operator surface does not (witnessed at
\texttt{algebra/ring.cppm}, \texttt{algebra/field.cppm},
\texttt{numbers/uint.cppm}, \texttt{order/lattice.cppm}).
\emph{Signed-overflow UB defeats $\mathbb{Z}$}: \cppinline{int} /
\cppinline{long} / \cppinline{long long} carry the literal ring
operator surface
(\cppinline{HasRingOperators<int>} fires) but \emph{not} the
axiomatic ring witness, because closure-under-arithmetic fails as a
stable axiom under signed-overflow UB; the negative pins
\cppinline{!Group_ℤ<int>}, \cppinline{!IsArithmeticAdditiveGroup<int>},
\cppinline{!category::IsRing<int, std::plus, std::multiplies>} at
\texttt{numbers/sint.cppm} make the rejection mechanical.
\emph{IEEE 754 breaks reflexivity}: \cppinline{NaN <= NaN}
evaluates to \cppinline{false}, so
\cppinline{is_reflexive_v<double, std::less_equal<>>} is deliberately
not registered, and \cppinline{!IsTotallyOrdered<double>},
\cppinline{!IsTotallyOrdered<long double>}, and
\cppinline{!category::IsRing<double, std::plus, std::multiplies>}
all hold (\texttt{numbers/floating\_point.cppm} +
\texttt{order/total.cppm}); the operational ordered-field reading
survives via \cppinline{IsOrderedField<double>} in
\texttt{morphologies/archimedean.cppm}.
\emph{\cppinline{std::variant} NTTP-unfriendliness}: the variant
$\mathbb{Z}$-proxy \cppinline{SignedCardinality} (variant of
\cppinline{SignedExtensionalCardinal<>}, $\pm\aleph_0$,
\cppinline{NaZ}) cannot be a non-type template parameter without
workarounds, blocking compile-time arithmetic on the saturating
integer carrier (\texttt{sets/cardinality.cppm},
\texttt{numbers/rational.cppm}).
\emph{\cppinline{std::integral} unifies three structurally distinct algebras}:
\cppinline{std::integral T = std::unsigned_integral T || std::signed_integral T || std::same_as<T, bool>},
and the three siblings are a finite cyclic ring, an
operator-surface-without-axioms carrier, and (separately) the
Boolean rig + Galois field $\mathbb{F}_2$; nothing axiomatic
survives the union (\texttt{numbers/integral.cppm}). Each friction
is the structural reason an axiom slot is left
\emph{unsatisfied} for the relevant family.

\paragraph{The policy gate as the bridge.}
Where the strict axioms cannot be discharged but the operator surface
must still be programmed against, the library exposes an operational
policy gate. The canonical example is the rational-field claim,
which the showcases program against without committing the carrier
to a specific machine integer:
%
\begin{cpp}
// numbers/rational.cppm --- operational field-like witness for ℚ.
// HasFieldOperators is the operator-surface gate; the strict
// axiomatic field witness lives at category::IsField but is not
// load-bearing for downstream code that only needs +, -, *, /.
static_assert(
    dedekind::algebra::HasFieldOperators<Rational<default_integer>>,
    "Rational<I> must satisfy the operational field-like witness "
    "(ℚ is a field).");

static_assert(
    dedekind::algebra::HasFieldOperators<
        Rational<SignedExtensionalCardinal<>>>,
    "Rational<SignedExtensionalCardinal<>> must satisfy "
    "HasFieldOperators: the intended arbitrary-precision "
    "signed-rational carrier for ℚ.");
\end{cpp}
%
The two carriers differ in their machine-level friction profile
(\cppinline{default_integer} trips signed-overflow UB on extreme
inputs; \cppinline{SignedExtensionalCardinal<>} does not, at the
cost of variant-NTTP friction) but both discharge the operational
gate. Downstream combinators bind to \cppinline{HasFieldOperators}
rather than to \cppinline{category::IsField}, and the carrier swap
between the two is a one-line change at the \cppinline{using}
declaration. The gate is the load-bearing concept; the strict
axiomatic field is reachable, on demand, when a use site warrants
the heavier obligation.

In short, where Section~\ref{sec:compiler-wall} names the
\emph{mechanical} boundaries of the technique (template depth,
\cppinline{consteval} budgets, combinatorial blowup, diagnostics),
this subsection names the \emph{algebraic} boundary: which axioms
discharge cleanly on which carriers, where the obligation falls to
the engineer, and which frictions block the discharge. The empirical
artefacts of the showcases sit one level below this chain---they
demonstrate that the library realises the axiom~$\to$~theorem
discipline in C++23 today, on the carriers in the Rosetta tables of
Section~\ref{sec:carrier-rosetta}.

% ============================================================
\section{Discussion}
\label{sec:discussion}
% ============================================================


\subsection{The Compiler Wall: Where This Stops}
\label{sec:compiler-wall}

Treating \texttt{clang++} as a poor man's theorem prover is useful
precisely because the compiler is bounded, predictable, and cheap. It
is also, for the same reason, incapable of doing everything a real
proof assistant would. The purpose of this subsection is to be honest
about the wall and to name the \texttt{clang} knobs that push it back
by a constant factor but cannot remove it.

\paragraph{Compile-time budgets.}
Every reduction in this paper is a variadic template, so two
\texttt{clang} budgets matter: \emph{instantiation depth} (default
$1024$, raised to $2048$ on the affected translation units; concepts
and NTTP lambdas each add frames so the limit hits earlier than the
naive count suggests) and \emph{\cppinline{constexpr} steps} (default
$\sim\!10^6$, capping a single \cppinline{constexpr} evaluation).  A
$2\times 2$ LP with four constraints is comfortably under both; a
$2\times 2$ LP with thirty constraints ($\binom{30}{2} = 435$ active
sets and nontrivial metadata) starts approaching the step limit.
Raising the caps is possible (\cppinline{-ftemplate-depth=N},
\cppinline{-fconstexpr-steps=N}) but trades compile time and resident
memory for reach; the more principled answer is to stop at the
boundary and hand off to a runtime solver --- which is exactly the
case our centrepiece is \emph{not}.

\paragraph{Combinatorial blowup is the pessimistic bound.}
Compile-time vertex enumeration over $n$ halfspaces is, in the worst
case, $\binom{n}{d}$ active-set solves for a $d$-dimensional LP
(six for $n = 4$, $190$ for $n = 20$, nearly half a million for
$n = 1000$).  Two facts soften this substantially in practice.  By
the fundamental theorem of linear programming, the optimum of a
$d$-dimensional LP is attained at a vertex with at most $d$ binding
constraints, and on a handwritten polytope most of the syntactic
constraint list is slack or redundant; the $\binom{n}{d}$ figure
counts enumeration cost, not what any one vertex actually witnesses.
Many of those redundancies are also \emph{statically} removable: the
library's \cppinline{structured_and} dispatch already collapses
linearly-dependent or contradictory halfspaces at compile time
(Showcases~3 and~4), so the combinatorial term tracks the pruned
count rather than the raw
one.\footnote{Concretely:
\cppinline{x > 5 && x > 7} collapses to \cppinline{x > 7} (stricter
pivot wins); \cppinline{x > 5 && x < 2} collapses to $\varnothing$
(contradiction); \cppinline{x > 3 && x < 5} over an integral carrier
collapses to \cppinline{Singleton<4>}.  The 2D extension --- pruning
halfspaces whose normals are linearly dependent --- follows the same
dispatch pattern.}
The instance regime this paper targets --- MPC steps with a handful
of physical constraints, sensor-fusion guards, parametric
configuration LPs --- rarely exceeds a few dozen raw constraints
before pruning, and remains comfortably tractable.  Industrial-scale
LP is emphatically not in scope; we do not claim it is.

\paragraph{Active-set hints as the user-side escape valve.}
Even within the compile-time regime, a user with domain knowledge of
\emph{which} halfspaces bind at the optimum can short-circuit the
$\binom{n}{d}$ enumeration entirely: because the constraint pack is
already an NTTP list, instantiating
\cppinline{maximize<T, cx, cy, H_i, H_j>()} with just the binding
pair (instead of the full pack) is exactly the active-set hint, and
the library evaluates only that single Cramer solve.  An explicit
\cppinline{with_active_set<...>} API that takes the hint as a separate
template argument and verifies feasibility against the full pack
without re-enumerating is a natural successor surface, but the
underlying NTTP-pack-as-hint mechanism is available today: dropping
slack constraints from the syntactic list is the user's prerogative,
and the combinatorial term then tracks the hinted count rather than
the raw one.

\subsection{Boundary Observations: An Axiom~\texorpdfstring{$\to$}{->}~Theorem Chain}
\label{sec:boundary-observations}

The preceding sections present the technique in an empirical sort of way: through worked showcases, IR artefacts, and concrete diagnostics.
The library admits a stricter framing whose structural
parallel is Wadler's \emph{Theorems for Free!}~\cite{wadler1989theorems}:
parametricity gives the typed-lambda calculus theorems from type
signatures alone; the dedekind library extends the move into
\emph{algebraic} territory, where a carrier's axiomatic profile
(declared via opt-in traits and concept witnesses) discharges
carrier-level theorems at the type level. The empirical artefacts of
the showcases are then \emph{realisations} of an underlying
axiom~$\to$~theorem chain rather than the substance of the chain
itself.

\paragraph{Axioms as the engineer's honesty obligation.}
A carrier's algebraic profile is asserted explicitly. Where the
proof of an axiom is beyond what C++ concepts can quantify, the claim
is made via an opt-in trait specialisation, and the engineer who
declares the trait carries Wadler's
\emph{propositions-as-types}~\cite{wadler2015propositions}
obligation under the NSPE Code of Ethics canons III and IV (truthful
public statements; faithful agency). The library ships several such
opt-in surfaces:
\cppinline{is_monic_arrow_v} for monomorphism declarations,
\cppinline{is_initial_ring_v} for the initial-object reading of
$\mathbb{Z}$, \cppinline{is_grothendieck_group_v} for the
free-abelian-group reading, \cppinline{is_initial_f_algebra_v} for the
NNO-as-initial-F-algebra reading, and a family of axiom-shape concepts
including \cppinline{IsRing}, \cppinline{IsField},
\cppinline{IsTotallyOrdered}, \cppinline{IsArchimedean}, and
\cppinline{IsCyclicGroup}. Each is the engineer's claim, not a
machine-discharged proof; the structural shape (signatures, closure,
opt-in trait set to \texttt{true}) is what the type system can pin
mechanically.

\subsection{Related Work}
\paragraph{Functional and dependently-typed approaches.}
Agda~\cite{norell2007towards} and Coq provide dependent type systems in which set-membership
proofs are first-class terms.  The expressiveness is greater than C++23, but the
runtime model differs fundamentally: proof terms are erased at runtime, whereas
\texttt{dedekind}'s structural pruning targets machine-code generation in a
systems-programming context.
Haskell~\cite{GHC2024} supports type-class-based structural reasoning, and libraries
such as \texttt{algebra} encode ring and field hierarchies.  C++ templates differ in that
they operate on concrete machine types (with hardware-level algebraic imperfections)
rather than erased polymorphic dictionaries.

\paragraph{Exact arithmetic.}
The GNU Multiple Precision library (GMP) and \texttt{mpfr} support arbitrary-precision
arithmetic at runtime.  \texttt{iRRAM}~\cite{mueller2001irram} and similar libraries
implement computable real arithmetic.
\texttt{dedekind} is complementary: exact reals (Dedekind cuts) are modelled
\emph{intensionally as types}, and realization to a specific arithmetic backend is
an explicit programmer choice.

\paragraph{Set-builder type systems.}
Refinement types~\cite{freeman1991refinement} attach predicates to types, enabling
a form of set-builder representation in ML-family languages.
Liquid Haskell~\cite{vazou2014refinement} extends Haskell with SMT-backed
refinement predicates.
\texttt{dedekind} achieves a subset of this functionality in standard C++23 without
a custom type checker, using NTTP-embedded lambdas as the predicate carrier.

\paragraph{Linear programming.}
The standard LP toolchain---simplex~\cite{dantzig1963simplex} and
interior-point methods~\cite{karmarkar1984interior} as implemented in
Gurobi, CPLEX, GLPK, and related solvers---treats the problem instance
as runtime data: the objective vector, constraint matrix, and right-hand
side are passed to a solver that iterates to a vertex. \texttt{dedekind}
sits orthogonally: problem instances whose size and coefficients are
\emph{compile-time data} reduce via type-level vertex enumeration to
the optimum as a typed constant (Section~\ref{sec:lp-centrepiece}), with
the entire solver collapsing out of the emitted binary. This is not a
replacement for runtime LP---for large, data-dependent instances the
classical pipeline remains indispensable---but a complement for
instances that appear during code generation (problem-shape metadata,
compile-time configuration polytopes, template-parametric optimizer
settings). Exact-rational arithmetic over $\mathbb{Q}$ connects the
approach to the exact-LP literature~\cite{gleixner2016exact}; the
specific contribution is the realisation at translation time with
zero runtime overhead, witnessed in the emitted LLVM~IR fixtures.

\paragraph{Automatic differentiation.}
Runtime AD libraries such as Ceres~\cite{ceres-solver},
Adept~\cite{hogan2014adept}, Sacado, dco/c++, and
CasADi~\cite{andersson2019casadi} provide forward- and reverse-mode AD
via expression templates or operator overloading on a specialised
arithmetic type. The theoretical view taken by
Elliott~\cite{elliott2018simple}---that forward-mode AD is the
$(\text{value}, \text{tangent})$ product with chain-rule composition,
not the dual-number object \emph{per se}---clarifies why the same
reduction that solves the LP also computes its sensitivities when
instantiated over the right ring (Section~\ref{sec:ad-elliott}).
\texttt{dedekind}'s contribution is not a new AD algorithm but the
observation that a single compile-time reduction, written once over an
\texttt{HasRingOperators} carrier, serves as LP solver over $\mathbb{Q}$ and
as forward-mode AD engine over $\mathbb{Q}[\varepsilon]/(\varepsilon^2)$
with no extra code path.

\paragraph{Embedded machine-learning runtimes and edge computing.}
The embedded / edge-computing~\cite{shi2016edge} landscape has a
mature runtime ecosystem that \texttt{dedekind} deliberately sits
\emph{beside} rather than competes with. TensorFlow Lite
Micro~\cite{david2021tflm} compiles models to a small-footprint
interpreter targeted at microcontroller-class devices; ONNX Runtime
and microTVM generate target-specific code from a model graph; the
Eigen linear-algebra library is the de facto substrate for embedded
numerics. These runtimes take the model as input and the target as
fixed; they do not treat the hosting application's algebraic
invariants as proof obligations for the C++ compiler to discharge.

A deliberate choice on the linear-algebra backend is worth flagging
here, since it connects the library's position to the
abstract-naturals-versus-IEEE-numerics tension of
Section~\ref{sec:numeric-tension}. Eigen~\cite{eigen2010} is
mature, widely deployed, and excellent at what it targets---but it
is also \emph{carrier-opinionated}: its matrix types default to
\texttt{float} or \texttt{double} and its optimisations assume a
field-like IEEE carrier. A natural path for \texttt{dedekind}'s
runtime linear-algebra layer is instead toward
GraphBLAS~\cite{kepner2011graphalgs, davis2019graphblas}, whose
explicit \emph{(carrier, $+$, $\times$)} semiring parameterisation
makes the ring-over-which-matrices-live a first-class configurable
property. That matches the discipline this paper argues for:
matrices over \cppinline{bool} (with OR/AND) are perfectly
well-defined objects when inversion is not required, matrices over
\cppinline{char} are useful for quantised embedded inference, and
both sit naturally inside the \cppinline{HasRingOperators} /
\cppinline{HasSemiringOperators} concept hierarchy. The cost of this
generality is library complexity rather than carrier simplicity.
The trade we advocate is correctness-and-performance (tight control
over the ring structure) at the price of implementation effort,
against Eigen's opposite trade---an ergonomic default carrier
whose optimisations are largely locked to floating-point semantics.

\texttt{dedekind} addresses a different surface: the \emph{inline}
numerical primitives (Kalman filters, pose-estimation math, MPC
steps, parametric optimisation) a robotics, automotive, or
instrumentation engineer writes directly in their control loop. For
these, a fixed-topology LP, a compile-specialised Kalman update,
or an exact-rational sensor-fusion step expressed against
\texttt{HasRingOperators} concepts exhibits three properties the
model-runtime path does not aim at: a binary with no
interpreter, a dependency graph no heavier than the system's
\texttt{clang++}, and bit-for-bit deterministic arithmetic across
platforms when the carrier is $\mathbb{Q}$. We see this as a
complement to the runtime ecosystem, not a replacement: model-driven
inference continues to belong to TFLite Micro and its siblings, while
the numerical glue around such inference---the Kalman front-end, the
control-loop LP, the gradient for online adaptation---is where
Axiomatic Systems Programming is strongest.

\paragraph{Data-pipeline DSLs.}
Apache Spark, Flink, and the Python \texttt{pandas}/\texttt{polars} ecosystem
provide declarative data pipelines with runtime domain checks. None of these
ground their APIs in formal set theory, so intersections and filters are
runtime pipeline operations rather than compile-time type invariants.
\texttt{dedekind} is positioned one layer down: it is a systems-programming
library that could in principle back such pipelines, but the contribution
we demonstrate here is the compile-time reduction at the C++ layer itself.

\paragraph{C++ template metaprogramming lineage.}
Compile-time computation in C++ has a long tradition that
\texttt{dedekind} builds on rather than supplants. Boost.Hana
~\cite{chagnon2016hana} and \texttt{mp11}~\cite{pdimov2017mp11} provide
compile-time heterogeneous containers and type-level algorithm
support; Hana in particular already demonstrates concept-based
structural dispatch at compile time and would be a natural
lower-layer substrate for several of our combinators.
CTRE~\cite{dusikova2018ctre} shows that serious computation
(regular-expression matching, compiled to a type-level automaton)
can live entirely inside the type system and emit optimised
machine code; the shape of that reduction---intensional description
to a named extensional object at translation time---is close kin to
our halfspace and LP reductions, differing in subject matter but
not in mechanism. Expression-template AD libraries including
ENOKI~\cite{jakob2019enoki} and the runtime AD systems
Ceres~\cite{ceres-solver}, Adept~\cite{hogan2014adept},
and CasADi~\cite{andersson2019casadi} have carried pieces of the
dual-number story into production workloads.

\texttt{dedekind}'s position relative to this lineage is specific
and narrow: it targets \emph{algebraic-semantic} content---ring
homomorphisms, structural cardinality, ordering laws, the
$(f, f')$ product as an Elliott~\cite{elliott2018simple} sibling---rather
than the syntactic structure manipulation that is the traditional focus
of TMP libraries. The contribution is not another metaprogramming
trick but the claim, supported by worked reductions, that the
\emph{discipline} of treating mainstream C++23 as a structural-gate
theorem prover (Axiomatic Systems Programming) is now coherent enough
to practise end-to-end.
% ============================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================

The thesis of this paper, once more: \emph{named algebraic structure
at the type level, discharged by the compiler, narrows the gap
between abstract mathematics and machine code without a separate
proof assistant.} \texttt{dedekind} is the existence proof. Sets are
rules rather than buckets; intensional descriptions defer
materialisation to explicit boundaries; the LLVM backend carries
discharged proofs through to object-code optimisation. The
centrepiece existential proof (Section~\ref{sec:lp-centrepiece}) shows
a textbook linear-programming instance reducing at compile time to
its optimal vertex as a typed constant---the six candidate solves,
six feasibility checks, and six objective comparisons vanish in the
emitted IR. The same reduction over the dual-number ring recovers
exact forward-mode sensitivities (Section~\ref{sec:ad-elliott}); at
the linear-algebra layer it certifies the ring homomorphisms
$\mathbb{C}, \mathbb{D} \hookrightarrow M_2(\mathbb{Q})$ as
compile-time facts (Section~\ref{sec:lin-alg-embedding}). The
companion repository's polynomial-calculus partitions discharge the
Fundamental Theorem of Calculus at translation time on the same
discipline. In every case the binary contains only literals.

\paragraph{Axiomatic Systems Programming as a practised discipline.}
The viewpoint we name in Section~\ref{sec:introduction}---treating
\texttt{clang++} as a poor man's theorem prover, and the
\texttt{HasRingOperators} concept as a proof obligation the carrier must
discharge---is the paper's intended export. Taken seriously, it
turns each algebraic invariant into a structural gate the compiler
polices; taken habitually, it shifts the library author's default
from ``runtime check'' to ``type-level obligation''. Three axes of
type-directed collapse remain open (Section~\ref{sec:gap}): carrier
tightening, smooth optimization, and lattice-law rewriting; each
sits on the same NTTP-plus-concepts foundation as the halfspace and
LP reductions demonstrated here.
We leave them as worked targets for successor libraries that adopt
the same discipline.

\paragraph{An absent failure mode worth reporting.}
A practical observation, offered honestly: across this library's
development we did not encounter the two failure modes mainstream
C++ template programming is famed for --- \emph{template
instantiation explosion} (compile times that grow super-linearly
with template depth) and \emph{template error hell} (diagnostics
buried inside long substitution-failure cascades, far from the
call site that triggered them).  Neither was a recurring concern.
This is not nothing: both phenomena are well-documented hazards
in production C++ template programming.  The disciplined fragment
of §\ref{sec:axiomatic} (Table~\ref{tab:admissibility-rules})
appears to rule out, by construction, the mechanisms that produce
each.  The concept-constrained-template machinery fails at
\emph{constraint-evaluation} time --- before template-body
instantiation, with named-axiom diagnostics --- so the SFINAE-
cascade error path is excluded.  Uniform parametric polymorphism
(System-F-style; no specialisation-driven branching admitted in the
public surface) keeps the set of compiled code paths linear in the
number of callsites rather than combinatorial in the type tuples.
And the bounded-instantiation-depth admissibility rule (capped at
the carrier-lattice's three axes in the current surface) prevents
the recursively-unfolding template-metaprogramming pattern that
produces the explosion.  Whether this implies a \emph{formal}
result --- a theorem that the disciplined fragment lies inside a
decidable type-checking regime --- the paper does not claim.  The
fragment is small enough that decidability is plausible
(Lambek--Scott~\cite{lambek1988introduction} maps the constructive
half to a CCC fragment; the propositions-as-types
reading~\cite{wadler2015propositions} restricts to inhabited
propositional types), but a proof would have to argue against
the full C++ overload-resolution and ADL machinery and lies outside
the scope of this engineering existence-proof.  We report the
observation; the explanation we leave to the type-systems
literature.

\paragraph{A forward claim, in its weakest defensible form.}
As numerical workloads saturate available hardware, the marginal
returns of brute-force optimisation are diminishing, and the lever
with a different cost curve is specialisation at translation time
rather than run time. That lever also has a sustainability
dimension: cycles not spent at run time are watt-hours not drawn
from the grid, and the machine-learning research community has begun
to measure its compute footprint in exactly those
terms~\cite{schwartz2020greenai}. The operating point at which the
lever compounds most visibly is not the data centre but the edge:
CPU-first and embedded machine-learning
deployments~\cite{shi2016edge, david2021tflm}, where the problem
topology is fixed, the runtime footprint is under hard constraint,
and the toolchain of least resistance is the system's
\texttt{clang++}. We do not prove here
that axiomatic discipline applied at library scale yields a
meaningful fraction of cycles back at either end of the spectrum;
we claim only that the ingredients---compile-time verification,
structural optimisation, ring-aware code generation, exact
arithmetic carriers---are now assembled in mainstream C++23, and
that the discipline is worth practising precisely because its
payoffs accrue at scales larger than any single library. If even a
small fraction of the arithmetic done in contemporary edge
deployments---model-predictive control in vehicles, Kalman-style
sensor fusion in robotics, gradient evaluations in micro-inference
loops---ran against libraries whose algebraic invariants had been
discharged upstream, the compounded effect would be difficult to
dismiss. The shape of that opportunity is what Axiomatic Systems
Programming claims. Whether the shape materialises into the numbers
it suggests is an engineering question, and one the next decade of
library authors will answer.

\paragraph{Reproducibility under heterogeneous tooling.}
The discipline this paper names is, deliberately, falsifiable.  The
constraints under which it stays honest---textbook-anchored
vocabulary, a named-module dependency graph, structural assertions
discharged at translation time, a CI-policed test suite---do not
depend on which agent typed any particular line of code, nor on the
specific host language.  As Section~\ref{sec:axiomatic} notes, the
technique transfers with modest adjustment to any
strongly-typed-and-tested substrate whose type system can name a
predicate over compile-time data and check it mechanically: Scala
under refined types and a strict build, C\# with constrained
generics under Roslyn analysers, Java with sealed types, strictly-
typed Python under \texttt{mypy} or \texttt{pyright}.  C++23 is the
worked example here---chosen because of the LLVM lever from
discharged proof to optimised object code---but the discipline does
not turn on it.  A practitioner adopting the same constraints and a
comparable toolchain in 2026 should be able to produce comparable
artefacts on comparable timescales, irrespective of host language or
of how the keystrokes were generated.  The discipline is meant to
outlast any particular generation of authoring tool---and,
increasingly, of programming language.

% ============================================================
\section*{Acknowledgements}
\label{sec:acknowledgements}
% ============================================================

The development of \texttt{dedekind} relied substantively on AI assistance throughout, and the writing of this paper did so as well.
The disclosure here is the concrete instance of the cost-benefit-calculus claim in §\ref{sec:introduction}: that paragraph names the shift in the abstract; this one names how it shifted on this artefact.
Per the ‹Programming› journal's policy on AI tools~\cite{programming_journal_ai_policy}, the use of these tools is acknowledged and disclosed; the AI systems named below are tools whose contribution is recorded here as assistance, not authorship.
The author is a single human person, named in the byline, who bears full responsibility for content, scope, correctness, and editorial judgement.

The primary assistant was Anthropic's \emph{Claude} family (Sonnet and Opus across the project's lifetime; the 1M-context Opus 4.7 variant for the most recent slices, where reproducibility against extended-context interactions matters), invoked through the Claude Code CLI.
Google's Gemini was used for second-opinion structural critique on substantive mathematical claims (size-restriction notation, adjunction-encoding feasibility, foundational-vs-mechanical disentanglement).
GitHub Copilot's pull-request reviewer was invoked on every PR for scope-vs-claim consistency checks at the prose / table / listing level.
The assistance covered draft generation (prose, code, listings, tabular structured elements), structural refactoring (concept naming, partition placement, hub-and-spoke dispatch shape), test scaffolding, review-thread response drafting, and CI / GitHub workflow scaffolding.

The pattern is visible at commit-level granularity in the public repository at \url{https://github.com/vincentk/dedekind}: every commit message, every pull-request review thread, and the public CI build log compose into a complete audit trail.
Reviewers can inspect the assistance pattern at any commit.
``Assisted'' rather than ``co-authored'' is the operative voice throughout, both for this paper and for the codebase it documents.

\printbibliography

\appendix

\section{Continuous Iteration towards the Faithful Translation}
\label{sec:eventually}

The development follows a Top-Down Refinement model~\cite{wirth1971stepwise}, where the library acts as an Executable Ontology: the mathematical Forms are anchored upstream, machine carriers are staged downstream, and each step in between is a refinement the compiler is asked to police.
We treat the alignment between C++ types and mathematical axioms as an Eventual Consistency target~\cite{vogels2009eventually}, where the Semantic Gap is incrementally closed through the addition of Compile-Time Constraints rather than asserted globally up front.
The framing is deliberately borrowed from distributed-systems literature: the system holds multiple replicas of the same fact (a Form, a concept, a trait specialisation, a machine type); the replicas may temporarily disagree (a carrier may not yet witness an axiom the Form claims); convergence is monotone under bounded operations, in the sense that each pull request can only \emph{tighten} the alignment, never loosen it.
The quiescent state is the one in which every node in the partition DAG is green; the Semantic Gap is whatever sits between the current state and that target at any given moment.
The §\ref{sec:introduction} framing --- ``aim at, not assert'' --- is the same statement read from this angle: the bridge $\mathcal{F}: \mathbf{Cat} \to \mathbf{Cpp}$ is the convergence target, and each concrete carrier's \cppinline{static_assert} discharge is a partial inhabitant accumulating against it.

To support this approach in a transparent fashion, we reach deep into the CI/CD toolbox and enable whatever we can find: an open-source licence and public review for the social audit trail; public issue management for the backlog of refinements still owed; a reproducible build pipeline that re-runs the compile-time discipline on every push; a test harness whose \cppinline{static_assert} obligations are checked mechanically; pull-request reviews that frame each new trait specialisation as a public claim under engineering ethics (cf.\ §\ref{sec:introduction}'s NSPE-canon footnote); version control that pins the chain of refinements to specific commits; and test-coverage reports that flag the spokes of the hub-and-spoke architecture for which no carrier yet witnesses the trait.
None of this is technically novel.
What matters is that the engineering practices of mainstream open-source software development \emph{compose} cleanly with the type-system discipline, turning what would otherwise be a paper-only argument into an artefact whose every refinement step is reviewable, reproducible, and dated.

\section{The System F $\leftrightarrow$ Juliet Posture $\leftrightarrow$ Bourbaki Rosetta}
\label{app:jlt-systemf}

§\ref{sec:jlt} reports the System-F reading as a side effect of the Juliet Posture rather than a design goal: the disciplined subset chosen for alignment with $\mathbf{Alg}(\Sigma)$ and clean integration with \cppinline{std::regular} carriers turns out to be a faithful image of System~F (Pierce~\cite{pierce2002tapl} ch.~23, 29) and of the simply-typed $\lambda$-calculus under Lambek--Scott~\cite{lambek1988introduction}, and the same fragment lines up with Bourbaki structuralism~\cite{marquis2022structuralist}.
The triangulation across three independently-developed traditions is the kind of evidence the structuralist tradition takes seriously --- not because any single tradition committed to all the others' conclusions, but because the alignment was not engineered for and fell out gratis on the disciplined fragment.
This appendix records the artefact: the admissibility rules an engineer must follow for the System-F reading to hold, and a row-by-row mapping across the three vocabularies.

\paragraph{Admissibility rules.}
``Carefully selected'' is not informal: the library polices a fixed list of admissibility rules on every \cppinline{export}-qualified declaration.
Each rule closes a specific escape hatch by which C++ would otherwise escape its System-F reading (Table~\ref{tab:admissibility-rules}); together they delineate the disciplined fragment over which the rosetta of Table~\ref{tab:structuralist-bridge} holds.
The list is not novel --- working C++ teams police variants of it informally --- but the contribution is that the rules are what make the rosetta accurate, and that each rule has a mechanical witness in the partition graph (the public-API headers exported by each \cppinline{export module} declaration are the audit surface).

\begin{table}[htbp]
\centering
\caption{\textbf{The admissibility rules}: which C++ idioms are admitted
in the library's public surface, and what each rule rules out.
The discipline is the price of the System-F reading; each rule below is
the cost of one row of the structuralist-bridge rosetta
(Table~\ref{tab:structuralist-bridge}).  Internal
partitions may use the listed escape hatches in bounded scopes, but
they do not appear in any \cppinline{export}-qualified signature.}
\label{tab:admissibility-rules}
\small
\begin{tabular}{@{}p{0.42\textwidth} p{0.52\textwidth}@{}}
\toprule
\textbf{Rule on the public surface} & \textbf{What it rules out, and why} \\
\midrule
No \cppinline{reinterpret_cast} in any \cppinline{export}-qualified API. &
Defeats type abstraction at the value level.  Bit-level reinterpretation
(e.g., IEEE-754 inspection) lives in non-exporting internal partitions. \\
No SFINAE branches in the public surface. &
Substitution-failure-as-overload-selection makes the same call
mean different things at different call sites.  Concept-constrained
templates fail with named-axiom messages instead. \\
No ADL-dependent dispatch except for textbook operator overloads. &
Argument-dependent lookup is admitted only where the textbook reads
\texttt{a + b} or \texttt{a == b} on library carriers; free-function
calls in user code dispatch through explicit qualification. \\
No template-template parameters in \cppinline{export}-qualified signatures. &
Higher-kinded structure is named at the level of concepts
(\cppinline{IsFunctor<F>}, etc.), not by passing template-templates
through public function signatures. \\
No public-API specialisation that branches behaviour. &
Generic templates are parametric across the public surface; per-type
specialisations live in the trait registry, each carrying an
algebraic claim that is reviewed and CI-checked against a
\cppinline{static_assert}. \\
Bounded template instantiation depth. &
The Turing-complete instantiation machinery is bounded by the
carrier-lattice depth (Figure~\ref{fig:carrier-lattice}: scalar,
shape, quotient axes; depth $\leq 3$ in the current surface), which
keeps the type checker's work within a decidable cone. \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{The structuralist bridge.  Each row reads three ways: a System F property (logic / type-theory tradition, Pierce~\cite{pierce2002tapl}~ch.~23, 29 + Wadler~\cite{wadler1989theorems}); the Juliet-Posture restriction in C++23 that recovers it; and the corresponding Bourbaki-structuralist move (Marquis~\cite{marquis2022structuralist}; Bourbaki's \emph{L'Architecture des mathématiques} 1948).  The triangulation is the point: each \textbf{Jlt} realisation is simultaneously a recovery of a System F property and a Bourbaki structuralist move.  The C++ column states the \emph{discipline} an engineer adopts; the consequences live in the outer columns.}
\label{tab:structuralist-bridge}
\begin{tabular}{@{}p{0.30\textwidth}p{0.32\textwidth}p{0.32\textwidth}@{}}
\toprule
\textbf{System F property} & \textbf{Juliet Posture (C++)} & \textbf{Structuralist style (Bourbaki)} \\ \midrule
\textbf{Parametricity} (\emph{Theorems for Free!}~\cite{wadler1989theorems})\footnote{C++23 concepts realise parametricity in an \emph{implicit-style} discipline: the constraint is checked at instantiation but no concept witness is passed at runtime.  This is closer to Haskell-style type classes than to pure System F's explicit type abstraction, but the parametricity conclusions follow once the public surface excludes ad-hoc dispatch (Table~\ref{tab:admissibility-rules}'s no-SFINAE / no-ADL-beyond-textbook / no-public-specialisation rules).}  &
\textbf{Concept-gating}: \cppinline{requires}-clauses forbid the function from inspecting type internals; interaction happens only via the allowed \emph{scents} (axiom-shape concepts). &
\textbf{Typification}: ignoring internal set-elements in favour of relations among them. \\ \addlinespace
\textbf{Universal quantification} ($\forall\alpha.\dots$) &
\textbf{Constrained templates}: \cppinline{template<Scent T>} reads as universal quantification over the species defined by the concept \emph{Scent}. &
\textbf{Species of structure}: the genus of all objects sharing a given structure. \\ \addlinespace
\textbf{Type erasure} (computation is type-blind) &
\textbf{Zero-cost abstraction}: \cppinline{std::regular} objects with transparent value semantics let the compiler optimise on structure while the program's logic remains independent of the carrier name. &
\textbf{Indistinguishability}: isomorphic structures are functionally identical. \\ \addlinespace
\textbf{Strong normalisation} (termination)\footnote{C++ template instantiation is Turing-complete in general~\cite{veldhuizen2003templates}; strong normalisation is recovered specifically by Table~\ref{tab:admissibility-rules}'s bounded-instantiation-depth rule and the exclusion of template-template-parameter recursion in public signatures.  These restrictions are the price of the strong-normalisation reading.}  &
\textbf{Structural pruning}: rejecting Turing-complete template recursion in favour of finite algebraic compositions targets a fragment guaranteed to terminate (and to collapse logically empty constructions to zero machine instructions). &
\textbf{Echelon construction}: finite, well-founded builds from base sets and axioms. \\ \addlinespace
\textbf{Impredicativity} (types may be applied to types)\footnote{System F has prenex-style impredicativity at the kind level; C++23 concepts admit a practical form by allowing concepts to be defined in terms of other concepts (e.g.\ \cppinline{IsFunctor<F>} ranging over type-templates).  The two coincide on the fragment we use; the broader impredicative core (universe-level self-quantification) is not in scope.}  &
\textbf{Meta-concepts}: concepts defined in terms of other concepts let the library name 2-arrow structures (morphisms between morphisms) at the type level. &
\textbf{Higher-order structure}: structures built upon structures. \\ \addlinespace
\textbf{Ad-hoc rejection} (a function does not behave differently for \cppinline{int} vs \cppinline{float}) &
\textbf{Ban on ADL / specialisation}: ADL beyond textbook operators, public-API specialisation, and SFINAE branching are excluded so that lineage cannot hijack scent (Table~\ref{tab:admissibility-rules}). &
\textbf{Axiomatic purity}: rejection of ``substance'' or non-law behaviour at the level of the structure. \\ \bottomrule
\end{tabular}
\end{table}

\paragraph{Three load-bearing threads.}
Three threads run through the rosetta and stand on their own as the substantive claims of the System-F recovery:

\begin{enumerate}
\item \textbf{The postural ban on overloading.}  System F does not admit ad-hoc polymorphism: a function cannot behave differently for \cppinline{int} versus \cppinline{float}.
$\mathbf{Jlt}$ recovers this by banning ADL beyond textbook operator overloads and public-API specialisation.
If a function is defined for \cppinline{Scent T}, it must work the same for every rose that smells like \cppinline{T}.
\item \textbf{Axiomatic honesty.}  System F assumes its types follow their definitions perfectly.
$\mathbf{Jlt}$ recovers this by auditing carriers (e.g.\ rejecting \cppinline{int} for ring operations where overflow would defeat associativity) so the carrier-rosetta (Tables~\ref{tab:rosetta-quotient} and~\ref{tab:rosetta-rank}) matches the textbook expectations rather than papering over them.
\item \textbf{Transparency over encapsulation.}  System F types are \emph{transparent} to the calculus.
$\mathbf{Jlt}$ recovers this by rejecting OO-style encapsulation and private state, so type-directed collapse can see all the way through an object to perform formal reductions.
\end{enumerate}

Taken together, these features give \texttt{clang++} a workflow that carries a striking resemblance to a System F type-checker, with optimised machine code falling out as a side effect rather than as a separate compilation step.\footnote{The well-known alternative to this \emph{disciplined embedding} is the \emph{two-language problem}~\cite{bezanson2017julia}: research code is written in an expressive high-level language (Python, R, Julia) while only performance-critical execution is delegated to a separately maintained C or C++ backend.  Bridging tools such as \texttt{nanobind}~\cite{jakob2022nanobind} close the \emph{binary} gap but not the \emph{semantic} gap: the C++ backend cannot reason about high-level mathematical invariants defined in the Python layer.  The Juliet Posture is the alternative: keep the high-level mathematical vocabulary in the same language the optimiser sees.}

\paragraph{An old capability, newly accessible.}
None of the calculus underneath is new.
Templates have always been a System-F-flavoured substrate \emph{within a carefully selected subset of C++}: type-parameter polymorphism, $\beta$-style substitution via \cppinline{using} aliases, normalisation through \cppinline{constexpr} evaluation --- the Lambek--Scott correspondence~\cite{lambek1988introduction} and the Curry--Howard reading~\cite{wadler2015propositions} have applied to that fragment for as long as it has existed.
What changed in C++20/23 was not the lambda calculus underneath; it was the addition of \cppinline{concept}, \cppinline{requires}-clauses, and NTTPs, which together turned a long-standing latent capability into a surface the type checker can discharge mechanically.
Practitioners with deep template-metaprogramming experience --- Boost.Hana~\cite{chagnon2016hana} and CTRE~\cite{dusikova2018ctre} are the visible end of a long lineage --- have known this terrain for over a decade; the contribution of this paper is to walk it under the named-structure / discharged-by-the-compiler discipline, end-to-end, in standard C++23.
The translation table at §\ref{sec:cpp-stlc-mapping} is the dictionary between what was always there and what the modern surface now lets one say out loud.


\subsubsection{The Traits Registry: Abstract Naturals vs.\ IEEE Numerics}
\label{sec:numeric-tension}

Here is the tension in one sentence: the \texttt{int} in your source
file is not the $\mathbb{Z}$ in your textbook, and everybody quietly
agrees to pretend it is. The ring $\mathbb{Z}$ obeys a few laws:

\begin{itemize}
  \item Associativity of addition: $(a + b) + c = a + (b + c)$.
  \item Existence of additive inverse: $\forall a,\, \exists -a$ such that $a + (-a) = 0$.
  \item No upper or lower bound.
\end{itemize}

A 32-bit signed \texttt{int} obeys \emph{none} of them exactly: addition
overflows, negation of \texttt{INT\_MIN} is undefined behaviour, and the
representable range is finite. IEEE~754 \texttt{double}~\cite{ieee754_2019}
is further out still: addition is non-associative because rounding
happens every step, and \texttt{NaN} and $\pm\infty$ puncture the law of
excluded middle in the exact way Section~\ref{sec:sets-rules} will ask
the subobject classifier $\Omega$ to repair. A generic algorithm that
silently assumes the ring laws on these carriers is, in the literal
technical sense, wrong---it just happens not to be wrong on most inputs.

\texttt{dedekind} encodes these divergences explicitly in the \texttt{:species} trait registry:

% Source: src/main/modules/dedekind/category/species.cppm (partial specialisations around ll. 485-500)
\begin{listing}[htbp]
\caption{The species registry records algebraic properties---and their absence. Signed \texttt{int} is explicitly rejected as associative under \texttt{+} because signed overflow is undefined behaviour.}
\label{lst:species}
\begin{cpp}

// Variable template: is_associative_v<T, Op> is the structural trait.
// The library partial-specialises it per-carrier, per-operation.

// Unsigned integrals are associative under +  (modular arithmetic).
template <std::unsigned_integral T>
inline constexpr bool is_associative_v<T, std::plus<T>> = true;

// Signed integrals are REJECTED: overflow of a signed integer is
// undefined behaviour, so the addition operator is not a law-abiding
// monoid on signed carriers.
template <std::signed_integral T>
inline constexpr bool is_associative_v<T, std::plus<T>> = false;

// IEEE double is rejected for a different reason: rounding breaks
// bit-exact associativity.
template <>
inline constexpr bool is_associative_v<double, std::plus<double>> = false;

static_assert(is_associative_v<unsigned int, std::plus<unsigned int>>);
static_assert(!is_associative_v<int,         std::plus<int>>);
static_assert(!is_associative_v<double,      std::plus<double>>);
\end{cpp}
\end{listing}

Generic algorithms that require associativity will fail to compile for
\texttt{int} or \texttt{double} alike, unless an explicit numeric policy
(modular semantics on signed overflow, compensated summation, an
interval-arithmetic wrapper, or a lift to \cppinline{Rational<long>})
is supplied. The realization boundary is the point at which the
programmer chooses which policy to apply.


\subsubsection{Carriers: From Exact Rationals to Packed Integers}
\label{sec:carrier-rosetta}

A \emph{carrier} in this paper is the underlying set on which an
algebraic structure lives --- the substrate of operations.  The
universal-algebra formulation is introduced in §\ref{sec:alg-sigma}
and applies here.  The same set may serve as carrier for several
distinct algebras (e.g.\ $\mathbb{B}$ as $(\{0,1\}, \lor, \land)$
vs as $(\mathbb{F}_2, \oplus, \land)$); and the same algebra may be
obtained as either a primitive (a chosen carrier) or as the result
of a construction (a quotient, free, or localisation functor)
applied to a smaller carrier.  $\mathbb{Q}$ is
both a carrier (the rationals, used as base for
\cppinline{Rational<I>}) and the field-of-fractions construction
$\operatorname{Frac}(\mathbb{Z})$~\cite{lang2002algebra}; $\mathbb{C}$
is both a carrier and the quotient
$\mathbb{R}[i]/(i^2 + 1)$.  The Algebraic Lattice
(Figure~\ref{fig:carrier-lattice}) separates these two readings onto
distinct axes; the rosetta tables that follow itemise each axis.\footnote{%
\emph{Meets vs. joins.} Type-level meets (the largest shared
substructure of two algebras) are computed via dispatch ---
\cppinline{structured_and} on intersecting predicate-sets is the
canonical example.  Joins (the smallest common extension) are
\emph{not} synthesised by the library: pushouts / tensor products
require ambient choice (which base, which extension functor) that
the project's Honest Rejection stance refuses to make
automatically.  Joins are handled instead via named carrier-crossing
arrows the user invokes explicitly (\cppinline{embed_ℤ_ℚ} for
$\mathbb{Z} \hookrightarrow \mathbb{Q}$, etc.).  The trade-off is
honest: the library refuses to fabricate the answers that aren't
canonical.}

\paragraph{Re-enabling refused carriers.}  The Honest Rejection
policy refuses carriers like \cppinline{int} as a witness for
\emph{axiomatic} ring laws (signed-overflow UB defeats closure as a
stable axiom, per Listing~\ref{lst:species}).  An engineer with a
local proof that overflow cannot occur on their inputs has two
sanctioned re-entry paths:

\begin{itemize}
\item \emph{Operator-surface tier.}  \cppinline{HasRingOperators<int>}
fires unconditionally; downstream combinators that bind to this
weaker concept (rather than to the strict \cppinline{IsRing} bundle)
admit \cppinline{int} directly.  Most paper showcases bind to the
operator-surface tier for exactly this reason.
\item \emph{Opt-in trait specialisation.}  Where a stronger axiom is
needed, the engineer specialises the relevant species trait
explicitly: e.g.\ \cppinline{is_associative_v<MyType, std::plus<>> = true;}
for a user-defined type whose addition is provably associative.  The
specialisation is a public statement under the truthfulness canons
(NSPE Code of Ethics, Canons~III + IV \cite{nspe_ethics}); the
engineer carries the proof obligation, the library trusts the trait.
\end{itemize}

The mechanism is intentionally narrow: the engineer can re-enable a
specific (carrier, axiom) pair, but cannot re-enable an entire
concept bundle without specialising each axiom trait individually.
Pedantic by default; opt-in by deliberate engineer action.

The intensional-first rule applied to \emph{every} numeric role at
once is spelled out in three tables. Table~\ref{tab:carriers-exact}
catalogs the \emph{exact} side: the library's own intensional
carriers for each mathematical role, and the algebraic concept each
realises. Table~\ref{tab:carriers-primitive} catalogs the
\emph{primitive} side: the \cppinline{std}-library types the hardware
runs natively, and the concept each of them carries (with the usual
IEEE / overflow caveats). Table~\ref{tab:carriers-crossings} is the
state-machine that connects the two: given a role, which named
embedding or realisation arrow moves you from exact to primitive, and
which moves you back.

\begin{table}[htbp]
\centering
\caption{\textbf{Exact carriers.} The library's intensional types, one
per mathematical role, together with the algebraic concept each
satisfies. These are the carriers a disciplined engineer writes the
\emph{reference implementation} against --- laws are provably
satisfied, not approximated.}
\label{tab:carriers-exact}
\small
\begin{tabular}{@{}l l l@{}}
\toprule
\textbf{Role} & \textbf{Exact carrier} & \textbf{Concept satisfied} \\
\midrule
$\mathbb{B}$ & \cppinline{bool} under $\land, \lor$ & \cppinline{IsBoolean} \\
$\mathbb{F}_2$ & \cppinline{bool} under $(\oplus, \land)$ & \cppinline{IsGaloisField<bool,^,&>} (order 2) \\
$\mathbb{F}_{64}$ & \texttt{𝔽64} ($\mathbb{F}_2[x]/(x^6{+}x{+}1)$, 6-bit in \cppinline{uint8_t}) & \cppinline{IsGaloisField<T,+,*>} (order 64) \\
$\mathbb{F}_2^{64}$ & \cppinline{uint64_t} under $\oplus$, scalar action \cppinline{Bit2U64Action} & \cppinline{IsVectorSpace<uint64_t, bool>} \\
$\mathbb{N}$ bounded & \cppinline{ExtensionalCardinal<N>} & \cppinline{IsCyclicGroup}, \cppinline{IsAbelianGroup<T,+>} \\
$\mathbb{N}$ unbounded & \cppinline{Cardinality} (variant over $\aleph_0$) & saturating commutative monoid via \cppinline{add}/\cppinline{mul} helpers \\
$\mathbb{Z}$ bounded & \cppinline{SignedExtensionalCardinal<N>} & \cppinline{IsAbelianGroup<T,+>}, \cppinline{IsRing} \\
$\mathbb{Z}$ unbounded & \cppinline{SignedCardinality} (variant over $\pm \aleph_0$, NaZ) & \cppinline{IsAbelianGroup<T,+>}, \cppinline{IsRing} \\
$\mathbb{Q}$ & \cppinline{Rational<SignedExtensionalCardinal<>>} & \cppinline{HasFieldOperators<Rational<I>>}\textsuperscript{$\dagger$} \\
$\mathbb{R}$ exact & \cppinline{ExactReal<I>} = \cppinline{Real<Rational<I>>} & \cppinline{IsDedekindComplete} (exact rational scalar) \\
$\mathbb{C}$ & \cppinline{Complex<Rational<SignedExtensionalCardinal<>>>} & \cppinline{Algebra_}$\mathbb{C}$ \\
$\mathbb{D}$ (dual) & \cppinline{Dual<Rational<SignedExtensionalCardinal<>>>} & \cppinline{IsRing<T,+,*>}, $\varepsilon^2 = 0$ \\
\bottomrule
\end{tabular}

\vspace{0.3em}
\begin{flushleft}
{\footnotesize \textsuperscript{$\dagger$} The operator-surface concept \cppinline{HasFieldOperators<Rational<I>>} fires today; the strict axiomatic \cppinline{category::IsField} witness on $\mathbb{Q}$ is parked behind an architectural gate (the \cppinline{IsTotal} ladder admits periodic / idempotent / saturating paths but not yet an exact path).  The operator surface is what ships; the strict axiomatic certificate is queued.}
\end{flushleft}
\end{table}

\begin{table}[htbp]
\centering
\caption{\textbf{Primitive carriers.} The \cppinline{std}-library types
the hardware runs natively, with the algebraic concept each carries
under its C++-standard arithmetic semantics. These are the carriers a
disciplined engineer drops to for performance, \emph{after} verifying
the reference implementation on the exact side.}
\label{tab:carriers-primitive}
\small
\begin{tabular}{@{}l l l@{}}
\toprule
\textbf{Primitive type} & \textbf{Arithmetic semantics} & \textbf{Concept satisfied} \\
\midrule
\cppinline{bool} & Galois field $\mathbb{F}_2$ under $(\oplus, \land)$; Boolean lattice under $(\lor, \land)$ on the same carrier & \cppinline{category::IsField<bool, std::bit_xor<bool>, std::bit_and<bool>>}; \cppinline{order::HasLatticeOperators<bool>} \\
\cppinline{unsigned long}, \cppinline{size_t} & Modular, wraps at $2^{N}$ & \cppinline{IsCyclicGroup}, \cppinline{IsAbelianGroup<T,+>} \\
\cppinline{int}, \cppinline{long} & Partial (UB on signed overflow) & \cppinline{HasRingOperators}\textsuperscript{a} \\
\cppinline{int8_t}, \cppinline{int16_t} & Functor-surface only (literal arithmetic promotes to \cppinline{int}) + SIMD-vectorisable & \cppinline{IsAlgebra<T, std::plus<T>, std::multiplies<T>>}\textsuperscript{b} \\
\cppinline{float}, \cppinline{double} & IEEE 754, rounded, non-associative & \cppinline{Continuum_}$\mathbb{R}$ (non-assoc.\ \cppinline{+}) \\
\cppinline{std::complex<double>} & Cartesian pair over IEEE \cppinline{double} & \cppinline{Algebra_}$\mathbb{C}$ (non-assoc.) \\
\bottomrule
\end{tabular}

\vspace{0.3em}
\begin{flushleft}
{\footnotesize \textsuperscript{a} \cppinline{HasRingOperators} holds (literal $+$, binary $-$, unary $-$, $*$ all close on the carrier); the strict \cppinline{IsRing} bundle refuses because signed-overflow UB makes the operation partial, which doesn't satisfy any of the \cppinline{IsTotal} paths (\cppinline{IsPeriodic} / \cppinline{IsIdempotent} / \cppinline{IsSaturating}) the totality gate admits (math-wins-over-C++). Total alternatives: \cppinline{SignedCardinality} (saturating), \cppinline{SignedExtensionalCardinal<>} (cyclic), or the \cppinline{:partial} Maybe-monad chain for honest fail-on-overflow.

\textsuperscript{b} For narrow signed integers \cppinline{HasRingOperators} \emph{refuses}: integer promotion (C++ [conv.prom]) lifts \cppinline{int8_t + int8_t} to \cppinline{int}, so literal arithmetic does not close on the carrier.  The functor-parametric surface \cppinline{IsAlgebra<T, std::plus<T>, std::multiplies<T>>} fires because \cppinline{std::plus<T>} returns \cppinline{T} by signature --- this is the structural-shape level, not a ring-axiom certification: signed-overflow UB still defeats the strict \cppinline{IsRing} bundle along the same path footnote~a names for \cppinline{int} / \cppinline{long}, and the project's math-wins-over-C++ stance therefore withholds ring-axiom certification on narrow signed carriers.  Total alternatives are the same as in footnote~a.}
\end{flushleft}
\end{table}

\begin{table}[htbp]
\centering
\caption{\textbf{Crossings state-machine.} The named morphisms that
move between exact and primitive sides, per role. Each row shows
\emph{realise} (exact $\to$ primitive: drop exactness for speed) and
\emph{embed} (primitive $\to$ exact: lift a machine value into the
certified carrier). A dash marks a direction that requires going
through an intermediate role.}
\label{tab:carriers-crossings}
\small
\begin{tabular}{@{}l l l@{}}
\toprule
\textbf{Role} & \textbf{Realise: exact $\to$ primitive} & \textbf{Embed: primitive $\to$ exact} \\
\midrule
$\mathbb{B}$ & (identity) & (identity) \\
$\mathbb{N}$ & \cppinline{.realize_to_size_t(sentinel)} & \cppinline{embed_unsigned_integral<N>(v)} \\
$\mathbb{Z}$ & \cppinline{static_cast<int64_t>(x)} & \cppinline{embed_signed_integral<Z>(v)} \\
$\mathbb{Q}$ & \cppinline{.resolve()} $\to$ \cppinline{double} & \cppinline{Rational<Z>(num, den)} \\
$\mathbb{R}$ symbolic & evaluate cut at a rational bound & (intensional; no simple lift) \\
$\mathbb{R}$ policy & \cppinline{discharge_ieee(IEEE<double>)} & \cppinline{assume_ieee(Real<double>)} \\
$\mathbb{C}$ & per-component \cppinline{.resolve()} & \cppinline{Complex<R>(re, im)} \\
$\mathbb{D}$ (dual) & \cppinline{.val} (primal only; tangent discarded) & \cppinline{Dual<F>(v, d)} \\
\bottomrule
\end{tabular}
\end{table}

A fourth table addresses where textbook ring structure and literal
C++ operators diverge: the library expresses the disagreement
through three concept tiers --- \emph{strict} (categorical proof),
\emph{functor-parametric shape} (operators close on the chosen
functors), and \emph{literal shape} (the C++ \cppinline{+},
\cppinline{-}, \cppinline{*} close on the carrier directly) ---
sealed by \cppinline{IsArithmeticRing} when all three agree under
\cppinline{(std::plus, std::multiplies)}.
Table~\ref{tab:rosetta-textbook-vs-cpp} catalogues the divergences
on the carriers used in the worked examples (Issue~\#393); the
friction list is summarised structurally in
Section~\ref{sec:boundary-observations}.

\begin{table}[htbp]
\centering
\caption{\textbf{Math textbook $\leftrightarrow$ C++ standard.} Where the
literal C++ operators agree with the textbook (rows that tick all
columns), the bundled seal \cppinline{IsArithmeticRing} fires and
generic code can be written in plain syntax. Where they diverge
(integer promotion, signed-overflow UB), the table names the precise
failure point and the tier that still fires. ``$\circ$'' marks an
assertion that fires only at the literal-shape tier (the strict half
is gated by an open FIXME on the species-trait registry); ``--''
marks not-currently-asserted.}
\label{tab:rosetta-textbook-vs-cpp}
\small
\begin{tabular}{@{}l l c c c c@{}}
\toprule
\textbf{Carrier} & \textbf{Math claim} & \textbf{Strict} & \textbf{Functor} & \textbf{Literal} & \textbf{Seal} \\
\midrule
\cppinline{unsigned int}, \cppinline{unsigned long}, \cppinline{size_t} & ring $\mathbb{Z}/2^N\mathbb{Z}$ & \checkmark & \checkmark & \checkmark & \checkmark \\
\cppinline{unsigned short}, \cppinline{unsigned char} & ring $\mathbb{Z}/2^N\mathbb{Z}$ & \checkmark & \checkmark & $\times$\textsuperscript{a} & $\times$ \\
\cppinline{int}, \cppinline{long} & partial (signed-overflow UB) & $\times$\textsuperscript{b} & \checkmark & \checkmark & $\times$ \\
\cppinline{bool} under $(\oplus, \wedge)$ & Galois field $\mathbb{F}_2$ & \checkmark & \checkmark & $\times$\textsuperscript{a} & $\times$\textsuperscript{c} \\
\cppinline{Rational<long>} & field $\mathbb{Q}$ & $\circ$\textsuperscript{d} & $\circ$ & \checkmark & $\circ$ \\
\cppinline{RigPolynomial<unsigned int>} & polynomial ring $R[x]$ & \checkmark & \checkmark & \checkmark & \checkmark \\
\cppinline{Matrix2x2V<unsigned int>} & non-commutative ring & $\circ$ & $\circ$ & \checkmark & -- \\
\cppinline{Vec2V<T>} & module / vector space, \emph{not a ring} & n/a & n/a & $\times$\textsuperscript{e} & n/a \\
\bottomrule
\end{tabular}

\vspace{0.5em}
{\footnotesize
\begin{tabular}{@{}l p{0.85\textwidth}@{}}
\textsuperscript{a} & integer promotion lifts the result of literal \cppinline{+} / \cppinline{*} to \cppinline{int}; the functor surface returns \cppinline{T} by signature. \\
\textsuperscript{b} & signed-overflow UB means the carrier admits no \cppinline{IsTotal} label (it is neither periodic, nor idempotent, nor saturating); the categorical \cppinline{IsRing} chain --- which composes \cppinline{IsTotal} --- therefore breaks for this carrier. \\
\textsuperscript{c} & \cppinline{bool}'s ring structure lives over $(\oplus, \wedge)$, not $(+, *)$; \cppinline{IsArithmeticRing}, which fixes \cppinline{(std::plus, std::multiplies)}, asks the wrong question for this carrier. \\
\textsuperscript{d} & operational under the active numeric policy: \cppinline{HasFieldOperators} is the operational variant, the categorical proof is intentionally not specialised. \\
\textsuperscript{e} & vectors have no internal product \cppinline{Vec2V $\times$ Vec2V $\to$ Vec2V}, only scalar multiplication; the shape concept correctly refuses. \\
\end{tabular}
}
\end{table}

Reading rule for callsites: pick \emph{seal} for plain C++ syntax
over a textbook-aligned carrier; pick \emph{strict + functor} when
the carrier carries its ring structure over non-standard functors
(\cppinline{bool} as $\mathbb{F}_2$, narrow unsigneds with
\cppinline{std::plus} / \cppinline{std::multiplies}); pick
\emph{literal} for the operator surface without the algebraic
axioms (elementwise-lifted matrix bodies, halfspace evaluation).

\paragraph{Migration discipline.}
The three tables form a migration map: \emph{start correct} on
Table~\ref{tab:carriers-exact} (typically
\cppinline{Rational<SignedExtensionalCardinal<>>} for
arithmetic-heavy code, with bit-for-bit provable laws); \emph{cross
over} via Table~\ref{tab:carriers-crossings} only in profiled inner
loops where arithmetic cost dominates; \emph{land} on
Table~\ref{tab:carriers-primitive} with the named caveat the
concept cell records (e.g.\ loss of associativity under
\cppinline{+}, which some algorithms tolerate and others do not).
The rule: \textbf{start correct, finish fast, and the type system
enforces that the laws your code relies on are the laws the carrier
actually provides --- or fails to compile.}


The second pillar.  Sets as predicates over an ambient species (the
Comprehension View), and the symbolic composition of set operations
under the subobject classifier $\Omega$.  Where §\ref{sec:axiomatic}
fixes the substrate, the sets layer reifies its first concrete
inhabitants — sets that are rules, not buckets.  The sequence-layer
counterpart (functions $\mathbb{N} \to T$ as infinite-dimensional
vectors, sharing the same intensional posture) extends this pillar
in a sibling slice.
\section{The $\mathbf{Ddk}$ DSL: Rosetta and Worked Exhibits}
\label{app:ddk-rosetta}

§\ref{sec:cpp-stlc-mapping} reports the $\mathbf{Ddk}$ surface as the intersection $\mathbf{Jlt} \cap \mathbf{Alg}(-)$ and pins Honest Rejection as the headline policy.  This appendix collects the verification artefacts named in passing there: a row-by-row mapping from the disciplined C++23 fragment to STLC and category theory, an operator-shape rosetta linking C++ operators to the algebraic concepts they witness, and one worked Functor exhibit showing the apparatus end-to-end on a concrete example.

\begin{table}[htbp]
\centering
\caption{\textbf{From disciplined C++23 to set builder notation to simply typed $\lambda$-calculus and category theory.}  Correspondence /
mapping table from concepts and keywords in the disciplined fragment
of the C++ language to STLC; exhibits of realisations are referenced
in-line in column~1.  The midrule separates standard C++23 features
(top) from the set-builder DSL identifiers (bottom) that name the
subobject classifier $\Omega$ and the Cartesian-closed-category
witness on which the set-comprehension surface
(§\ref{sec:sets-rules}) is built.}
\label{tab:cpp-stlc-mapping}
\small
\begin{tabular}{@{}p{0.42\textwidth} p{0.52\textwidth}@{}}
\toprule
\textbf{Disciplined C++23} & \textbf{Lambda calculus / category theory} \\
\midrule
\cppinline{template <typename T>} (§\ref{sec:axiomatic}) & Type variable $\alpha$; System-F universal quantification at the type level (parametric \emph{in the disciplined subset}; no specialisation, no SFINAE branch in the public surface) \\
\cppinline{concept} (§\ref{sec:axiomatic}) & Type-class / kind constraint --- a predicate on types restricting the domain of a polymorphic function \\
\cppinline{requires}-clause (§\ref{sec:axiomatic}) & Evidence / constraint satisfaction --- a proof obligation that the type carries the term-level operations \\
\cppinline{static_assert} (throughout) & Mechanical proof that proposition $P$ is inhabited at translation time (Curry--Howard~\cite{wadler2015propositions}) \\
\cppinline{using} alias (§\ref{sec:axiomatic}) & Type-level redex --- the compiler reduces by $\beta$-style substitution \\
\cppinline{constexpr} evaluation (§\ref{sec:lp-centrepiece}, IR fixtures) & Normalisation ($\beta$-reduction) --- reduction of a term to its value or literal form \\
\cppinline{std::same_as} (§\ref{sec:lp-centrepiece}) & Definitional type equality $\tau = \sigma$ \\
\midrule
\texttt{dedekind} & \\
\midrule
NTTP-carried lambda $\lambda x. P(x)$ (§\ref{sec:sets-rules}) & Characteristic morphism $\chi_S : A \to \Omega$ defining a set by its membership rule \\
\cppinline{Ø<T>} (empty-set type; §\ref{sec:pruning}, Showcase 3) & Bottom type $\bot$ --- the uninhabited type witnessing a logical contradiction \\
\cppinline{HasCanonicalSetCCC<T>} (§\ref{sec:sets-rules}) & The Cartesian-closed-category witness ($1$, $\times$, $(-)^{(-)}$) on the ambient species \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\caption{\textbf{Operator Rosetta: what the C++ programmer types
↔ what they are operationally claiming.}  The
\cppinline{Has*Operators} family pins the \emph{operator surface}
of an algebraic structure.  Each section heading names the concept;
each row maps the C++ spelling to the operation it witnesses.  The
operator-surface concepts are the operational gate; the strict
axiomatic concepts (\cppinline{IsRing}, \cppinline{IsField},
\cppinline{IsVectorSpace}, \dots; see Lang~\cite{lang2002algebra}
for the standard structures and Burris--Sankappanavar
\cite{burris1981universalalgebra} for the universal-algebra anchor)
compose them with the species-trait registry that pins the laws.
Carriers may carry the operators (\cppinline{Has*Operators}) without
the laws holding strictly --- e.g.\ \cppinline{double} satisfies
\cppinline{HasFieldOperators} but not strict \cppinline{IsField}
because IEEE~754 rounding defeats associativity.}
\label{tab:operator-rosetta}
\small
\begin{tabular}{@{}p{0.40\textwidth} p{0.54\textwidth}@{}}
\toprule
\textbf{C++ operator surface} & \textbf{Algebraic concept} \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasGroupOperatorsAdd}} --- additive group $(G, +, -, 0)$} \\
\midrule
\cppinline{a + b}, \cppinline{-a}, \cppinline{a - b} & group operation, inverse, $a + (-b)$ \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasGroupOperatorsMul}} --- multiplicative group $(G, \cdot, ^{-1}, 1)$} \\
\midrule
\cppinline{a * b}, \cppinline{a / b}, \cppinline{T{1}} & group operation, $a \cdot b^{-1}$, multiplicative unit (reciprocal of $x$ is \cppinline{T{1} / x}; no \cppinline{.inverse()} method required) \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasSemiringOperators}} --- semiring (rig: no additive inverse)} \\
\midrule
\cppinline{a + b}, \cppinline{a * b}, \cppinline{T{}}, \cppinline{T{1}} & both monoid operations + their identities; \emph{no} unary \cppinline{-} \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasRingOperators}} --- ring operator surface $(+, -, \cdot)$} \\
\midrule
\cppinline{a + b}, \cppinline{a - b}, \cppinline{-a}, \cppinline{a * b} & ring addition + multiplication; absorbing zero is structural \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasFieldOperators}} --- field $(F, +, -, \cdot, /, 0, 1)$, $\equiv$ \cppinline{HasRingOperators} $\wedge$ \cppinline{HasGroupOperatorsMul}} \\
\midrule
\cppinline{a / b}, \cppinline{T{1}} (in addition to \cppinline{HasRingOperators}) & multiplicative inverse $a \cdot b^{-1}$ on non-zero elements; multiplicative unit \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasVectorSpaceOperators<V, F, Act>}} --- vector space over scalar field $F$ (action functor \cppinline{Act}, defaults to \cppinline{std::multiplies<>})} \\
\midrule
\cppinline{v + w}, \cppinline{-v} (additive group on $V$) & vector addition; \cppinline{HasGroupOperatorsAdd<V>} \\
\cppinline{Act{}(s, v)} with \cppinline{s : F} & scalar action $F \times V \to V$ via the action functor; \cppinline{s * v} under the default \cppinline{std::multiplies<>}. The right-action \cppinline{v * s} is conventional, not required by the concept. \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasLatticeOperators}} --- lattice $(L, \vee, \wedge)$} \\
\midrule
\cppinline{a | b}, \cppinline{a & b}, \cppinline{a ^ b}, \cppinline{~a} & join, meet, symmetric difference, complement (bitwise on \cppinline{unsigned}) \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasPartialOrderOperators} / \cppinline{HasTotalOrderOperators}}} \\
\midrule
\cppinline{a < b}, \cppinline{a <= b}, \cppinline{a > b}, \cppinline{a >= b} & poset / chain comparisons; results in $\Omega$ of the chosen logic \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasKleisliBindOperators}} --- Kleisli (monadic) extension} \\
\midrule
\cppinline{ma >>= f}, \cppinline{ma >> f} & $\kappa(ma, f) = \mu(\varphi(ma, f))$; bind / fish \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasComonadicExtendOperators}} --- co-Kleisli (comonadic) extend} \\
\midrule
\cppinline{wa <<= f}, \cppinline{wa << f} & $\sigma(wa, f) = \varphi(\delta(wa), f)$; extend / cofish \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasSubscriptOperator}} / \cppinline{HasArrowDereferenceOperator} --- CCC eval / comonadic extract (PR \#533)} \\
\midrule
\cppinline{s[i]} & CCC eval counit $\varepsilon : (\textit{Idx} \times T^{\textit{Idx}}) \to T$ \\
\cppinline{m->member} & comonadic extract $\varepsilon : W\langle T\rangle \to T$ (Wadler~\cite{wadler1992comprehending}) \\
\midrule
\multicolumn{2}{@{}l}{\textbf{\cppinline{HasCanonicalSetCCC}} --- CCC primitives on the ambient species} \\
\midrule
$1$, $\times$, $(-)^{(-)}$ & terminal object, binary product, exponential (the Lambek--Scott bridge to STLC) \\
\bottomrule
\end{tabular}
\end{table}

Three rows are doing more work than the others, and the paper's
contribution lies on top of them: the subobject classifier
$\Omega$ (the codomain of every NTTP-carried lambda; see
§\ref{sec:sets-rules}) is what makes a predicate a \emph{set} in
the structuralist sense; the Lambek--Scott correspondence is what
makes the set-builder DSL a syntax for a Cartesian-closed
category---hence, equivalently, for STLC; and the
\cppinline{constexpr}-as-normalisation row is what lets type-directed
collapse (§\ref{sec:gap}) act as a form of cut elimination, with the
LLVM backend treating the reduced bottom type as dead code.

\paragraph{The textbook functor diagram for Listing~\ref{lst:isfunctor}.}
Figure~\ref{fig:functor-diagram} is the commutative-square diagram for a functor $F: \mathcal{C} \to \mathcal{D}$~\cite{maclane1971categories} (Mac Lane §I.3): an object map $A \mapsto F(A)$ and an arrow map $f \mapsto F(f)$ that respects identities and composition.
The body's \cppinline{IsFunctor} concept (Listing~\ref{lst:isfunctor}, §\ref{sec:cpp-stlc-mapping}) encodes the same content structurally; the Curry--Howard discharge that \texttt{maybe\_functor} inhabits the concept is the \cppinline{static_assert} in the worked-instance Listing~\ref{lst:maybe_functor}.
Together, Listings~\ref{lst:isfunctor} and~\ref{lst:maybe_functor} land the System-F-recovery rosetta (Table~\ref{tab:structuralist-bridge}) at one concrete site: the \cppinline{template <typename F>} header is the \emph{Constrained Templates} row ($\equiv \forall\alpha.\dots$); the \cppinline{requires}-clause is the \emph{Concept-gating} row discharging parametricity; and Listing~\ref{lst:maybe_functor}'s \cppinline{static_assert} is the Curry--Howard term.

\begin{figure}[htbp]
\centering
\begin{tikzcd}[column sep=large, row sep=large, ampersand replacement=\&]
A \arrow[r, "f"] \arrow[d, mapsto, "F"', shorten <=2pt, shorten >=2pt]
  \arrow[rr, bend left=35, "g \circ f"] \&
B \arrow[r, "g"] \arrow[d, mapsto, "F", shorten <=2pt, shorten >=2pt] \&
C \arrow[d, mapsto, "F", shorten <=2pt, shorten >=2pt] \\
F(A) \arrow[r, "F(f)"']
     \arrow[rr, bend right=35, "F(g \circ f)\;=\;F(g) \circ F(f)"'] \&
F(B) \arrow[r, "F(g)"'] \&
F(C)
\end{tikzcd}
\caption{The textbook functor diagram, extended with a second source arrow $g: B \to C$ so the composition law is observable rather than only stated.  $F$ sends each source object $X \in \mathcal{C}$ to a target object $F(X) \in \mathcal{D}$ and each source arrow $f: A \to B$ to a target arrow $F(f): F(A) \to F(B)$, in a way that respects identities ($F(\mathrm{id}_A) = \mathrm{id}_{F(A)}$) and composition ($F(g \circ f) = F(g) \circ F(f)$, witnessed by the curved arrows on top and bottom).  Listing~\ref{lst:isfunctor} encodes the same content in C++23 concepts, and Listing~\ref{lst:maybe_functor} exhibits a worked instance.}
\label{fig:functor-diagram}
\end{figure}

\paragraph{The hub-vs-carrier obligation placement.}
An architectural choice worth naming: the functorial obligation in Listing~\ref{lst:isfunctor} lives in the \emph{hub} type (\cppinline{maybe_functor<T>}), not in the carrier (\cppinline{std::optional<T>}).
In Haskell the carrier itself is the functor; \cppinline{Maybe a} is declared an instance of \cppinline{Functor} with the laws supplied implicitly.
In \texttt{dedekind} we deliberately keep \cppinline{std::optional<T>} a Plain Old C++ Object and externalise the functorial proof obligation into a separate hub: an engineer who wants \cppinline{std::optional} to act as a functor writes \cppinline{maybe_functor} and discharges the laws there.
The C++23 concept machinery does not project structural witnesses onto carriers automatically (concepts are checked, not synthesised).
Crowding self-descriptive bits onto \cppinline{std::optional} would defeat the POCO discipline that lets the disciplined fragment recover System F (Table~\ref{tab:structuralist-bridge}).
The real-world functions an engineering library hosts are not generic in the strong Haskell sense: the functorial structure is supplied at the library boundary by a hub that names it, not inherited from the carrier.

\paragraph{Every carrier is a structure-preserving image of its base.}
Each carrier in the numeric tower (Figure~\ref{fig:carrier-lattice}) is the image of a structure-preserving construction applied to its base.
Each carrier sits as the image of an H or P operation of the HSP theorem~\cite{burris1981universalalgebra} (§II.10).\footnote{H, S, P are the three structure-preserving operations on algebras: \textbf{H}omomorphic images, \textbf{S}ubalgebras, direct \textbf{P}roducts.  A class is a variety iff it is closed under all three.  S is not heavily exercised by today's carriers and is left absent.  $\operatorname{End}_R$ at the rank axis is \emph{not} an HSP operation: it is the internal hom in $\operatorname{Mod}_R$, which preserves only some axioms (matrix multiplication is non-commutative even when $R$ is commutative).  Its partial trait propagation lives behind a separate \cppinline{is_endomorphism_ring_v} trait.  The arrow-side cousin to \cppinline{IsQuotientAlgebra} is \cppinline{IsQuotientMorphism} in \texttt{algebra:universal} (a declared homomorphism + declared surjectivity, naming the canonical projection $\pi: A \to A/{\sim}$).}
The library reifies these as marker-trait + detection-concept pairs in \texttt{algebra:quotient} (Table~\ref{tab:hsp-reification}).
A single declaration at each carrier site (specialising the marker trait's \cppinline{::type} member to the base) fires the structural-trait propagation (associativity, commutativity, distributivity, the \cppinline{IsTotal} certificate via its periodic / idempotent / saturating paths) from base to image.

\begin{table}[htbp]
\centering
\caption{HSP reification in \texttt{algebra:quotient}.  Each row maps a structure-preserving operation to its marker trait, detection concept, and a representative carrier-site witness.  S is omitted because it is not heavily exercised by today's carriers; $\operatorname{End}_R$ at the rank axis is not an HSP operation and is treated separately (preceding paragraph's footnote).}
\label{tab:hsp-reification}
\small
\begin{tabular}{@{}p{0.10\textwidth}p{0.27\textwidth}p{0.22\textwidth}p{0.30\textwidth}@{}}
\toprule
\textbf{HSP op} & \textbf{Marker trait} & \textbf{Detection concept} & \textbf{Example carrier} \\
\midrule
H (quotient) & \cppinline{quotient_algebra_base<Q>::type} & \cppinline{IsQuotientAlgebra<Q>} & \cppinline{Rational<Z>} over \cppinline{Z} (Frac construction) \\
P (product)  & \cppinline{product_algebra_base<Q>::type}  & \cppinline{IsProductAlgebra<Q>}  & \cppinline{Vec2V<T>} over \cppinline{T} (rank-2 free module) \\
\bottomrule
\end{tabular}
\end{table}

\section{The Compiler as Structural Gate and Optimization Engine}

Two durable uses of a compile-time type system underlie this paradigm, neither
of which requires the compiler to be a theorem prover:

\begin{enumerate}
  \item \textbf{Ruling out entire classes of errors.}  Encoding algebraic laws as
        concept requirements makes every instantiation of a generic template an
        implicit proof obligation discharged at translation time.  Failed
        obligations are type errors, not runtime exceptions.  The obligation
        itself is finite, decidable, and mechanical---the compiler \emph{checks}
        it, but does not \emph{search for} it.  Failed obligations surface
        as named-axiom messages (e.g.\ \texttt{Invertible2x2 requires full
        rank: a*d - b*c != 0}) rather than as deep template-cascade
        traces; the structural discipline behind this and a worked
        diagnostic are exhibited in Section~\ref{sec:compiler-wall}.
  \item \textbf{Enabling entire classes of optimizations.}  The LLVM backend
        exploits the structural knowledge inherited from discharged obligations.
        When a set is defined by a predicate and two predicates are logically
        contradictory, the compiler determines at translation time that the
        resulting intersection is empty and emits no machine instructions for
        it.  This is what we term \emph{structural pruning}, and it is the
        subject of Section~\ref{sec:pruning}.
\end{enumerate}

Both uses are bounded but open-ended.  clang's type checker is not a
theorem prover: it has no dependent types, no universal-quantification
inference, no proof-search engine.  What it does have is
\cppinline{static_assert}, \cppinline{requires}-clauses, and
\cppinline{constexpr} evaluation---enough to verify a predetermined
proof sketch, not enough to generate one.  The paradigm claims the
reliable half: ruling out, and optimizing under, the classes of
structural failures the programmer has named in the type system.  The
Curry--Howard correspondence between propositions and types is
suggestive here, but its full force (dependent products, inductive
proofs, universe hierarchies) is not available in C++23; what
\emph{is} available is enough to anchor the two uses above.  Each use
ruled out exposes a next class that the current machinery does not
handle but the library's structural vocabulary already names; the four
axes of Section~\ref{sec:gap} fall out of the halfspace reductions
almost by construction.\footnote{Penrose's \emph{The Road to
Reality}~\cite{penrose2004roadtoreality} offers an analogous
observation about G\"odel's incompleteness theorem --- a formal
system closing at one level motivates the next, not as exhaustion
but as natural successor; the closest formal analogue is Turing's
ordinal logics~\cite{turing1939ordinals}, where each stage's
consistency feeds the next stage's formation rule.  The
systems-programming reading is benign: the issue tracker plays the
successor role.}

\paragraph{Three honesty-anchoring constraints.}
Three compounding constraints keep the library's artefacts honest:
\emph{domain textbook anchoring} (concept names and law statements
taken directly from the published mathematical vocabulary ---
Lawvere, Mac Lane, Lambek, Wadler, Moggi); \emph{mechanical
validation} (every algebraic claim becomes a
\cppinline{static_assert} the compiler discharges; every concept
composes through a named-module DAG; every behavioural claim passes
CI); and \emph{codomain textbook anchoring} (concepts validated
against the C++ standard library's own ontology ---
\cppinline{std::ranges::input_range}, \cppinline{std::regular},
\cppinline{std::numeric_limits}, the iterator
hierarchy).\footnote{The library is developed with AI assistance
(Anthropic's Claude as the primary code-writing agent; Gemini and
GitHub Copilot in supporting roles); the human role is
prioritisation, orchestration, and consensus-before-merge.  The
three constraints above are what keep the workflow honest: the
mathematical content stays in mathematical vocabulary all the way
down to where the machine takes over.}

A successful type-check is thus a \emph{weak but universal} proof about the overall health of the Ontology: it asserts that every type-shaped claim in the build graph composes without contradiction, even though the individual assertions are coarse (concept satisfaction, not law-by-law verification).
The automated test suite is the converse, a growing set of \emph{strong but existential} proofs about the same Ontology: each test pins a concrete carrier-instance against a concrete law and discharges the obligation as a \cppinline{static_assert} witness rather than a runtime equality check.
Universal coverage is the build's job; existential depth is the test suite's; both accumulate monotonically under the convergence discipline.
As a convenient and notable side effect: source code listings presented in this document are lifted (occasionally abridged for paper-page budget) from the repository's test suite; each listing's source-tree path is recorded as a LaTeX-source comment immediately above the listing, so a reader auditing against the repository can find the verbatim original directly.

Similarly, where the output of LLVM optimizations is reported here, these are examples which are lifted from the test suite.
Here, the compiler is run during CI with appropriate flags to emit a portable representation of the optimized backend instructions.
The emitted instructions are then compared to LLVM IR fixtures checked into the source tree.
This way, it can be checked during CI that algebraic compile-time optimizations are recognized by the compiler during the development cycle.
Regressions would lead to build failures and be flagged during CI as blocking the corresponding PR.


\section{Use of Agentic Systems}

Agentic systems --- LLMs invoked as code-review and refactoring assistants --- enter this picture as a third party that can act on either side of the universal/existential split, provided the source carries enough domain context to prime them on each invocation.
The library therefore embeds the domain language deliberately deep into the source: exhaustive Doxygen on every public surface, opinionated quotes from practitioners (e.g.\ references to the Polish school of logic and semantic epistemology~\cite{ajdukiewicz1985jezyk} where the underlying register is at issue), and named provenance on each trait specialisation.
The audience is not only the human reader; it is also the agentic context window on every pull-request review.
Prompts to such systems are accordingly written liberally and in \emph{loose language} --- in the careful, sparing register of RFC~2119~§6~\cite{bradner1997rfc2119}, where the normative weight of \textsc{MUST} / \textsc{SHOULD} / \textsc{MAY} is reserved for the cases that actually demand it --- so that the agent can catalyse convergence not only at the syntactic layer the conventional build pipeline polices, but also at the semantic layer where loose intent has to be translated into precise concept-discharge obligations.


\end{document}
